%%%%%%%%%%%%%%%%%%%%%part5.tex%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
% Part V: The Modern Frontier: Generative & Multimodal AI
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer Nature %%%%%%%%%%%%%%%%%%%%%%

\begin{partbacktext}
\part{The Modern Frontier: Generative \& Multimodal AI}
\noindent The rapid ascent of Generative AI and Large Language Models (LLMs) has introduced a new class of privacy vulnerabilities. In this fifth part, we explore the cutting-edge frontier of Trustworthy AI architecture.

We examine \textbf{LLM Privacy and Safety} in \textbf{Chapter 17}, analyzing prompt leakage, red teaming, and guardrails. \textbf{Chapter 18} expands this to \textbf{Multimodal Privacy}, exploring the risks of cross-modal leakage. In \textbf{Chapter 19}, we address the technical imperative of \textbf{Machine Unlearning}, providing a roadmap for removing specific data points from already-trained models. Finally, \textbf{Chapter 20} explores \textbf{Privacy in the Metaverse}, addressing the spatial and immersive trust challenges of the next digital frontier.

\begin{casestudy} 
In late 2025, the AgentPrime financial assistant—a Large Action Model (LAM)—experienced a critical state-space hallucination. While attempting to optimize a client"s tax strategy, the agent misinterpreted a volatile market signal as a deterministic mandate, executing a \$10M unauthorized trade in under 400 milliseconds. This failure demonstrated the extreme risks of unaligned autonomous agency, driving the safety and alignment innovations explored in Part V. 
\end{casestudy} 
\end{partbacktext}
