%%%%%%%%%%%%%%%%%%%%%%% references.tex %%%%%%%%%%%%%%%%%%%%%
% 
% References for The Trustworthy AI Architect
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer Nature %%%%%%%%%%%%%%%%%%

\begin{thebibliography}{99.}%

\bibitem{science-mono} Dwork, C., Roth, A.: The Algorithmic Foundations of Differential Privacy. Foundations and Trends in Theoretical Computer Science (2014)

\bibitem{science-online} Nissenbaum, H.: Privacy in Context: Technology, Policy, and the Integrity of Social Life. Stanford Law Books (2010)

\bibitem{science-journal} Kairouz, P., et al.: Advances and Open Problems in Federated Learning. Foundations and Trends in Machine Learning \textbf{14}(1--2), 1--210 (2021)

\bibitem{science-DOI} Tran-Truong, P.T., et al.: POP2TIC: Performance Optimization for Privacy-Preserving Fog Computing. Journal of Network and Computer Applications (JNCA) (2024)

\bibitem{sweeney2002} Sweeney, L.: k-Anonymity: A Model for Protecting Privacy. International Journal of Uncertainty, Fuzziness and Knowledge-Based Systems \textbf{10}(05), 557--570 (2002)

\bibitem{machanavajjhala2007} Machanavajjhala, A., Kifer, D., Gehrke, J., Venkitasubramaniam, M.: l-Diversity: Privacy Beyond k-Anonymity. ACM Transactions on Knowledge Discovery from Data (TKDD) \textbf{1}(1), 3--es (2007)

\bibitem{shokri2017} Shokri, R., Stronati, M., Song, C., Shmatikov, V.: Membership Inference Attacks Against Machine Learning Models. In: IEEE Symposium on Security and Privacy (SP), pp. 3--18 (2017)

\bibitem{zhu2019} Zhu, L., Liu, Z., Han, S.: Deep Leakage from Gradients. In: Advances in Neural Information Processing Systems (NeurIPS) (2019)

\bibitem{trantruong2020} Tran-Truong, P.T., et al.: Data Poisoning Attack \& Defenses. In: Proceedings of the 13th International Conference on Advanced Computing and Applications (ACOMP) (2020)

\bibitem{mcmahan2017} McMahan, B., Moore, E., Ramage, D., Hampson, S., y Arcas, B.A.: Communication-Efficient Learning of Deep Networks from Decentralized Data. In: Artificial Intelligence and Statistics (AISTATS), pp. 1273--1282 (2017)

\bibitem{bonawitz2017} Bonawitz, K., et al.: Practical Secure Aggregation for Privacy-Preserving Machine Learning. In: Proceedings of the 2017 ACM SIGSAC Conference on Computer and Communications Security (CCS), pp. 1175--1191 (2017)

\bibitem{costan2016} Costan, V., Devadas, S.: Intel SGX Explained. Cryptology ePrint Archive (2016)

\bibitem{blanchard2017} Blanchard, P., Guerraoui, R., Stainer, J., et al.: Machine Learning with Adversaries: Byzantine Tolerant Gradient Descent. In: Advances in Neural Information Processing Systems (NeurIPS) (2017)

\bibitem{goodfellow2014} Goodfellow, I.J., Shlens, J., Szegedy, C.: Explaining and Harnessing Adversarial Examples. arXiv preprint arXiv:1412.6572 (2014)

\bibitem{ribeiro2016} Ribeiro, M.T., Singh, S., Guestrin, C.: ``Why Should I Trust You?'': Explaining the Predictions of Any Classifier. In: Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining, pp. 1135--1144 (2016)

\bibitem{lundberg2017} Lundberg, S.M., Lee, S.I.: A Unified Approach to Interpreting Model Predictions. In: Advances in Neural Information Processing Systems (NeurIPS) (2017)

\bibitem{han2015} Han, S., Mao, H., Dally, W.J.: Deep Compression: Compressing Deep Neural Networks with Pruning, Trained Quantization and Huffman Coding. arXiv preprint arXiv:1510.00149 (2015)

\bibitem{carlini2021} Carlini, N., et al.: Extracting Training Data from Large Language Models. In: 30th USENIX Security Symposium (2021)

\bibitem{hu2021} Hu, E.J., et al.: LoRA: Low-Rank Adaptation of Large Language Models. arXiv preprint arXiv:2106.09685 (2021)

\bibitem{bourtoule2021} Bourtoule, L., et al.: Machine Unlearning. In: IEEE Symposium on Security and Privacy (SP), pp. 141--159 (2021)

\bibitem{katz2017} Katz, G., Barrett, C., Dill, D.L., et al.: Reluplex: An Efficient SMT Solver for Verifying Deep Neural Networks. In: Computer Aided Verification (CAV) (2017)

\bibitem{cohen2019} Cohen, J., Rosenfeld, E., Kolter, Z.: Certified Adversarial Robustness via Randomized Smoothing. In: International Conference on Machine Learning (ICML) (2019)

\bibitem{ouyang2022} Ouyang, L., et al.: Training Language Models to Follow Instructions with Human Feedback. In: Advances in Neural Information Processing Systems (NeurIPS) (2022)

\bibitem{cisa2021} CISA: Software Bill of Materials (SBOM). Cybersecurity and Infrastructure Security Agency (2021)

\bibitem{vn-decree13} Government of Vietnam: Decree No. 13/2023/ND-CP on Personal Data Protection (2023)

\bibitem{eu-ai-act} European Parliament: Artificial Intelligence Act (2024)

\bibitem{opacus} Yousefpour, A., et al.: Opacus: User-Friendly Differential Privacy in PyTorch. arXiv preprint arXiv:2109.12298 (2021)

\bibitem{flower} Beutel, D.J., et al.: Flower: A Friendly Federated Learning Framework. arXiv preprint arXiv:2007.14390 (2020)

\bibitem{li2007} Li, N., Li, T., Venkatasubramanian, S.: t-Closeness: Privacy Beyond k-Anonymity and l-Diversity. In: IEEE 23rd International Conference on Data Engineering (ICDE), pp. 106--115 (2007)

\bibitem{hardt2016} Hardt, M., Price, E., Srebro, N.: Equality of Opportunity in Supervised Learning. In: Advances in Neural Information Processing Systems (NeurIPS) (2016)

\bibitem{verma2018} Verma, S., Rubin, J.: Fairness Definitions Explained. In: IEEE/ACM International Workshop on Software Fairness (FairWare), pp. 1--7 (2018)

\bibitem{bai2022} Bai, Y., et al.: Constitutional AI: Harmlessness from AI Feedback. arXiv preprint arXiv:2212.08073 (2022)

\bibitem{shrestha2023} Shrestha, R., et al.: A Survey on Zero-Knowledge Machine Learning. arXiv preprint arXiv:2305.10547 (2023)

\bibitem{schwartz2020} Schwartz, R., et al.: Green AI. Communications of the ACM \textbf{63}(12), 54--63 (2020)

\bibitem{nair2022} Nair, V., et al.: Exploring the Privacy Risks of Virtual Reality. In: IEEE Symposium on Security and Privacy (SP) (2022)

\end{thebibliography}
