%%%%%%%%%%%%%%%%%%%%% Chapter 7 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Federated Learning: Decentralized Intelligence
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\motto{``Your data stays on your device. Only model updates are shared.'' This is the promise of Federated Learning, but it is also where the hidden risks begin.}

\chapter{Federated Learning: Decentralized Intelligence}
\label{chap:federated_learning}

\abstract{Federated Learning (FL) has emerged as the leading architecture for collaborative AI without data centralization. However, as this chapter demonstrates, the mere decentralization of data does not guarantee privacy. We explore the core mechanisms of Federated Learning, analyze the "Privacy Illusion" created by raw gradient sharing, and introduce the combined framework of DP-FL. We specifically address the challenges of Non-IID data and communication overhead, drawing on optimizations for resource-constrained Fog and edge infrastructure.}

\section{The Federated Promise}
\label{sec:federated_promise}

Traditional AI training requires moving all raw data into a single centralized server---a massive, vulnerable honeypot for attackers and a regulatory nightmare for architects. \textbf{Federated Learning (FL)} \cite{mcmahan2017} fundamentally changes this pattern. It allows multiple clients (such as hospitals, banks, or mobile devices) to train a global model collaboratively without ever sharing their raw records \cite{science-journal}.

For industries like healthcare, the promise is transformative: ten hospitals can train a state-of-the-art disease prediction model while ensuring that no patient's private history ever leaves the local hospital firewalls.

\section{The FL Architecture: Local Training and Global Aggregation}
\label{sec:fl_architecture}

In a standard Federated Learning loop (often using the \textbf{FedAvg} algorithm \cite{mcmahan2017}), the process follows four repeating steps:

\begin{enumerate}
    \item \textbf{Broadcasting:} The global server sends the current model weights to selected clients.
    \item \textbf{Local Training:} Each client trains the model on its local, private data.
    \item \textbf{Update Submission:} Clients send only the model's weight updates (gradients) back to the server.
    \item \textbf{Aggregation:} The server aggregates the updates (typically via weighted averaging) to improve the global model.
\end{enumerate}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    server/.style={draw, rounded corners=5pt, fill=blue!10, draw=blue!60,
                   inner sep=9pt, font=\small\bfseries, align=center,
                   minimum width=3.6cm, minimum height=1.1cm},
    client/.style={draw, rounded corners=4pt, fill=green!8, draw=green!55,
                   inner sep=7pt, font=\footnotesize\bfseries, align=center,
                   minimum width=2.8cm, minimum height=1.0cm},
    data/.style={draw, rounded corners=3pt, fill=gray!10, draw=gray!50,
                 inner sep=5pt, font=\scriptsize, align=center,
                 minimum width=2.4cm},
    broad/.style={->, >=Stealth, thick, blue!60, align=center, dashed},
    grad/.style={->, >=Stealth, thick,align=center, green!55!black},
    note/.style={font=\scriptsize\itshape, gray!70}
  ]

  %% Global Server
  \node (srv) [server] at (5.5,0)
      {\textbf{Global Server}\\{\scriptsize FedAvg Aggregator}};

  %% Four clients arranged below
  \node (c1) [client] at (0,-3.0)
      {\textbf{Client 1}\\{\scriptsize Hospital A}};
  \node (c2) [client] at (3.5,-3.0)
      {\textbf{Client 2}\\{\scriptsize Hospital B}};
  \node (c3) [client] at (7.0,-3.0)
      {\textbf{Client 3}\\{\scriptsize Bank C}};
  \node (c4) [client] at (10.5,-3.0)
      {\textbf{Client 4}\\{\scriptsize Mobile Device}};

  %% Private data icons below each client
  \node (d1) [data] at (0,-4.3)    {\scriptsize Private Data\\(stays local)};
  \node (d2) [data] at (3.5,-4.3)  {\scriptsize Private Data\\(stays local)};
  \node (d3) [data] at (7.0,-4.3)  {\scriptsize Private Data\\(stays local)};
  \node (d4) [data] at (10.5,-4.3) {\scriptsize Private Data\\(stays local)};

  \draw[gray!40, dashed, thin] (c1.south) -- (d1.north);
  \draw[gray!40, dashed, thin] (c2.south) -- (d2.north);
  \draw[gray!40, dashed, thin] (c3.south) -- (d3.north);
  \draw[gray!40, dashed, thin] (c4.south) -- (d4.north);

  %% Step 1: Broadcast (server → clients, dashed blue)
  \draw[broad] (srv.south) -- node[note, left] {(1) broadcast\\$\theta_t$} (c1.north);
  \draw[broad] (srv.south) -- (c2.north);
  \draw[broad] (srv.south) -- (c3.north);
  \draw[broad] (srv.south) -- node[note, right] {(1) broadcast\\$\theta_t$} (c4.north);

  %% Step 2/3: Local training + gradient return (solid green)
  \draw[grad] (c1.north) -- node[note, left] {(3) $\Delta\theta_1$} ++(0,1.0) -| (srv.south west);
  \draw[grad] (c2.north) -- node[note, left, xshift=2pt] {(3) $\Delta\theta_2$} ++(0,0.5) -| ([xshift=-0.5cm]srv.south);
  \draw[grad] (c3.north) -- node[note, right,xshift=-2pt] {(3) $\Delta\theta_3$} ++(0,0.5) -| ([xshift=0.5cm]srv.south);
  \draw[grad] (c4.north) -- node[note, right] {(3) $\Delta\theta_4$} ++(0,1.0) -| (srv.south east);

  %% Step labels
  \node[note, align=center, blue!70] at (5.5,-1.5)
      {(2) Local training on private data (no raw data leaves device)};
  \node[note, align=center, blue!60] at (5.5,1.1)
      {(4) Server aggregates: $\theta_{t+1} = \theta_t - \eta\,\tfrac{1}{K}\sum_i \Delta\theta_i$};

\end{tikzpicture}
\caption{The Federated Learning Architecture. (1)~The global
         server broadcasts the current model weights $\theta_t$ to selected clients.
         (2)~Each client trains locally on its private data ---
         no raw records leave the device. (3)~Clients upload
         only gradient updates $\Delta\theta_i$. (4)~The server
         aggregates via weighted FedAvg to produce $\theta_{t+1}$.}
\label{fig:ch7_fl_arch}
\end{figure}

\section{The Privacy Illusion: Why FL Alone Is Not Enough}
\label{sec:privacy_illusion}

Despite the decentralization of raw data, Federated Learning alone is not inherently privacy-preserving. Research has shown that these shared model gradients can be reverse-engineered through \textbf{Gradient Inversion Attacks} \cite{zhu2019}, effectively reconstructing the original training images or records from the numerical updates.

\begin{table}
\caption{Primary Risks in Standard Federated Learning}
\label{tab:fl_risks}
\begin{tabular}{p{3cm}p{5cm}p{2cm}}
\hline\noalign{\smallskip}
Risk Type & Mechanism & Impact \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Gradient Inversion & Adversary reconstructs data from shared updates. & \textcolor{red}{$\bullet$} High \\
Membership Inference & Determining if an individual was in a client's set. & \textcolor{orange}{$\bullet$} Medium \\
Model Poisoning & Malicious clients corrupting the global weights. & \textcolor{red}{$\bullet$} High \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\section{The Solution: DP-FL}
\label{sec:dp_fl}

To turn the "Privacy Illusion" into a concrete guarantee, the Trustworthy AI Architect must combine Federated Learning with \textbf{Differential Privacy (DP-FL)}. By having each client clip their gradients and add calibrated Gaussian noise before submission, we ensure that no individual record's signal can be extracted from the global aggregate.

\section{Real-World Challenges: Overhead and Stability}
\label{sec:fl_challenges}

Implementing DP-FL in the wild introduces two major bottlenecks:

\begin{description}
  \item[Communication Overhead:] 
        Sending high-dimensional updates frequently consumes massive bandwidth. In the \textbf{POP2TIC} framework \cite{science-DOI}, we found that by optimizing the compression and noise strategies, overhead can be reduced by up to 40 percent without compromising the $\epsilon$-guarantee.
  \item[Non-IID Data:]
        In many real-world settings, data is not "Independent and Identically Distributed" (Non-IID) across clients. A hospital in a rural area may have significantly different data than one in an urban center, making global convergence difficult when combined with DP noise.
\end{description}

% ------------------------------------------------------------------
\section{Regulatory Alignment: DP-FL and Vietnam's Data Localization Laws}
\label{sec:fl_regulatory}
% ------------------------------------------------------------------

A persistent challenge for AI architects in Southeast Asia is demonstrating to regulators that a distributed system is legally compliant. Vietnam's \textbf{Decree 13/2023/ND-CP} (Personal Data Protection) and \textbf{Decree 356} (Cybersecurity) establish two core obligations: (1) \emph{data localization} — personal data of Vietnamese residents must be stored on servers located in Vietnam; and (2) \emph{data minimization} — only the minimum necessary data should cross jurisdictional boundaries.

The DP-FL architecture satisfies both requirements \textbf{by design}:

\begin{table}[ht]
\centering
\caption{DP-FL compliance mapping against Vietnam Decree 13/2023/ND-CP and Decree 356. Raw personal data never moves; only calibrated, DP-noised gradient tensors cross boundaries.}
\label{tab:fl_decree_mapping}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{p{3.2cm} p{3.8cm} p{4.8cm}}
\hline\noalign{\smallskip}
\textbf{Legal Obligation} & \textbf{Decree Clause} & \textbf{DP-FL Architectural Response} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Data Localization & Decree 13, Art.\ 25: Cross-border transfer requires legal basis & Raw records never leave the local device or institution; only gradient tensors cross province boundaries. \\[3pt]
Data Minimization & Decree 13, Art.\ 6: Collect only strictly necessary data & DP-noised gradients contain no individually identifiable signal; $\varepsilon$ budget enforces an information ceiling. \\[3pt]
Purpose Limitation & Decree 13, Art.\ 6: Data restricted to declared purpose & Gradients are consumed by aggregation only; the server cannot reconstruct individual records. \\[3pt]
Security Obligation & Decree 356, Art.\ 26: Implement technical safeguards & SecAgg masking + TEE-protected aggregation satisfies the "technical and organizational measures'' standard. \\[3pt]
Audit Rights & Decree 13, Art.\ 8: Data subjects have right to processing records & Privacy Budget Ledger provides an immutable per-round $\varepsilon$ audit trail (see Chapter~\ref{chap:differential_privacy}). \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\begin{svgraybox}
\textbf{The ``No Personal Data Crosses the Commune Boundary'' Argument:} When presenting a DP-FL system to Vietnam's Ministry of Information and Communications (MIC), the architect's most effective compliance statement is: ``The raw electronic health records of Patient $i$ never leave Hospital $A$'s server. What Hospital $A$ transmits is a gradient vector $\tilde{\mathbf{g}}$, which is: (a) a statistical summary, not a record; (b) clipped and noise-injected to satisfy $(\varepsilon, \delta)$-DP; and (c) indistinguishable from the gradient that would have been produced even if Patient $i$ were absent.'' This argument maps directly to Article 25 of Decree 13's legitimate transfer basis: the transmitted data is no longer ``personal data'' under the DP definition.
\end{svgraybox}

% ------------------------------------------------------------------
\section{Connectivity-Aware FL for ASEAN Infrastructure}
\label{sec:fl_asean_connectivity}
% ------------------------------------------------------------------

The standard FedAvg loop (Section~\ref{sec:fl_architecture}) assumes clients complete a local training round and submit updates within a fixed synchronization window. In ASEAN deployments --- particularly rural Vietnam, Indonesia's outer islands, and Myanmar's border regions --- intermittent 4G/5G connectivity means a synchronous FL loop will constantly fail, with \textbf{straggler} clients blocking the global round indefinitely.

\subsection{Asynchronous Federated Learning}
\label{subsec:async_fl}

\textbf{Asynchronous FL (AsyncFL)} decouples the global update from client synchronization. The server applies each client update immediately as it arrives, using a \textbf{staleness discount factor} to downweight updates from clients that are many rounds behind the current global model:

\begin{equation}
  \theta^{(t+1)} = \theta^{(t)} - \eta \cdot \alpha(\tau_i) \cdot \tilde{\mathbf{g}}_i,
  \qquad \alpha(\tau) = \frac{1}{(1 + \tau)^{\beta}}
  \label{eq:async_fl}
\end{equation}

where $\tau_i$ is the client's \textbf{staleness} (number of global rounds the client's local model lags behind) and $\beta \in [0.5, 1.0]$ is the staleness decay parameter. Under convex loss functions and bounded staleness, AsyncFL converges at rate $O(1/\sqrt{T})$ — identical to synchronous FL \cite{xie2019}.

\begin{description}
  \item[Straggler handling:] Clients with intermittent connectivity submit when their connection is available; the server never blocks waiting for a quorum.
  \item[Privacy accounting:] Each asynchronous update still carries its own DP noise; the global $\varepsilon$ budget accumulates over received updates, not wall-clock time.
  \item[Fairness:] A heavily discounted stale update cannot unfairly dominate the global model, preserving convergence even when a single high-bandwidth client floods the server.
\end{description}

\subsection{Gradient Compression for Low-Bandwidth Links}
\label{subsec:gradient_compression}

Even DP-noised gradients for a modest ResNet-18 backbone exceed 40\,MB per round --- prohibitive for rural 4G uplinks. Three complementary compression strategies reduce this to under 500\,KB while preserving model quality:

\begin{table}[ht]
\centering
\caption{Gradient compression techniques for low-bandwidth ASEAN deployments. All techniques are compatible with DP-SGD; apply compression \emph{before} the DP noise injection step to avoid amplifying quantization error.}
\label{tab:gradient_compression}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{p{3.2cm} p{4.0cm} p{2.0cm} p{2.4cm}}
\hline\noalign{\smallskip}
\textbf{Technique} & \textbf{Mechanism} & \textbf{Compression Ratio} & \textbf{Accuracy Cost} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Top-$k$ Sparsification & Transmit only the $k$ largest-magnitude elements; zero the rest & $100\times$ & Low ($<$1\% AUC) \\[3pt]
Random $k$ Dropout & Randomly select $k$ coordinates; rescale by $d/k$ for unbiasedness & $100\times$ & Minimal \\[3pt]
Scalar Quantization & Represent 32-bit floats as 8-bit integers with learned scale factor & $4\times$ & Negligible \\[3pt]
PowerSGD & Low-rank matrix factorization $G \approx PQ^\top$ of the gradient tensor & $10$--$50\times$ & Low \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\begin{svgraybox}
\textbf{ASEAN Deployment Pattern (Vietnam T\^ay Nguy\^en Highlands):} For rural commune health centers with intermittent 4G connectivity, the recommended stack is \textbf{AsyncFL + Top-$k$ Sparsification ($k{=}0.01{\cdot}d$) + DP-SGD ($\varepsilon{=}3.0$, $\sigma{=}1.1$)}. This reduces uplink bandwidth to $\sim$400\,KB per round for a ResNet-18 backbone --- within the free-tier data budget of a standard Vietnamese SIM card. The staleness discount with $\beta{=}0.7$ handles connectivity gaps typical of mountainous terrain without destabilizing global convergence.
\end{svgraybox}

\section{Conclusion}
\label{sec:conclusion_fl}

Federated Learning provides the architectural scaffolding for decentralized trust, but it requires the mathematical shield of DP and the cryptographic protection of \textbf{Secure Aggregation} to be fully secure. In low-connectivity ASEAN environments, AsyncFL and gradient compression are not optional optimizations --- they are prerequisites for a system that works in production. In the next chapter, we look at how to protect the update process itself from a curious or compromised central server.


