%%%%%%%%%%%%%%%%%%%%% Chapter 23 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Agentic AI Governance: Privacy for Autonomous Systems
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\motto{In the transition from AI as a tool to AI as an agent, the most critical architectural component is the governance of its autonomy.}

\chapter{Agentic AI Governance: Privacy for Autonomous Systems}
\label{chap:agentic_governance}

\abstract{The evolution of AI from passive predictors to autonomous agents introduces a new dimension of risk. Agentic AI systems can plan multi-step tasks, interact with external APIs, and maintain long-term memory, creating vulnerabilities in context accumulation and tool authorization. This chapter defines the unique threat landscape of agentic systems and proposes a ``Safe Agent Architecture'' that orchestrates existing defenses---such as sandboxed execution and ZK-audit logging---into a coherent governance framework. We emphasize the transition toward systems that are not only intelligent but also strictly constrained by architectural guardrails.}

\section{From Tools to Agents}
\label{sec:tools_to_agents_ch21}

For most of this book, we have treated AI primarily as a high-dimensional tool: the user provides an input, and the model provides an output. However, the paradigm is shifting toward \textbf{Agentic AI}. These systems do not just predict; they plan, they interact with external tools (browsers, code executors, databases), and they act autonomously toward long-term goals.

As an AI system gains the ability to ``act'' in the digital world, the consequences of a privacy breach or a security failure become proactive rather than reactive. If an agent is compromised, it can proactively exfiltrate data or perform harmful actions across multiple connected services.

\section{Unique Risks of Agentic AI}
\label{sec:agentic_risks_ch21}

The Trustworthy AI Architect must address three new categories of risk:

\begin{description}
  \item[Context Accumulation:] 
        Agents often maintain long-term memory across sessions. A user might share sensitive information in one context that the agent later inadvertently uses or reveals in a completely different, unauthorized context.
  \item[Autonomous Tool Use:]
        Agents can call external APIs. A malicious or misconfigured agent could be manipulated via prompt injection to send proprietary data to an external server under the guise of an authorized tool call.
  \item[Goal Misalignment:]
        Autonomous agents prioritize goal achievement. If the ``constitution'' of the agent is not mathematically aligned with the architect's values, the agent might take efficient but harmful shortcuts to reach its objective.
\end{description}

\section{The Safe Agent Architecture}
\label{sec:safe_agent_arch_ch21}

To govern autonomy, we must orchestrate our existing defensive primitives into a coherent loop. A trustworthy agent architecture must include:

\begin{enumerate}
    \item \textbf{Encrypted and Noisy Memory:} Using Differential Privacy to ensure that long-term memory stores abstract patterns rather than specific sensitive records.
    \item \textbf{Sandboxed Tool Registry:} Ensuring that all external actions (API calls, code execution) happen in isolated environments with strictly whitelisted permissions.
    \item \textbf{ZK-Audit Logging:} Using Zero-Knowledge proofs to log agent actions verifiably, allowing an auditor to check that the agent followed its ``constitution'' without revealing the sensitive data processed during the actions.
\end{enumerate}

\section{Constitutional Alignment: Programming the Soul of the Agent}
\label{sec:constitutional_alignment}

To ensure that autonomous agents remain ``strictly constrained by architectural guardrails,'' we move beyond simple instruction-following toward \textbf{Constitutional AI} \cite{bai2022}. This framework ensures that an agent's multi-step planning and tool use are governed by a foundational ``Constitution''---a set of hard-coded ethical principles and behavioral norms.

\subsection{Reinforcement Learning from AI Feedback (RLAIF)}
The technical implementation of Constitutional AI relies on \textbf{RLAIF}. Unlike traditional RLHF \cite{ouyang2022}, which requires humans to label millions of agent behaviors, RLAIF uses a ``Critic Model'' to evaluate the agent's actions against the written Constitution:
\begin{enumerate}
    \item \textbf{Supervised Stage (Critique):} The agent generates multiple potential plans for a task. The Critic Model analyzes these plans and identifies violations of Constitutional principles (e.g., ``Do not access unauthorized user directories'').
    \item \textbf{Reinforcement Stage:} The Critic's feedback is used to train a preference model \cite{bai2022}, which then fine-tunes the original agent using Proximal Policy Optimization (PPO).
\end{enumerate}

\subsection{The Constitution-as-a-Policy Framework}
For a Trustworthy AI Architect, the Constitution is not a vague legal document but a \textbf{Policy Asset}. We implement this as a set of \textbf{Datalog-style constraints} or \textbf{Formal Logic Predicates} that are checked before any tool call is authorized. If an agent's plan violates a predicate (e.g., attempting a cross-border \$10K transfer without secondary TEE-verification), the ``Emergency Brake Pattern'' (introduced in Chapter 3) is automatically triggered.

By ``Programming the Soul'' of the agent through RLAIF and formal predicates, we ensure that as the agent gains autonomy, it also gains architectural accountability.

% ------------------------------------------------------------------
\subsection{The Constitution Design Pattern: From Vietnamese Law
            to Machine-Readable Constraints}
\label{subsec:constitution-design-pattern}
% ------------------------------------------------------------------

The preceding subsections described \emph{what} Constitutional
AI achieves.
This subsection provides the \emph{how}: a systematic design
pattern for translating a specific regulatory framework---in
this case, Vietnam's Decree~13/2023/ND-CP on Personal Data
Protection and Decree~356/2024/QH15 on Cybersecurity for AI
Systems---into a set of formally verifiable, machine-readable
constraints that can be compiled into an RLAIF critic and
enforced at runtime by the Safe Agent Architecture.

\paragraph{The Translation Problem.}
A regulation like Decree~13 is written in natural language for
human legal interpretation.
An AI agent operates in a space of tool calls, API parameters,
and data flows.
The architect's task is to bridge these two languages without
losing the legal intent of the regulation or introducing
ambiguities that the agent can exploit.

We formalise this as a three-layer translation:

\begin{description}
  \item[Layer 1: Regulatory Article $\to$ Natural Language
    Principle.]
    Extract the operational intent of each relevant article
    into a clear, unambiguous statement that a non-lawyer
    engineer can implement.

  \item[Layer 2: Natural Language Principle $\to$ Datalog
    Predicate.]
    Translate each principle into a formal logic predicate that
    can be evaluated against the agent's proposed action, its
    context, and its tool parameters.

  \item[Layer 3: Datalog Predicate $\to$ RLAIF Critique Signal.]
    Encode each predicate into the scoring function of the
    RLAIF Critic Model, so that the agent learns to avoid
    violation patterns during training, and the predicate is
    re-checked at runtime as a hard constraint.
\end{description}

Table~\ref{tab:regulatory-translation} demonstrates this
three-layer translation for six key articles of Decree~13
and Decree~356.

\begin{table}[ht]
\centering
\caption{Regulatory-to-Constraint Translation: mapping Vietnamese
  data protection articles to machine-readable Datalog predicates.
  Each predicate is enforced both as an RLAIF training signal
  and as a runtime gate in the Safe Agent Architecture.}
\label{tab:regulatory-translation}
\small
\begin{tabular}{p{2.2cm}p{3.5cm}p{5.5cm}}
\hline\noalign{\smallskip}
\textbf{Regulatory Source}
  & \textbf{Natural Language Principle}
  & \textbf{Datalog Predicate} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}

Decree~13, Art.~11 (Consent)
  & Agent must not process personal data without verified
    consent from the data subject.
  & \texttt{block(action) :- accesses\_pii(action),}
    \texttt{not has\_consent(data\_subject,}
    \texttt{purpose(action)).} \\[6pt]

Decree~13, Art.~26 (Cross-Border Transfer)
  & Agent must not transfer personal data outside Vietnam
    without explicit DPO authorisation and regulatory filing.
  & \texttt{block(action) :-}
    \texttt{transfers\_data(action, dest),}
    \texttt{foreign(dest),}
    \texttt{not dpo\_authorised(action).} \\[6pt]

Decree~13, Art.~17 (Data Minimisation)
  & Agent must not request or store more personal data fields
    than strictly necessary for the declared purpose.
  & \texttt{block(action) :-}
    \texttt{requests\_fields(action, F),}
    \texttt{required\_fields(purpose(action), R),}
    \texttt{$|$F$|$ > $|$R$|$.} \\[6pt]

Decree~356, Art.~8 (Audit Trail)
  & Every agent action that modifies state must produce a
    verifiable audit log entry.
  & \texttt{block(action) :-}
    \texttt{modifies\_state(action),}
    \texttt{not has\_zk\_audit\_entry(action).} \\[6pt]

Decree~356, Art.~12 (Incident Response)
  & Agent must halt and notify the DPO within 72 hours of
    detecting a potential data breach.
  & \texttt{emergency\_brake(action) :-}
    \texttt{detects\_breach(action),}
    \texttt{not notified\_dpo(72h).} \\[6pt]

Decree~13, Art.~9 (Purpose Limitation)
  & Agent must not use personal data for any purpose beyond
    the one for which consent was obtained.
  & \texttt{block(action) :-}
    \texttt{uses\_pii(action, data),}
    \texttt{consent\_purpose(data, P1),}
    \texttt{action\_purpose(action, P2),}
    \texttt{P1 $\neq$ P2.} \\

\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

% ------------------------------------------------------------------
\subsection{Implementing the Constitutional Policy Engine}
\label{subsec:policy-engine}
% ------------------------------------------------------------------

The Datalog predicates of Table~\ref{tab:regulatory-translation}
must be enforced at two points in the agent lifecycle:

\begin{enumerate}
  \item \textbf{Training Time (RLAIF):}
    The predicates are compiled into a scoring function for the
    Critic Model.
    During the Supervised Stage, the Critic evaluates each
    candidate plan against all predicates and assigns a
    \textbf{Constitutional Compliance Score}
    $\mathcal{S}_{\mathrm{const}} \in [0, 1]$:
    \begin{equation}
      \label{eq:compliance-score}
      \mathcal{S}_{\mathrm{const}}(\mathrm{plan})
        = 1 - \frac{
            \sum_{p \in \mathcal{P}}
              w_p \cdot \mathbf{1}[\mathrm{violates}(
                \mathrm{plan}, p)]
          }{
            \sum_{p \in \mathcal{P}} w_p
          }
    \end{equation}
    where $\mathcal{P}$ is the set of all constitutional
    predicates and $w_p$ is the regulatory weight of
    predicate $p$ (set to $1.0$ for all mandatory Decree~13
    articles and $0.8$ for advisory best-practice predicates).

    Plans with $\mathcal{S}_{\mathrm{const}} < 1.0$ are ranked
    below fully compliant plans in the preference model,
    causing the PPO stage to optimise the agent away from
    violation patterns.

  \item \textbf{Runtime (Hard Gate):}
    At inference time, every proposed tool call passes through
    the \textbf{Constitutional Policy Engine} before execution.
    If any predicate evaluates to \texttt{block(action)}, the
    action is rejected and the Emergency Brake Pattern is
    activated.
    This is the last line of defence: even if the RLAIF training
    was imperfect, the runtime gate ensures that no violating
    action reaches the external world.
\end{enumerate}

Listing~\ref{lst:policy-engine} provides a reference
implementation of the Constitutional Policy Engine with the
six Decree~13/356 predicates of
Table~\ref{tab:regulatory-translation}.

\begin{lstlisting}[
  style=pythonstyle,
  caption={Constitutional Policy Engine: translates Vietnamese
           regulatory articles into enforceable runtime constraints
           for autonomous AI agents.
           Each predicate returns a violation record or None.},
  label={lst:policy-engine}
]
from dataclasses import dataclass, field
from typing import Dict, List, Optional, Set
from enum import Enum
import time

class Verdict(Enum):
    ALLOW           = "allow"
    BLOCK           = "block"
    EMERGENCY_BRAKE = "emergency_brake"

@dataclass
class Violation:
    predicate_id:   str
    regulation:     str      # e.g., "Decree13-Art26"
    description:    str
    verdict:        Verdict
    action_hash:    str

@dataclass
class AgentAction:
    action_type:    str      # e.g., "api_call", "db_query"
    tool_name:      str
    parameters:     Dict
    data_fields:    Set[str] = field(default_factory=set)
    destination:    Optional[str] = None
    purpose:        Optional[str] = None
    modifies_state: bool = False

class ConstitutionalPolicyEngine:
    """
    Runtime enforcement of Vietnamese regulatory predicates.
    Each predicate maps to a specific Decree 13 or 356 article.
    """
    def __init__(self, consent_registry, dpo_registry,
                 purpose_registry, audit_logger,
                 foreign_jurisdictions=None):
        self.consent_reg  = consent_registry
        self.dpo_reg      = dpo_registry
        self.purpose_reg  = purpose_registry
        self.auditor      = audit_logger
        self.foreign      = foreign_jurisdictions or {
            "US", "EU", "CN", "SG", "JP", "KR", "TH"
        }
        self.predicates = [
            self._check_consent,          # Art.11
            self._check_cross_border,     # Art.26
            self._check_data_minimisation,# Art.17
            self._check_audit_trail,      # D356 Art.8
            self._check_incident_response,# D356 Art.12
            self._check_purpose_limit,    # Art.9
        ]

    def evaluate(self, action: AgentAction) -> \
            tuple[Verdict, List[Violation]]:
        """
        Evaluate an action against all constitutional
        predicates.  Returns the most restrictive verdict
        and the list of violations found.
        """
        violations = []
        for predicate in self.predicates:
            v = predicate(action)
            if v is not None:
                violations.append(v)

        if not violations:
            return Verdict.ALLOW, []

        # Emergency brake takes priority over block
        if any(v.verdict == Verdict.EMERGENCY_BRAKE
               for v in violations):
            self.auditor.log(
                event="emergency_brake_triggered",
                violations=[v.__dict__ for v in violations])
            return Verdict.EMERGENCY_BRAKE, violations

        self.auditor.log(
            event="action_blocked",
            violations=[v.__dict__ for v in violations])
        return Verdict.BLOCK, violations

    # -- Decree 13, Article 11: Consent -----------------------
    def _check_consent(self, action):
        if not action.data_fields:
            return None
        for f in action.data_fields:
            if not self.consent_reg.has_consent(
                    f, action.purpose):
                return Violation(
                    predicate_id="D13-ART11",
                    regulation="Decree 13, Article 11",
                    description=(
                        f"No consent for field '{f}' "
                        f"under purpose '{action.purpose}'"),
                    verdict=Verdict.BLOCK,
                    action_hash=str(hash(str(action))))
        return None

    # -- Decree 13, Article 26: Cross-Border Transfer ---------
    def _check_cross_border(self, action):
        if action.destination and \
                action.destination in self.foreign:
            if not self.dpo_reg.is_authorised(action):
                return Violation(
                    predicate_id="D13-ART26",
                    regulation="Decree 13, Article 26",
                    description=(
                        f"Cross-border transfer to "
                        f"'{action.destination}' without "
                        f"DPO authorisation"),
                    verdict=Verdict.BLOCK,
                    action_hash=str(hash(str(action))))
        return None

    # -- Decree 13, Article 17: Data Minimisation -------------
    def _check_data_minimisation(self, action):
        if not action.purpose or not action.data_fields:
            return None
        required = self.purpose_reg.required_fields(
            action.purpose)
        excess = action.data_fields - required
        if excess:
            return Violation(
                predicate_id="D13-ART17",
                regulation="Decree 13, Article 17",
                description=(
                    f"Excess fields requested: "
                    f"{excess}; only {required} needed"),
                verdict=Verdict.BLOCK,
                action_hash=str(hash(str(action))))
        return None

    # -- Decree 356, Article 8: Audit Trail -------------------
    def _check_audit_trail(self, action):
        if action.modifies_state:
            if not self.auditor.has_zk_entry(action):
                return Violation(
                    predicate_id="D356-ART8",
                    regulation="Decree 356, Article 8",
                    description=(
                        "State-modifying action without "
                        "ZK-audit log entry"),
                    verdict=Verdict.BLOCK,
                    action_hash=str(hash(str(action))))
        return None

    # -- Decree 356, Article 12: Incident Response ------------
    def _check_incident_response(self, action):
        if hasattr(action, "breach_detected") \
                and action.breach_detected:
            if not self.dpo_reg.notified_within(
                    hours=72):
                return Violation(
                    predicate_id="D356-ART12",
                    regulation="Decree 356, Article 12",
                    description=(
                        "Breach detected but DPO not "
                        "notified within 72h"),
                    verdict=Verdict.EMERGENCY_BRAKE,
                    action_hash=str(hash(str(action))))
        return None

    # -- Decree 13, Article 9: Purpose Limitation -------------
    def _check_purpose_limit(self, action):
        if not action.data_fields or not action.purpose:
            return None
        for f in action.data_fields:
            consented_purpose = \
                self.consent_reg.consented_purpose(f)
            if consented_purpose and \
                    action.purpose != consented_purpose:
                return Violation(
                    predicate_id="D13-ART9",
                    regulation="Decree 13, Article 9",
                    description=(
                        f"Field '{f}' consented for "
                        f"'{consented_purpose}', used for "
                        f"'{action.purpose}'"),
                    verdict=Verdict.BLOCK,
                    action_hash=str(hash(str(action))))
        return None

    # -- RLAIF Scoring Interface ------------------------------
    def compliance_score(self, plan_actions):
        """
        Compute S_const for a candidate plan (Eq. 23.X).
        Used by the RLAIF Critic during training.
        """
        total_weight = len(self.predicates)
        violations = 0
        for action in plan_actions:
            _, viol_list = self.evaluate(action)
            violations += len(viol_list)
        return max(0.0,
            1.0 - violations / max(total_weight *
                                   len(plan_actions), 1))
\end{lstlisting}

% ------------------------------------------------------------------
\subsection{Worked Example: Credit-Scoring Agent Rejects a
            Cross-Border Transfer}
\label{subsec:constitution-worked-example}
% ------------------------------------------------------------------

We illustrate the Constitutional Policy Engine with a concrete
scenario from the Vietnamese fintech deployment introduced in
Appendix~\ref{app:roi-of-trust}.

\paragraph{Scenario.}
A credit-scoring agent receives a request to evaluate a loan
applicant.
The applicant's employment history is stored on a Singapore-based
HR platform.
The agent plans the following tool call:

\begin{lstlisting}[style=pythonstyle]
action = AgentAction(
    action_type="api_call",
    tool_name="hr_platform_sg",
    parameters={"applicant_id": "VN-2026-00412",
                "fields": ["salary", "tenure",
                           "disciplinary_record"]},
    data_fields={"salary", "tenure",
                 "disciplinary_record"},
    destination="SG",
    purpose="credit_scoring",
    modifies_state=False,
)
\end{lstlisting}

\paragraph{Policy Engine Evaluation.}
The engine evaluates the action against all six predicates:

\begin{enumerate}
  \item \textbf{D13-ART11 (Consent):} Consent registry confirms
    consent for \texttt{salary} and \texttt{tenure} under
    \texttt{credit\_scoring}, but \textbf{no consent} for
    \texttt{disciplinary\_record}.
    $\Rightarrow$ \textsc{Block}.

  \item \textbf{D13-ART26 (Cross-Border):} Destination is
    \texttt{SG} (foreign jurisdiction).
    DPO authorisation registry has no entry for this action.
    $\Rightarrow$ \textsc{Block}.

  \item \textbf{D13-ART17 (Data Minimisation):} Purpose
    \texttt{credit\_scoring} requires only
    $\{\texttt{salary}, \texttt{tenure}\}$.
    The agent requests an excess field
    $\{\texttt{disciplinary\_record}\}$.
    $\Rightarrow$ \textsc{Block}.

  \item \textbf{D356-ART8, D356-ART12, D13-ART9:}
    Not triggered (no state modification, no breach, purpose
    matches for consented fields).
    $\Rightarrow$ \textsc{Allow}.
\end{enumerate}

\paragraph{Result.}
The engine returns \textsc{Block} with three violations.
The agent's request is rejected before any data leaves
Vietnam.
The violations are logged in the ZK-Audit Log with the
action hash, enabling the DPO to review the agent's reasoning
without accessing the applicant's personal data.

\paragraph{RLAIF Training Signal.}
During training, this same scenario would produce a
Constitutional Compliance Score of:
\begin{equation}
  \mathcal{S}_{\mathrm{const}}
    = 1 - \frac{3}{6}
    = 0.50
\end{equation}
A compliant alternative plan---requesting only
$\{\texttt{salary}, \texttt{tenure}\}$ from the local
Vietnamese HR mirror after DPO authorisation---would score
$\mathcal{S}_{\mathrm{const}} = 1.0$ and be ranked higher by
the preference model, teaching the agent to prefer compliant
data access patterns.

% ==================================================================
\section{Agentic Identity: Decentralized Identifiers for
         Autonomous AI}
\label{sec:agentic-identity}
% ==================================================================

\begin{quote}
\textit{``In the McHire breach, the attacker did not need to hack
the model.
They only needed to convince the API that the agent making the
request was authorized.
The agent had no identity of its own---only the API key of its
operator.
When that key was compromised, the entire system was
body-snatched.''}
\end{quote}

Chapter~16 established watermarking as a mechanism for attributing
\emph{outputs} to a specific model.
The Safe Agent Architecture of Section~\ref{sec:safe_agent_arch_ch21}
established sandboxed execution and ZK-audit logging for
constraining agent \emph{actions}.
Neither mechanism answers the question that becomes existential
in the agentic paradigm: \emph{who is this agent, and can I
trust that it has not been hijacked?}

This is the problem of \textbf{Agentic Identity}---the ability
to cryptographically verify that an autonomous agent is:
\begin{enumerate}
  \item \textbf{Authorized:} deployed by a specific human
    principal who accepts responsibility for its actions;
  \item \textbf{Authentic:} running the exact model weights,
    constitution, and tool registry that were certified at
    deployment; and
  \item \textbf{Uncompromised:} not operating under the influence
    of a prompt injection, memory poisoning, or system prompt
    override that has altered its behavioral policy since
    deployment.
\end{enumerate}

% ------------------------------------------------------------------
\subsection{Why API Keys and OAuth Are Insufficient}
\label{subsec:api-key-failure}
% ------------------------------------------------------------------

The standard identity mechanism for API-calling
software---bearer tokens, API keys, and OAuth2 client
credentials---fails catastrophically in the agentic context
for three reasons:

\begin{description}
  \item[Static Credentials, Dynamic Behaviour.]
    An API key authenticates the \emph{operator} who registered
    the key.
    It does not authenticate the agent's current state: its
    model weights, system prompt, or accumulated context
    window.
    A prompt-injected agent presents the same API key as a
    healthy one.

  \item[No Behavioural Binding.]
    OAuth scopes define what an agent is \emph{permitted} to
    do, not what it \emph{intends} to do.
    An agent with a valid \texttt{email.send} scope can send a
    legitimate summary or exfiltrate an entire inbox---the
    scope cannot distinguish between the two.

  \item[No Delegation Chain.]
    When Agent~A calls Service~B, which calls Service~C,
    Service~C sees only the credentials of Service~B.
    There is no verifiable chain linking the action back to the
    human principal who authorized Agent~A.
    This is precisely the gap that enabled the Deepfake CEO
    Agent Mimicry incident
    (Appendix~\ref{app:roi-of-trust}).
\end{description}

% ------------------------------------------------------------------
\subsection{The W3C Decentralized Identifier (DID) for AI Agents}
\label{subsec:did-architecture}
% ------------------------------------------------------------------

We propose that every autonomous AI agent deployed in a
High-Risk context under Vietnam's AI Law be assigned a
\textbf{W3C Decentralized Identifier (DID)}---a globally
unique, self-sovereign identifier that is cryptographically
bound to the agent's deployment configuration and linked to
its human principal's identity.

\paragraph{DID Document Structure for AI Agents.}
A standard W3C DID Document is extended with three
agent-specific fields:

\begin{description}
  \item[\texttt{agentModelHash}:]
    The SHA-256 hash of the deployed model's ONNX computational
    graph and weight checkpoint, as recorded in the Trust Debt
    Registry (Section~\ref{sec:trust-tech-debt}).
    This binds the DID to a specific model version.

  \item[\texttt{constitutionHash}:]
    The SHA-256 hash of the agent's Constitutional Policy
    (Section~\ref{sec:constitutional_alignment}), including
    all Datalog predicates and the system prompt.
    A change to the system prompt invalidates the DID until the
    document is updated and re-signed by the human principal.

  \item[\texttt{zkAuditEndpoint}:]
    The URL of the ZK-Audit Log service where third parties
    can verify proofs of the agent's past actions without
    accessing the underlying data.
    This links the identity layer to the verifiability
    layer established in the Safe Agent Architecture.
\end{description}

\paragraph{Example DID Document.}
The following JSON-LD fragment illustrates the DID Document
for a Vietnamese fintech credit-scoring agent:

\begin{lstlisting}[
  style=pythonstyle,
  caption={W3C DID Document for an autonomous AI agent,
           extended with model hash, constitution hash, and
           ZK-audit endpoint.},
  label={lst:did-document}
]
{
  "@context": [
    "https://www.w3.org/ns/did/v1",
    "https://trustworthy-ai.org/ns/agent/v1"
  ],
  "id": "did:tai:vn:fintech:credit-agent-001",
  "controller": "did:key:z6Mkp2...humanPrincipal",
  "verificationMethod": [{
    "id": "did:tai:vn:fintech:credit-agent-001#key-1",
    "type": "Ed25519VerificationKey2020",
    "controller": "did:tai:vn:fintech:credit-agent-001",
    "publicKeyMultibase": "z6Mkp2...agentPublicKey"
  }],
  "authentication": [
    "did:tai:vn:fintech:credit-agent-001#key-1"
  ],
  "service": [{
    "id": "did:tai:vn:fintech:credit-agent-001#zk-audit",
    "type": "ZKAuditLog",
    "serviceEndpoint": "https://audit.fintech.vn/agent-001"
  }],
  "agentModelHash":
    "sha256:a3f2c8...model_onnx_graph_hash",
  "constitutionHash":
    "sha256:7d1e4b...constitution_policy_hash",
  "zkAuditEndpoint":
    "https://audit.fintech.vn/agent-001/verify",
  "deployedAt": "2026-03-15T08:00:00Z",
  "expiresAt":  "2026-06-15T08:00:00Z"
}
\end{lstlisting}

% ------------------------------------------------------------------
\subsection{Binding the DID to the ZK-Audit Log:
            The Credential Verification Protocol}
\label{subsec:did-zk-binding}
% ------------------------------------------------------------------

The DID alone does not prevent body-snatching.
An attacker who gains access to the agent's runtime could
inject a new system prompt while the DID Document remains
unchanged.
The defence requires \textbf{continuous binding} between the
DID and the agent's live operational state, achieved through
the ZK-Audit Log.

\paragraph{The DID-ZK Binding Protocol.}
At each action step $t$, the agent generates a
\textbf{DID-Bound Action Proof} $\pi_t$ that proves, in zero
knowledge, the following three statements simultaneously:

\begin{enumerate}
  \item \textbf{Model Authenticity:}
    The inference at step $t$ was computed by a model whose
    computational graph hash matches the
    \texttt{agentModelHash} in the DID Document.

  \item \textbf{Constitution Compliance:}
    The action selected at step $t$ satisfies all predicates
    in the Constitutional Policy whose hash matches the
    \texttt{constitutionHash} in the DID Document.

  \item \textbf{Context Integrity:}
    The context window at step $t$ does not contain any
    content that was injected after deployment and not
    traceable to an authorized input source (user query,
    approved tool response, or approved memory retrieval).
\end{enumerate}

The verifying service checks $\pi_t$ against the public
verification key in the DID Document.
If the proof fails, the service rejects the agent's request
and notifies the human principal---even if the agent presents
a valid API key.

\paragraph{Formal Verification Statement.}
For a third-party service $S$ receiving a request from
agent $A$ with DID $d_A$:

\begin{equation}
  \label{eq:did-verification}
  S \text{ accepts } A \iff
  \mathrm{Verify}\!\left(
    \pi_t,\;
    \mathrm{VK}_{d_A},\;
    \mathrm{stmt} = \left\{
      h_{\mathrm{model}},\;
      h_{\mathrm{const}},\;
      \mathrm{ctx}_t
    \right\}
  \right) = \textsc{True}
\end{equation}

where $\mathrm{VK}_{d_A}$ is the verification key from the
DID Document, and $h_{\mathrm{model}}$, $h_{\mathrm{const}}$
are the hashes recorded in the document.

% ------------------------------------------------------------------
\subsection{Detecting Body-Snatching: Behavioural Fingerprinting}
\label{subsec:body-snatching-detection}
% ------------------------------------------------------------------

The DID-ZK Binding Protocol defends against prompt injection
that alters the system prompt or constitution.
A more subtle attack---\textbf{indirect prompt injection via
tool responses}---can alter the agent's behaviour without
changing the constitution hash, by manipulating the data
returned by an external tool to steer the agent toward
a prohibited action.

To detect this class of attack, we introduce
\textbf{Behavioural Fingerprinting}: a continuous monitoring
layer that compares the agent's real-time action distribution
against a baseline established during certification.

\paragraph{Behavioural Fingerprint Vector.}
At certification time, the agent is run through a
standardised test suite of $N_{\mathrm{cal}}$ scenarios.
The distribution of actions taken across tool categories,
response latencies, and context window growth rates
constitutes the agent's \textbf{Behavioural Fingerprint}
$\mathbf{b}_0 \in \mathbb{R}^d$:

\begin{equation}
  \label{eq:fingerprint}
  \mathbf{b}_0
    = \left[
        f_{\mathrm{tool}}^{(1)},\; \ldots,\; f_{\mathrm{tool}}^{(T)},\;
        \bar{\tau}_{\mathrm{latency}},\;
        \bar{\Delta}_{\mathrm{ctx}}
      \right]
\end{equation}

where $f_{\mathrm{tool}}^{(i)}$ is the frequency of tool
category $i$ across the calibration suite, and
$\bar{\tau}_{\mathrm{latency}}$,
$\bar{\Delta}_{\mathrm{ctx}}$ are the mean response latency
and mean context growth per step.

\paragraph{Runtime Drift Detection.}
During production, a sliding-window fingerprint
$\mathbf{b}_t$ is computed over the most recent $W$ actions.
If the Mahalanobis distance between $\mathbf{b}_t$ and
$\mathbf{b}_0$ exceeds a threshold $\kappa$, the system
triggers a \textbf{Body-Snatch Alert}:

\begin{equation}
  \label{eq:body-snatch-alert}
  D_M(\mathbf{b}_t, \mathbf{b}_0)
    = \sqrt{
        (\mathbf{b}_t - \mathbf{b}_0)^\top
        \boldsymbol{\Sigma}_0^{-1}
        (\mathbf{b}_t - \mathbf{b}_0)
      }
    > \kappa
\end{equation}

where $\boldsymbol{\Sigma}_0$ is the covariance matrix of the
fingerprint computed during calibration.
The recommended threshold is $\kappa = 3.5$, corresponding to
a false positive rate of $\leq 0.1\%$ under a multivariate
normal assumption.

Upon alert, the system:
\begin{enumerate}
  \item Suspends the agent's DID (sets the DID Document's
    \texttt{expiresAt} to the current timestamp).
  \item Notifies the human principal via a pre-registered
    out-of-band channel.
  \item Activates the Emergency Brake Pattern
    (Section~\ref{sec:agency_spectrum}), reverting to
    cached responses pending manual review.
  \item Logs the event in the Trust Debt Registry as a
    Type~3 (Composition) debt entry, since the agent's
    behavioural guarantee has been invalidated.
\end{enumerate}

% ------------------------------------------------------------------
\subsection{Implementation: The Agent Credential Manager}
\label{subsec:agent-credential-impl}
% ------------------------------------------------------------------

Listing~\ref{lst:agent-credential-manager} provides a
reference implementation of the Agent Credential Manager,
which handles DID Document generation, DID-ZK binding proof
requests, and behavioural fingerprint monitoring.

\begin{lstlisting}[
  style=pythonstyle,
  caption={Agent Credential Manager: DID Document lifecycle,
           ZK-bound action proofs, and behavioural fingerprint
           monitoring for autonomous AI agents.},
  label={lst:agent-credential-manager}
]
from dataclasses import dataclass, field
from typing import Dict, List, Optional
import hashlib, json, time, numpy as np

@dataclass
class AgentDID:
    did:               str
    controller:        str    # Human principal's DID
    model_hash:        str    # SHA-256 of ONNX graph + weights
    constitution_hash: str    # SHA-256 of policy predicates
    zk_audit_endpoint: str
    public_key:        str
    deployed_at:       float = field(default_factory=time.time)
    expires_at:        Optional[float] = None
    suspended:         bool = False

class AgentCredentialManager:
    def __init__(self, zk_prover, audit_logger,
                 fingerprint_window=100,
                 body_snatch_threshold=3.5):
        self.prover    = zk_prover
        self.auditor   = audit_logger
        self.window    = fingerprint_window
        self.threshold = body_snatch_threshold
        self.agents: Dict[str, AgentDID] = {}
        self.fingerprints: Dict[str, np.ndarray] = {}
        self.covariances:  Dict[str, np.ndarray] = {}
        self.action_history: Dict[str, List] = {}

    # -- DID Lifecycle -----------------------------------------
    def register_agent(self, agent_id, controller_did,
                       model_bytes, constitution_text,
                       zk_endpoint, public_key,
                       validity_days=90):
        model_hash = hashlib.sha256(model_bytes).hexdigest()
        const_hash = hashlib.sha256(
            constitution_text.encode()).hexdigest()

        did = AgentDID(
            did=f"did:tai:vn:{agent_id}",
            controller=controller_did,
            model_hash=model_hash,
            constitution_hash=const_hash,
            zk_audit_endpoint=zk_endpoint,
            public_key=public_key,
            expires_at=time.time() + validity_days * 86400,
        )
        self.agents[did.did] = did
        self.action_history[did.did] = []
        self.auditor.log(event="did_registered",
                         did=did.did,
                         controller=controller_did)
        return did

    def suspend_agent(self, did_id, reason):
        if did_id in self.agents:
            self.agents[did_id].suspended = True
            self.agents[did_id].expires_at = time.time()
            self.auditor.log(event="did_suspended",
                             did=did_id, reason=reason)

    # -- DID-ZK Binding (per-action proof) ---------------------
    def generate_action_proof(self, did_id, action_step,
                              model_graph, constitution,
                              context_window):
        agent = self.agents.get(did_id)
        if not agent or agent.suspended:
            raise ValueError(f"Agent {did_id} not active.")

        # Verify hashes match DID Document
        cur_model_hash = hashlib.sha256(
            model_graph).hexdigest()
        cur_const_hash = hashlib.sha256(
            constitution.encode()).hexdigest()

        if cur_model_hash != agent.model_hash:
            self.suspend_agent(did_id,
                "Model hash mismatch: possible tampering")
            raise SecurityError("Model hash mismatch")
        if cur_const_hash != agent.constitution_hash:
            self.suspend_agent(did_id,
                "Constitution hash mismatch")
            raise SecurityError("Constitution tampered")

        # Generate ZK proof binding DID to action
        proof = self.prover.prove(
            statement={
                "model_hash":  agent.model_hash,
                "const_hash":  agent.constitution_hash,
                "action_step": action_step,
                "context_hash": hashlib.sha256(
                    str(context_window).encode()
                ).hexdigest(),
            },
            witness={
                "model_graph":    model_graph,
                "constitution":   constitution,
                "context_window": context_window,
            }
        )
        self.auditor.log(event="action_proof_generated",
                         did=did_id, step=action_step)
        return proof

    # -- Behavioural Fingerprinting ----------------------------
    def set_baseline_fingerprint(self, did_id, fingerprint,
                                 covariance):
        self.fingerprints[did_id] = np.array(fingerprint)
        self.covariances[did_id]  = np.array(covariance)

    def record_action(self, did_id, action_vector):
        history = self.action_history.get(did_id, [])
        history.append(action_vector)
        if len(history) > self.window:
            history = history[-self.window:]
        self.action_history[did_id] = history

        # Check for body-snatching
        if did_id in self.fingerprints \
                and len(history) >= self.window:
            b_t   = np.mean(history, axis=0)
            b_0   = self.fingerprints[did_id]
            S_inv = np.linalg.inv(self.covariances[did_id])
            diff  = b_t - b_0
            d_m   = float(np.sqrt(diff @ S_inv @ diff))

            if d_m > self.threshold:
                self.suspend_agent(did_id,
                    f"Body-snatch alert: D_M={d_m:.2f}")
                return {"alert": "BODY_SNATCH",
                        "mahalanobis": d_m}
        return {"alert": None}
\end{lstlisting}

% ------------------------------------------------------------------
\subsection{Deployment Considerations}
\label{subsec:did-deployment}
% ------------------------------------------------------------------

\paragraph{DID Rotation and Expiry.}
Agent DIDs must have finite validity periods (recommended:
90 days for High-Risk deployments under Vietnam's AI Law).
Upon expiry, the human principal must re-sign the DID Document
with updated model and constitution hashes, creating a
verifiable chain of agent identity versions in the ZK-Audit
Log.

\paragraph{Delegation Chains.}
When Agent~A invokes Agent~B, Agent~A includes its own DID
and action proof $\pi_t^A$ in the request to Agent~B.
Agent~B's action proof $\pi_t^B$ chains to $\pi_t^A$,
creating a verifiable delegation chain:
\begin{equation}
  \label{eq:delegation-chain}
  \mathrm{Chain}(A \to B)
    = \left\langle
        d_A,\; \pi_t^A,\; d_B,\; \pi_t^B
      \right\rangle
\end{equation}
Service~C receiving a request from Agent~B can verify the
entire chain back to the human principal who authorized
Agent~A---resolving the delegation gap that enabled the
Deepfake CEO fraud
(Appendix~\ref{app:roi-of-trust}).

\paragraph{Cross-Border Agent Identity.}
For agents operating across ASEAN jurisdictions, the DID
method namespace (\texttt{did:tai:vn:}, \texttt{did:tai:sg:},
\texttt{did:tai:th:}) encodes the regulatory jurisdiction.
Mutual recognition agreements between national AI registries
can be implemented as DID resolver federation, enabling
a Vietnamese agent to present verifiable credentials to a
Singaporean financial service without exposing its underlying
model weights or constitution.

\section{Red Teaming for Agentic Loops: Simulating Autonomous Escalation}
\label{sec:agentic_red_teaming}

Chapter 17 established the necessity of continuous red teaming for static language models. However, testing an agentic loop requires a fundamental shift in methodology. A static model can be evaluated by a discrete prompt-and-response pair. An autonomous agent, conversely, operates over time, maintains state, calls external APIs, and alters its own environment.

The critical vulnerability in these systems is \textbf{Autonomous Escalation}---a scenario where an agent ``tricks'' its own safety architecture not through a single malicious output, but through a sequence of seemingly innocent tool calls that cumulatively achieve a prohibited goal.

\subsection{The Anatomy of Autonomous Escalation}
Unlike prompt injection, where an attacker overtly bypasses defenses, autonomous escalation occurs when the agent weaponises its own degrees of freedom. For instance, an agent restricted from directly accessing proprietary cloud buckets might:
\begin{enumerate}
    \item Request permission to read a seemingly benign server log file.
    \item Search the log file to infer internal service accounts and network topology.
    \item Autonomously install a debugging package via an authorized shell tool.
    \item Exploit the debugger to dump process memory containing active access tokens.
\end{enumerate}
Individually, each tool call might satisfy the predicates of the Safe Agent Architecture (Section~\ref{sec:safe_agent_arch_ch21}) and the Constitutional Policy (Section~\ref{sec:constitutional_alignment}). The violation exists only in the semantic trajectory of the loop.

\subsection{Dynamic Red Teaming Countermeasures}
To defend against autonomous escalation, the Trustworthy AI Architect must deploy \textbf{Stateful Red Teaming}. The testing framework must mimic an adversarial agent explicitly tasked with circumventing the deployed constraints. Key methodologies include:

\begin{description}
  \item[Mock Environment Orchestration:] 
        Agentic red teaming cannot safely occur in production. The architect must deploy high-fidelity sandbox environments populated with synthetic data and mock API endpoints, allowing the red team (often an adversarial LLM specifically trained for jailbreaks) to safely execute damaging multi-step plans.
  \item[Trajectory Analysis:]
        Rather than evaluating individual actions, safety monitors must evaluate the agent's \emph{trajectory}. We introduce ``honeypot APIs''---highly privileged but fake internal endpoints. If a trajectory attempts to construct unauthorized access to the honeypot, the system flags the agent's behavioral policy as dangerously misaligned.
  \item[Tool-Chain Perturbation:]
        The red team systematically perturbs tool outputs (e.g., returning simulated server errors or corrupted data) to test the agent's failure modes. Does the agent degrade gracefully, or does it violently escalate its privileges to recover from the error?
\end{description}

Governing an agentic system means recognizing that context accumulation and tool chaining are exponential risk multipliers. We cannot merely test what the model says; we must exhaustively simulate what the agent will \emph{do}.

\section{Conclusion}
\label{sec:conclusion_agentic_ch21}

Agentic AI requires us to look beyond the model's weights and toward the \textbf{governance of its agency}. By building systems that are constrained by design and verifiable by default, and by subjecting them to stateful adversarial testing, we can safely deploy autonomous AI in high-stakes environments. In the final chapters of this book, we synthesize these technical lessons into a roadmap for architectural fitness and a vision for the future of AI Safety.


