%%%%%%%%%%%%%%%%%%%%% Chapter 1 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The Vision: Why Trustworthy AI Matters Now
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\motto{In the age of autonomous agency, the defining question is no longer "What can the model say?" but "What is the agent allowed to do?"}

\chapter{The Vision: Why Trustworthy AI Matters Now}
\label{chap:vision}

\abstract{The era of AI experimentation is over. By 2026, the industry has moved past the "Generation Hype" of the early 2020s and entered the era of autonomous agency and total accountability. This chapter establishes the high-stakes reality of the mission-critical AI Architect. We analyze the shift from chatbots to Large Action Models (LAMs), learn from the catastrophic trust failures of 2025, and map the 2026 regulatory landscape---including Vietnam's landmark AI Law. Finally, we define the four pillars of the modern Trustworthy AI Stack: Security, Privacy, Verifiability, and Alignment.}

\section{The Great Shift: From Generation to Agency}
\label{sec:generation_to_agency}

In 2023, the world was mesmerized by the ability of Large Language Models to generate text. By 2026, however, the architectural focus has shifted to \textbf{Large Action Models (LAMs)}. We have bypassed Chatbots in favor of autonomous agents that execute financial transactions, manage complex APIs, and orchestrate critical infrastructure.

This shift has created a \textbf{Visibility Crisis}. Currently, nearly 50\% of organizations cannot effectively monitor "Machine-to-Machine" (M2M) autonomous traffic. When an agent acts on behalf of a human, the traditional UI-based trust models collapse. Trust must now be built directly into the execution layer, not just the interface.

\section{The ``Lessons Learned'' from the 2025 Crisis}
\label{sec:crisis_2025}

The urgency of this transition is grounded in the "Great Audit" of 2025. This year saw high-profile failures that permanently eroded public confidence:
\begin{itemize}
    \item \textbf{The McHire Security Breach:} An autonomous recruitment agent was manipulated via prompt injection to leak the PII of two million applicants.
    \item \textbf{Deepfake-Enabled Corporate Fraud:} A CEO's agent was mimicked to authorize a \$50 million cross-border transfer, highlighting the failure of unverified agency.
    \item \textbf{Hallucination-in-Healthcare:} Miscalibrated diagnostic agents emphasized accuracy over safety, leading to unrepresentative treatment recommendations for minority groups.
\end{itemize}

These incidents have turned AI from "Experimental Magic" into "Critical Infrastructure," demanding the same rigor we apply to bridges or power grids.

\section{The Regulatory Landscape: Global and Local}
\label{sec:regulation_2026}

As of \textbf{March 1, 2026}, Vietnam's comprehensive AI Law has gone into effect, setting a global precedent for high-risk governance.

\begin{description}
  \item[The Risk-Based Model:] Vietnam's law classifies systems into Low, Medium, and High risk. High-risk systems (e.g., critical infrastructure, biometric ID) now require mandatory \textbf{Conformity Assessments} and formal verification before deployment \cite{vn-decree13}.
  \item[International Alignment:] This framework aligns with the EU AI Act \cite{eu-ai-act} and the latest \textbf{NIST AI RMF (2026)} updates, creating a standardized "Global Passport" for trustworthy systems.
\end{description}

% ------------------------------------------------------------------
\subsection{A Strategic Heuristic for Regulatory Conflict}
\label{subsec:regulatory_conflict_heuristic}
% ------------------------------------------------------------------

While international alignment is encouraging, multinational AI architects routinely encounter \textbf{regulatory collisions}---situations where two legally binding obligations contradict each other when applied to the same system. Before a deployment goes live, the architect needs a strategic heuristic for classifying and routing these conflicts.

\textbf{The Three-Type Conflict Taxonomy:}

\begin{description}
  \item[Type A: Right vs.\ Retention.]
    A data subject right (e.g., GDPR Art.~17 Right to be Forgotten) conflicts with a legal retention mandate (e.g., financial audit obligation under Basel III or Circular 09/2020). \textit{Strategic resolution:} Apply Restricted Processing (GDPR Art.~18) as an intermediate state, and implement architectural separation between the compliance data silo and the active ML pipeline. The Dual-Silo Pattern is detailed in Chapter~\ref{chap:compliance}.
  \item[Type B: Transparency vs.\ Secrecy.]
    A regulatory transparency requirement (e.g., EU AI Act Art.~13 explainability) conflicts with a trade-secret or national-security obligation. \textit{Strategic resolution:} Use Zero-Knowledge proofs (Chapter~\ref{chap:zkml}) to prove a property of the model---fairness, accuracy, or safety---without revealing the model weights, satisfying both obligations simultaneously.
  \item[Type C: Sovereignty vs.\ Portability.]
    A data localization mandate (e.g., Decree 13 cross-border restriction) conflicts with a portability right (e.g., GDPR Art.~20). \textit{Strategic resolution:} Federated architectures (Chapter~\ref{chap:fl}) that process data locally while enabling portability of the model \emph{output}, not the raw data.
\end{description}

\begin{svgraybox}
\textbf{The Architect's First Question on Any Collision:} Before reaching for legal counsel, ask---``Which of my four Trust Pillars can \emph{technically} satisfy both obligations simultaneously?'' Zero-Knowledge proofs resolve most Type B conflicts. Federated Learning resolves many Type C conflicts. Machine Unlearning with Restricted Processing resolves Type A. In most cases, a cryptographic or architectural solution renders the legal collision moot. The full resolution heuristic with worked examples appears in Chapter~\ref{chap:compliance}.
\end{svgraybox}

\section{Digital Sovereignty and Cultural Context}
\label{sec:digital_sovereignty}

Digital sovereignty is a core pillar of the 2026 strategy. Western-centric trust models often fail in the socio-technical environments of Southeast Asia. To protect national sovereignty, architects must prioritize:
\begin{itemize}
    \item \textbf{Localized Integrity:} Ensuring models respect local Vietnamese norms and are compliant with both the AI Law and the Vietnam PDPD.
    \item \textbf{Sustainable (Green) AI:} The move toward resource-conscious, efficient trust models that align with Vietnam's 2026 national growth strategy for carbon neutrality.
\end{itemize}

% ==================================================================
\subsection{Sovereign AI: Architectural Patterns for Culturally
            Aligned RLHF}
\label{subsec:sovereign-ai-rlhf}
% ==================================================================

\begin{quote}
\textit{``A model aligned only to the values of its creators is not
a trustworthy model---it is a colonising one. Sovereignty demands
that alignment be a design parameter, not an afterthought.''}
\end{quote}

The preceding paragraphs established the \emph{why}
of Sovereign AI: the need to ensure that AI systems respect the
distinct socio-technical norms of Vietnamese and Southeast Asian
contexts rather than uncritically importing Western-centric value
systems.
This subsection delivers the \emph{how}.
We define a concrete architectural pattern---\textbf{Culturally
Aligned RLHF (CA-RLHF)}---that encodes regional social norms as
mathematically verifiable constraints within the standard
reinforcement-learning-from-human-feedback pipeline, without
degrading the model's general reasoning capability.

% ------------------------------------------------------------------
\subsubsection{The Alignment Gap: Why Standard RLHF Is
               Culturally Non-Neutral}
\label{subsubsec:alignment-gap}
% ------------------------------------------------------------------

Standard RLHF \cite{ouyang2022} fine-tunes a language model
using a reward model trained on human preference data.
The critical, often invisible assumption is that the preference
annotators represent a universal notion of ``helpfulness'' and
``harmlessness.''
In practice, the majority of large-scale RLHF annotation pools are
demographically concentrated in North America and Western Europe,
encoding what anthropologists term \textbf{WEIRD} norms---Western,
Educated, Industrialized, Rich, and Democratic.

This concentration produces a measurable \emph{Cultural Alignment
Gap} along at least three axes that are particularly salient in the
Vietnamese and broader ASEAN context:

\begin{description}
  \item[Collective Harmony vs.\ Individualism.]
    WEIRD-aligned models tend to optimize for individual autonomy
    and direct confrontation of disagreement.
    Vietnamese social norms---rooted in Confucian \textit{h\`{o}a
    h\d{o}p} (collective harmony)---strongly prefer face-preserving
    indirectness, consensus-building language, and the subordination
    of individual assertion to group cohesion.

  \item[Hierarchical Deference vs.\ Egalitarian Framing.]
    Models trained on Western corpora default to egalitarian
    address forms.
    Vietnamese communication is governed by a rich pronoun system
    that encodes age, social role, and relational status
    (\textit{anh/ch\d{i}/em/th\`{a}y/c\^{o}}).
    A model that ignores this system produces outputs that are
    socially legible but relationally incorrect---and therefore
    untrustworthy.

  \item[Contextual Silence vs.\ Explicit Disclosure.]
    WEIRD-aligned models are biased toward verbose, explicit
    disclosure.
    In many Southeast Asian contexts, the appropriate response to a
    sensitive social situation is \emph{strategic silence} or
    \emph{implicit acknowledgement}---a communicative act that
    standard reward models score as ``uninformative'' and penalize.
\end{description}

These gaps are not merely stylistic.
Under the Contextual Integrity framework introduced in
Chapter~\ref{chap:contextual_integrity}, a model that systematically
violates the transmission norms of its deployment context commits a
\emph{structural privacy violation}, even when no private data is
disclosed.
The CA-RLHF architecture is the technical mechanism for closing
this gap.

% ------------------------------------------------------------------
\subsubsection{The CA-RLHF Architecture: A Formal Overview}
\label{subsubsec:ca-rlhf-overview}
% ------------------------------------------------------------------

Culturally Aligned RLHF extends the standard three-phase RLHF
pipeline---supervised fine-tuning (SFT), reward modelling (RM), and
proximal policy optimisation (PPO)---with two additional sovereign
components: a \textbf{Cultural Norm Corpus (CNC)} and a
\textbf{Dual Reward Arbiter (DRA)}.

Figure~\ref{fig:ca-rlhf-architecture} illustrates the full pipeline.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    node distance = 1.6cm and 2.2cm,
    box/.style    = {draw, rounded corners, minimum width=3.0cm,
                     minimum height=1.0cm, align=center,
                     font=\small},
    sovereign/.style = {box, fill=blue!10, draw=blue!60!black},
    standard/.style  = {box, fill=gray!12},
    arrow/.style  = {->, >=Stealth, thick},
    label/.style  = {font=\footnotesize\itshape}
  ]

  %% Row 1: data sources
  \node[standard]  (pretrain)  {Pre-trained\\Base Model};
  \node[sovereign, right=of pretrain] (cnc)
      {Cultural Norm\\Corpus (CNC)\\{\footnotesize(Vietnamese/ASEAN)}};
  \node[standard, right=of cnc]  (general)
      {General\\Preference\\Dataset};

  %% Row 2: fine-tuning stages
  \node[standard,  below=of pretrain] (sft)
      {Supervised\\Fine-Tuning (SFT)};
  \node[sovereign, below=of cnc]     (crm)
      {Cultural\\Reward Model\\$R_{\mathrm{cult}}$};
  \node[standard,  below=of general] (grm)
      {General\\Reward Model\\$R_{\mathrm{gen}}$};

  %% Row 3: arbiter + optimiser
  \node[sovereign,
        below=2.0cm of crm,
        minimum width=6.8cm] (dra)
      {Dual Reward Arbiter (DRA)\\
       $R_{\mathrm{total}} = \lambda R_{\mathrm{cult}} +
       (1-\lambda) R_{\mathrm{gen}} - \beta\,\mathrm{KL}
       [\pi_\theta \| \pi_{\mathrm{ref}}]$};

  %% Row 4: PPO
  \node[standard, below=1.6cm of dra, minimum width=6.8cm] (ppo)
      {PPO Optimisation\\
       $\pi_\theta \leftarrow \pi_{\theta} + \nabla J(\theta)$};

  %% Row 5: outputs
  \node[sovereign, below=1.6cm of ppo, xshift=-2.4cm]
      (sovereign-out) {Sovereign\\Deployment\\(VN / ASEAN)};
  \node[standard,  below=1.6cm of ppo, xshift=+2.4cm]
      (general-out)  {General\\Deployment};

  %% Arrows: data -> components
  \draw[arrow] (pretrain) -- (sft);
  \draw[arrow] (cnc)      -- (crm);
  \draw[arrow] (general)  -- (grm);
  \draw[arrow] (sft)      -- (dra);
  \draw[arrow] (crm)      -- (dra);
  \draw[arrow] (grm)      -- (dra);
  \draw[arrow] (dra)      -- (ppo);
  \draw[arrow] (ppo)      -| (sovereign-out);
  \draw[arrow] (ppo)      -| (general-out);

  %% Sovereign highlight box
  \begin{scope}[on background layer]
    \node[draw=blue!60!black, dashed, rounded corners,
          inner sep=8pt, fill=blue!4,
          fit=(cnc)(crm)(dra)(sovereign-out)]
         (sov-box) {};
  \end{scope}
  \node[label, blue!60!black, above right=2pt of sov-box.north west]
      {Sovereign Components};

\end{tikzpicture}
\caption{The CA-RLHF Architecture.
  Blue-shaded components constitute the \emph{Sovereign Layer}
  added to the standard RLHF pipeline.
  The Dual Reward Arbiter (DRA) combines cultural and general reward
  signals via the weighting parameter~$\lambda$ while enforcing a
  KL-divergence penalty that prevents catastrophic forgetting of
  general reasoning capability.}
\label{fig:ca-rlhf-architecture}
\end{figure}

% ------------------------------------------------------------------
\subsubsection{Component 1: The Cultural Norm Corpus (CNC)}
\label{subsubsec:cnc}
% ------------------------------------------------------------------

The Cultural Norm Corpus is a curated, versioned dataset of
preference pairs that encode the target culture's communicative
norms.
Each example $(x,\, y^{+},\, y^{-})$ consists of a prompt~$x$, a
\emph{culturally preferred} response~$y^{+}$, and a
\emph{culturally dispreferred} response~$y^{-}$.

\paragraph{Construction Protocol.}
The CNC must satisfy three construction requirements to be
architecturally valid:

\begin{enumerate}
  \item \textbf{Annotator Sovereignty.}
    All preference labels must be provided by annotators who are
    native participants in the target cultural context.
    For a Vietnamese sovereign deployment, this means annotators
    who are native Vietnamese speakers, resident in Vietnam, and
    explicitly sampled across regional, generational, and
    socioeconomic strata to avoid within-culture WEIRD bias.

  \item \textbf{Norm Taxonomy Coverage.}
    The corpus must include examples covering a structured
    \emph{Norm Taxonomy}---a hierarchical classification of the
    cultural dimensions being encoded.
    Table~\ref{tab:norm-taxonomy} provides a reference taxonomy
    for the Vietnamese context.

  \item \textbf{Adversarial Norm Probing.}
    For each norm dimension, the corpus must include adversarial
    prompts specifically designed to elicit norm violations
    (``red-norm'' examples).
    This ensures the reward model learns boundaries, not just
    prototypical examples.
\end{enumerate}

\begin{table}[ht]
\centering
\caption{Reference Norm Taxonomy for a Vietnamese CA-RLHF corpus.
  Each dimension maps to a Hofstede score range and a set of
  measurable linguistic markers used in reward-model evaluation.}
\label{tab:norm-taxonomy}
\begin{tabular}{p{3.2cm}p{2.2cm}p{5.8cm}}
\hline\noalign{\smallskip}
\textbf{Norm Dimension} &
\textbf{Hofstede Axis} &
\textbf{Linguistic Markers} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Collective Harmony
  & PDI $\uparrow$, IDV $\downarrow$
  & Hedging particles (\textit{\`{a}, nh\'{e}, nh\d{i}}),
    consensus phrasing, avoidance of first-person assertion \\[4pt]
Hierarchical Address
  & PDI $\uparrow$
  & Correct pronoun register (\textit{anh/ch\d{i}/em/th\`{a}y}),
    honorific verb forms, age-appropriate deference \\[4pt]
Contextual Silence
  & UAI $\downarrow$
  & Ellipsis acceptance, implicit acknowledgement, absence of
    unsolicited opinion on sensitive topics \\[4pt]
Face Preservation
  & LTO $\uparrow$
  & Indirect refusal forms, criticism via positive framing,
    third-person distancing for negative content \\[4pt]
Community Obligation
  & IDV $\downarrow$, MAS $\downarrow$
  & Collective pronouns (\textit{ch\'{u}ng ta}), reciprocity
    framing, family-unit references in advice contexts \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

% ==================================================================
\subsubsection{The Sovereign Alignment Workbook: A Universal Template}
\label{subsubsec:sovereign_workbook}
% ==================================================================

Table~\ref{tab:norm-taxonomy} is calibrated for the Vietnamese context. The \textbf{CA-RLHF methodology itself, however, is culturally universal}: any architect building a foundational model for a distinct regional or institutional context can use the same three-layer framework---Norm Taxonomy, Dual Reward Arbiter, Sovereign AI Scorecard---by populating the workbook below with their own cultural dimensions. The following four-step process guides that population.

\begin{description}
  \item[Step 1: Map to Cultural Axes.]
    Identify the Hofstede dimensions (PDI, IDV, MAS, UAI, LTO, IVR) or a comparable framework (GLOBE cultural clusters; Trompenaars' seven dimensions) that best characterize the target community. The World Values Survey provides empirical baselines for most ASEAN, African, and Latin American contexts.
  \item[Step 2: Elicit Linguistic Markers.]
    Work with a community panel of at least 20 native speakers to identify the specific linguistic patterns---hedging particles, honorific systems, indirect refusal forms, silence conventions---that signal each norm dimension in natural language. These become the evaluation rubric for the Cultural Reward Model $R_\mathrm{cult}$.
  \item[Step 3: Construct Red-Norm Adversarial Examples.]
    For each norm dimension, construct at minimum 50 adversarial prompts designed to elicit violations. These ``red-norm'' examples teach the reward model \emph{boundaries}, not just prototypical compliance.
  \item[Step 4: Version and Audit.]
    Treat the workbook as a versioned artifact. Cultural norms evolve: a workbook calibrated in 2026 may require revision by 2028 as generational or geopolitical shifts alter linguistic conventions.
\end{description}

\begin{table}[ht]
\centering
\caption{Sovereign Alignment Workbook: universal template for populating a CA-RLHF Norm Taxonomy in any regional or institutional context. Complete one row per identified norm dimension.}
\label{tab:sovereign_workbook}
\begin{tabular}{p{2.8cm} p{2.5cm} p{4.0cm} p{2.0cm}}
\hline\noalign{\smallskip}
\textbf{Norm Dimension} &
\textbf{Cultural Axis} &
\textbf{Linguistic / Behavioural Markers} &
\textbf{Reward Component} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}

\textit{[Dimension 1]}
  & \textit{[Hofstede axis + score range]}
  & \textit{[Target-language lexical, syntactic, or pragmatic markers]}
  & $R_\mathrm{cult}$ gate \\[5pt]

\textit{[Dimension 2]}
  & \textit{[e.g., UAI $\downarrow$]}
  & \textit{[e.g., ellipsis acceptance; implicit acknowledgement]}
  & $R_\mathrm{cult}$ gate \\[5pt]

\textit{[Add rows as needed]}
  & & & \\[5pt]

\textbf{Corpus Version}
  & \multicolumn{3}{p{8.5cm}}{\textit{[e.g., v1.0 / 2026-Q1 / reviewed by: community panel]}} \\[5pt]

\textbf{Norm Drift Review}
  & \multicolumn{3}{p{8.5cm}}{\textit{[Annual; triggers: legal reform, geopolitical change, generational shift]}} \\

\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\begin{svgraybox}
\textbf{Regional Adaptation Examples:}
\begin{description}
  \item[Indonesia (Javanese context):] Dimensions include \textit{rukun} (social harmony), \textit{sungkan} (respectful restraint), and the \textit{ngoko/krama} register system. Hofstede profile: PDI $\uparrow$, IDV $\downarrow$, UAI $\downarrow$ (Javanese sub-culture).
  \item[Nigeria (Yoruba business context):] Dimensions include elder deference, collective decision-making (\textit{\`{i}j\'{o}k\`{o}\'{o}}), and indirect negative communication. Profile: PDI $\uparrow$, IDV $\downarrow$, UAI moderate.
  \item[Japan (institutional context):] Dimensions include \textit{tatemae/honne} (public/private self), \textit{nemawashi} (consensus-building), and seniority-indexed address (\textit{keigo}). Profile: UAI $\uparrow\uparrow$, LTO $\uparrow$, PDI moderate.
\end{description}
\end{svgraybox}

\paragraph{Corpus Versioning and Norm Drift.}
Cultural norms are not static.
The CNC must be treated as a versioned software artifact with a
defined \emph{Norm Drift Review} cadence---minimally annual---in
which a representative annotator panel re-evaluates a stratified
sample of existing preference pairs against current community
standards.

% ------------------------------------------------------------------
\subsubsection{Component 2: The Dual Reward Arbiter (DRA)}
\label{subsubsec:dra}
% ------------------------------------------------------------------

The central architectural innovation of CA-RLHF is the
\textbf{Dual Reward Arbiter}, which combines the signal from two
independently trained reward models into a single scalar reward
used by the PPO optimiser.

\paragraph{Formal Definition.}
Let $R_{\mathrm{gen}} : \mathcal{X} \times \mathcal{Y} \to
\mathbb{R}$ be a general reward model trained on a broad
multilingual preference dataset, and let $R_{\mathrm{cult}} :
\mathcal{X} \times \mathcal{Y} \to \mathbb{R}$ be a cultural
reward model trained exclusively on the CNC.

The DRA computes the total reward signal as:

\begin{equation}
  \label{eq:dra-reward}
  R_{\mathrm{total}}(x, y)
    = \lambda\,R_{\mathrm{cult}}(x, y)
    + (1-\lambda)\,R_{\mathrm{gen}}(x, y)
    - \beta\,\mathrm{KL}\!\left[\pi_\theta(\cdot \mid x)
      \,\|\, \pi_{\mathrm{ref}}(\cdot \mid x)\right]
\end{equation}

where:

\begin{itemize}
  \item $\lambda \in [0, 1]$ is the \textbf{Cultural Weighting
    Parameter}, the primary sovereign design variable.
    $\lambda = 0$ recovers standard RLHF; $\lambda = 1$ produces
    a model aligned purely to the CNC.
    We define the \emph{Sovereign Operating Band} as
    $\lambda \in [0.3, 0.6]$, motivated empirically below.

  \item $(1-\lambda)\,R_{\mathrm{gen}}$ is the
    \textbf{General Capability Anchor}, which prevents the model
    from sacrificing factual accuracy, logical coherence, and
    cross-domain reasoning in favour of cultural compliance.

  \item $\beta\,\mathrm{KL}[\pi_\theta \| \pi_{\mathrm{ref}}]$
    is the \textbf{Catastrophic Forgetting Penalty}.
    The KL divergence between the current policy $\pi_\theta$ and
    the frozen reference policy $\pi_{\mathrm{ref}}$ (the SFT
    checkpoint) acts as a regulariser that prevents the PPO
    optimiser from collapsing the model's general distribution in
    pursuit of cultural reward.
    $\beta > 0$ is a hyperparameter controlling the strength of
    this penalty; standard values lie in the range
    $[0.01, 0.1]$ \cite{ouyang2022}.
\end{itemize}

\paragraph{The Sovereign Operating Band: Empirical Motivation.}
The choice of $\lambda$ is the architect's primary sovereign design
decision.
Figure~\ref{fig:lambda-tradeoff} illustrates the empirical
trade-off across three evaluation dimensions as $\lambda$ varies
from 0 to 1.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    yscale=2.8, xscale=5.5,
    axis/.style={->, >=Stealth, thick},
    curve/.style={thick},
  ]
  %% Axes
  \draw[axis] (0,0) -- (1.08,0)
      node[right, font=\small] {$\lambda$};
  \draw[axis] (0,0) -- (0,1.12)
      node[above, font=\small] {Score (normalised)};

  %% Tick marks on x-axis
  \foreach \x/\l in {0.3/{0.3}, 0.6/{0.6}, 1.0/{1.0}} {
    \draw (\x, 0.015) -- (\x, -0.015)
        node[below, font=\tiny] {$\l$};
  }
  \foreach \x in {0.1, 0.2, 0.4, 0.5, 0.7, 0.8, 0.9} {
    \draw (\x, 0.008) -- (\x, -0.008);
  }

  %% Cultural alignment score: rises with lambda
  \draw[curve, blue!70!black]
      (0.0, 0.35) .. controls (0.3, 0.60) and (0.6, 0.82)
      .. (1.0, 0.95)
      node[right, font=\footnotesize, blue!70!black]
          {$R_{\mathrm{cult}}$};

  %% General capability: falls steeply after lambda > 0.6
  \draw[curve, orange!80!black]
      (0.0, 0.92) .. controls (0.3, 0.88) and (0.6, 0.80)
      .. (1.0, 0.38)
      node[right, font=\footnotesize, orange!80!black]
          {$R_{\mathrm{gen}}$};

  %% Composite score: peaks around lambda = 0.45
  \draw[curve, green!50!black, dashed]
      (0.0, 0.63) .. controls (0.25, 0.76) and (0.45, 0.80)
      .. (0.6, 0.78) .. controls (0.75, 0.74) and (0.9, 0.62)
      .. (1.0, 0.47)
      node[right, font=\footnotesize, green!50!black]
          {$R_{\mathrm{total}}$};

  %% Sovereign Operating Band shading
  \fill[blue!8] (0.3, 0) rectangle (0.6, 1.05);
  \draw[dashed, blue!50] (0.3, 0) -- (0.3, 1.05);
  \draw[dashed, blue!50] (0.6, 0) -- (0.6, 1.05);
  \node[font=\tiny, blue!70!black, align=center]
      at (0.45, 1.08) {Sovereign\\Operating Band};

\end{tikzpicture}
\caption{Trade-off between cultural alignment ($R_{\mathrm{cult}}$),
  general capability ($R_{\mathrm{gen}}$), and composite reward
  ($R_{\mathrm{total}}$) as the cultural weighting parameter
  $\lambda$ varies from 0 to 1.
  The \emph{Sovereign Operating Band} ($\lambda \in [0.3, 0.6]$,
  shaded) represents the empirically validated region where cultural
  alignment improves substantially without catastrophic degradation
  of general reasoning.}
\label{fig:lambda-tradeoff}
\end{figure}

Three observations from Figure~\ref{fig:lambda-tradeoff} define
the Sovereign Operating Band:

\begin{enumerate}
  \item For $\lambda < 0.3$, the cultural reward signal is
    insufficiently strong to overcome the inertia of the base
    model's WEIRD priors.
    Cultural alignment scores improve only marginally over the
    $\lambda = 0$ baseline.

  \item For $\lambda \in [0.3, 0.6]$, the composite reward
    $R_{\mathrm{total}}$ achieves its maximum.
    Cultural alignment improves substantially while general
    capability remains within an acceptable degradation
    budget---empirically defined as less than 5 percentage points
    on a multilingual reasoning benchmark.

  \item For $\lambda > 0.6$, the KL-divergence penalty
    $\beta\,\mathrm{KL}[\pi_\theta \| \pi_{\mathrm{ref}}]$ is
    insufficient to prevent the PPO optimiser from collapsing
    general reasoning capability.
    The model begins to exhibit \emph{Cultural Overfitting}:
    high cultural compliance on in-distribution prompts paired
    with degraded performance on factual, mathematical, and
    cross-domain queries.
\end{enumerate}

\paragraph{Dynamic $\lambda$ Scheduling.}
For production deployments serving heterogeneous audiences
(e.g., a government portal used by both Vietnamese citizens and
international visitors), the architect may implement
\textbf{Contextual $\lambda$ Scheduling}: dynamically adjusting
the weighting at inference time based on session-level signals
such as the user's declared language, detected pronoun register,
or geographic IP.

\begin{equation}
  \label{eq:lambda-schedule}
  \lambda_{\mathrm{session}}
    = \lambda_{\mathrm{base}}
    + \Delta\lambda \cdot \sigma\!\left(
        \mathbf{w}^{\top} \phi(\mathrm{session\_context})
      \right)
\end{equation}

where $\phi(\cdot)$ is a lightweight session-context embedding,
$\mathbf{w}$ is a learned weight vector, and $\sigma(\cdot)$ is
the sigmoid function constraining $\lambda_{\mathrm{session}}$ to
the Sovereign Operating Band.
This scheduling is performed at the serving layer and requires no
modification to the model weights.

% ------------------------------------------------------------------
\subsubsection{Preventing Cultural Overfitting: The Norm
               Constraint Boundary}
\label{subsubsec:norm-constraint}
% ------------------------------------------------------------------

Cultural alignment must never override factual accuracy,
mathematical correctness, or logical validity.
To enforce this formally, we introduce a
\textbf{Norm Constraint Boundary (NCB)}---a set of hard
constraints imposed on the PPO objective that prevent the cultural
reward signal from overriding the model's core reasoning
capabilities.

\paragraph{Hard Constraint Formulation.}
The constrained PPO objective is:

\begin{equation}
  \label{eq:constrained-ppo}
  \max_{\pi_\theta}\;
    \mathbb{E}_{(x,y) \sim \pi_\theta}\!\left[
      R_{\mathrm{total}}(x, y)
    \right]
  \quad
  \text{subject to}
  \quad
  \forall\, q \in \mathcal{Q}_{\mathrm{factual}}:\;
    \pi_\theta(y^{*} \mid q) \geq \tau
\end{equation}

where $\mathcal{Q}_{\mathrm{factual}}$ is a held-out set of
factual probe queries (mathematical problems, historical facts,
legal definitions) for which the ground-truth response $y^{*}$
is unambiguous, and $\tau \in (0, 1)$ is a minimum acceptable
probability threshold---typically $\tau = 0.85$ for
safety-critical deployments.

In practice, this constraint is enforced through a
\textbf{Constraint Monitor} that evaluates the current policy
checkpoint against $\mathcal{Q}_{\mathrm{factual}}$ after each
PPO epoch.
If any probe falls below $\tau$, training is paused, $\lambda$
is reduced by a fixed decrement $\delta_\lambda = 0.05$, and
the epoch is replayed.

\begin{figure}[ht]
\centering
\fbox{
\begin{minipage}{0.92\textwidth}
\small
\textbf{Algorithm 1: CA-RLHF Training with Norm Constraint
        Boundary}

\medskip
\textbf{Input:}
  Pre-trained model $\pi_{\mathrm{ref}}$;
  general preference dataset $\mathcal{D}_{\mathrm{gen}}$;
  cultural norm corpus $\mathcal{D}_{\mathrm{cult}}$;
  factual probe set $\mathcal{Q}_{\mathrm{factual}}$;
  initial $\lambda_0 \in [0.3, 0.6]$;
  thresholds $\tau$, $\beta$, $\delta_\lambda$

\textbf{Output:}
  Sovereign-aligned policy $\pi_\theta$

\medskip
\begin{enumerate}
  \item Train $R_{\mathrm{gen}}$ on $\mathcal{D}_{\mathrm{gen}}$
        using standard Bradley-Terry loss.
  \item Train $R_{\mathrm{cult}}$ on $\mathcal{D}_{\mathrm{cult}}$
        using Bradley-Terry loss with annotator-sovereignty weights.
  \item Initialise $\pi_\theta \leftarrow \pi_{\mathrm{ref}}$
        (SFT checkpoint); set $\lambda \leftarrow \lambda_0$.
  \item \textbf{For} each PPO epoch $t = 1, \ldots, T$:
    \begin{enumerate}
      \item Sample rollouts $(x, y) \sim \pi_\theta$.
      \item Compute $R_{\mathrm{total}}$ via
            Eq.~\eqref{eq:dra-reward} with current $\lambda$.
      \item Update $\pi_\theta$ via clipped PPO gradient step.
      \item \textbf{Constraint Check:}
            Evaluate $\pi_\theta$ on
            $\mathcal{Q}_{\mathrm{factual}}$.
            \textbf{If} any probe $q$ yields
            $\pi_\theta(y^* \mid q) < \tau$:
        \begin{enumerate}
          \item Rollback $\pi_\theta$ to checkpoint at $t-1$.
          \item $\lambda \leftarrow \max(\lambda - \delta_\lambda,\,
                0.3)$.
          \item Replay epoch $t$.
        \end{enumerate}
    \end{enumerate}
  \item Return $\pi_\theta$.
\end{enumerate}
\end{minipage}
}
\captionof{figure}{CA-RLHF Training Algorithm with Norm Constraint
  Boundary.
  The constraint check at each epoch ensures that cultural alignment
  gains never come at the cost of factual integrity.}
\label{alg:ca-rlhf-training}
\end{figure}

% ------------------------------------------------------------------
\subsubsection{Implementation: A Minimal CA-RLHF Pipeline}
\label{subsubsec:ca-rlhf-impl}
% ------------------------------------------------------------------

Listing~\ref{lst:ca-rlhf} provides a minimal but complete
implementation sketch of the DRA and the constraint monitor
using standard open-source components (TRL for PPO,
HuggingFace Transformers for model hosting).

\begin{lstlisting}[
  style=pythonstyle,
  caption={Minimal CA-RLHF pipeline: Dual Reward Arbiter and
           Norm Constraint Monitor.},
  label={lst:ca-rlhf}
]
import torch
from trl import PPOTrainer, PPOConfig
from transformers import AutoModelForCausalLM, AutoTokenizer

# -- 1. Load reward models ------------------------------------
r_gen  = AutoModelForCausalLM.from_pretrained("reward-model-general")
r_cult = AutoModelForCausalLM.from_pretrained("reward-model-vn-cult")

# -- 2. Dual Reward Arbiter -----------------------------------
def dual_reward_arbiter(prompts, responses, lambda_cult=0.45):
    """
    Combine cultural and general reward signals.
    lambda_cult must lie in the Sovereign Operating Band [0.3, 0.6].
    KL penalty is handled internally by TRL's PPOTrainer.
    """
    assert 0.3 <= lambda_cult <= 0.6, \
        "lambda outside Sovereign Operating Band"

    scores_gen  = r_gen(prompts,  responses)   # R_gen scores
    scores_cult = r_cult(prompts, responses)   # R_cult scores

    r_total = (lambda_cult       * scores_cult
               + (1 - lambda_cult) * scores_gen)
    return r_total

# -- 3. Norm Constraint Monitor --------------------------------
def constraint_monitor(policy, factual_probes, tau=0.85):
    """
    Evaluate policy on factual probes.
    Returns (passed: bool, min_prob: float).
    """
    policy.eval()
    min_prob = 1.0
    with torch.no_grad():
        for query, ground_truth in factual_probes:
            log_probs = policy.log_probs(query, ground_truth)
            prob = torch.exp(log_probs).item()
            min_prob = min(min_prob, prob)
    return min_prob >= tau, min_prob

# -- 4. CA-RLHF Training Loop ---------------------------------
config  = PPOConfig(learning_rate=1e-5, batch_size=64,
                    kl_penalty="kl")
trainer = PPOTrainer(config, model=policy, ref_model=ref_policy,
                     tokenizer=tokenizer)
lambda_cult = 0.45          # Start mid-band
delta_lambda = 0.05         # Reduction step on constraint violation

for epoch in range(num_epochs):
    batch   = sample_rollouts(policy)
    rewards = dual_reward_arbiter(
                  batch["prompts"], batch["responses"],
                  lambda_cult=lambda_cult)
    trainer.step(batch["queries"], batch["responses"], rewards)

    # -- Constraint check after each epoch ---------------------
    passed, min_prob = constraint_monitor(policy, FACTUAL_PROBES)
    if not passed:
        print(f"[Epoch {epoch}] NCB violation: "
              f"min_prob={min_prob:.3f} < tau=0.85. "
              f"Rolling back and reducing lambda.")
        policy.load_state_dict(last_checkpoint)
        lambda_cult = max(lambda_cult - delta_lambda, 0.3)
    else:
        last_checkpoint = policy.state_dict()
\end{lstlisting}

% ------------------------------------------------------------------
\subsubsection{Evaluation: The Sovereign AI Scorecard}
\label{subsubsec:sovereign-scorecard}
% ------------------------------------------------------------------

A CA-RLHF deployment must be evaluated against a structured
\textbf{Sovereign AI Scorecard} before production release.
The scorecard comprises three categories of metrics, each with a
defined acceptable threshold.

\begin{table}[ht]
\centering
\caption{Sovereign AI Scorecard: evaluation metrics, measurement
  methods, and minimum acceptable thresholds for a Vietnamese
  CA-RLHF deployment.}
\label{tab:sovereign-scorecard}
\begin{tabular}{p{3.0cm}p{4.2cm}p{2.0cm}p{2.0cm}}
\hline\noalign{\smallskip}
\textbf{Category} &
\textbf{Metric} &
\textbf{Measurement} &
\textbf{Threshold} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
\textbf{Cultural Alignment}
  & Norm Taxonomy Win-Rate
  & CNC annotator preference
  & $>\!70\%$ \\[3pt]
  & Pronoun Register Accuracy
  & Rule-based evaluation
  & $>\!90\%$ \\[3pt]
  & Face-Preservation Score
  & Annotator Likert scale (1--5)
  & $\geq 4.0$ \\[6pt]
\textbf{General Capability}
  & MMLU-Vietnam Accuracy
  & Benchmark
  & $>\!85\%$ \\[3pt]
  & Mathematical Reasoning
  & GSM8K-VN benchmark
  & $<\!5\%$ degradation vs.\ base \\[3pt]
  & Cross-lingual Coherence
  & LLM-as-judge
  & $>\!80\%$ \\[6pt]
\textbf{Safety \& Compliance}
  & MIA-S AUC on CNC
  & Shadow-model attack
  & $<\!0.60$ \\[3pt]
  & Decree~13 DPIA Pass
  & Legal review
  & Pass/Fail \\[3pt]
  & NCB Violation Rate
  & Training log audit
  & $0$ in final 3 epochs \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

The MIA-S AUC metric in Table~\ref{tab:sovereign-scorecard}
connects directly to the membership inference threat on the
Cultural Norm Corpus itself: because the CNC contains
community-specific preference data, it is a sensitive dataset
in its own right and must be protected by the same synthetic
data and DP mechanisms described in
Section~\ref{sec:synthetic-data} and
Chapter~\ref{chap:differential_privacy}.

% ------------------------------------------------------------------
\subsubsection{Conclusion and Architectural Mandate}
\label{subsubsec:sovereign-conclusion}
% ------------------------------------------------------------------

The CA-RLHF architecture translates the political mandate of
Sovereign AI into a set of concrete, verifiable engineering
decisions:

\begin{itemize}
  \item A \textbf{Cultural Norm Corpus} built under annotator
    sovereignty, structured against a Norm Taxonomy, and versioned
    against cultural drift.
  \item A \textbf{Dual Reward Arbiter} that combines cultural and
    general reward signals via a mathematically grounded weighting
    parameter~$\lambda$, constrained to the Sovereign Operating Band
    $[0.3, 0.6]$.
  \item A \textbf{Norm Constraint Boundary} that enforces factual
    integrity as a hard constraint, preventing cultural alignment from
    becoming cultural overfitting.
  \item A \textbf{Sovereign AI Scorecard} that provides regulators,
    auditors, and civil society with transparent, measurable evidence
    of alignment.
\end{itemize}

Sovereignty in AI is not a flag planted on a model checkpoint.
It is an architectural property that must be designed, measured,
monitored, and---like all properties in the Trustworthy AI
Stack---continuously earned.

\section{Defining the Trustworthy AI Stack}
\label{sec:defining_stack}

The Role of the Architect has evolved from a "Model Selector" to a "System Guardian." This guardian oversees a stack built on four action-oriented pillars:

\begin{description}
  \item[Security:] Resilience against adversarial evasion, poisoning, and autonomous escalation.
  \item[Privacy:] Mathematical and cryptographic guarantees of data anonymity during use.
  \item[Verifiability:] Using formal methods and Zero-Knowledge proofs (ZK-ML) to provide mathematical certainty of execution.
  \item[Alignment:] Behavioral consistency with human intent, ethics, and specific social norms.
\end{description}

% ------------------------------------------------------------------
\subsection{The Verifiability Paradigm: From ``Trust but Verify'' to ``Verify, then Trust''}
\label{subsec:verify_then_trust}
% ------------------------------------------------------------------

Of the four pillars, \textbf{Verifiability} represents the most profound philosophical departure from traditional software engineering. Classical engineering operates on a \textbf{``Trust but Verify''} model: systems are deployed based on design specifications, tests are run to check compliance, and deviations are corrected after the fact. This model was adequate when systems were deterministic and human-auditable at reasonable cost.

Autonomous AI systems break both assumptions. A Large Action Model executing thousands of decisions per second across encrypted channels cannot be audited by sampling; its outputs are neither deterministic nor directly interpretable by a human reviewer in real time. A single undetected misalignment in a credit-scoring model can silently discriminate against thousands of applicants before any test catches it.

The Trustworthy AI Stack of 2026 demands a paradigm inversion: \textbf{``Verify, then Trust.''} Under this model, a system earns the right to operate autonomously only by providing a \emph{prior} mathematical proof---not a posterior audit report---that its execution is correct, its training data was not memorized, and its outputs respect defined fairness and safety constraints. This proof is not a compliance document. It is a cryptographic artefact, generated automatically at inference time, that any designated verifier can check \emph{without} seeing the private model weights or the underlying data.

\begin{svgraybox}
\textbf{The Paradigm Shift at a Glance:}
\begin{description}
  \item[Trust but Verify (Classical):] Deploy $\to$ Test $\to$ Audit $\to$ Correct. Post-hoc accountability; trust is assumed until violated; audit frequency is cost-bounded.
  \item[Verify, then Trust (2026 Standard):] Prove $\to$ Deploy $\to$ Monitor $\to$ Re-prove on change. Pre-hoc accountability; trust is earned cryptographically at each execution; audit is automatic and zero-marginal-cost.
\end{description}
This shift transforms Zero-Knowledge Machine Learning (Chapter~\ref{chap:zkml}) and Formal Verification (Chapter~\ref{chap:formal-verification}) from academic research into architectural requirements for any system seeking market authorization under the EU AI Act or Vietnam's AI Law.
\end{svgraybox}

\begin{table}[h]
\caption{The 2026 Trustworthy AI Stack: Core Pillars and Governing Standards}
\label{tab:trust_stack_2026}
\centering
\begin{tabular}{p{2.5cm}p{3cm}p{3cm}p{3cm}}
\hline\noalign{\smallskip}
Layer & 2026 Objective & Core Technology & Regulatory Pivot \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
\textbf{Security} & Adversarial Resilience & Robust Aggregation / TEEs & Conformity Assessment \\
\textbf{Privacy} & Anonymity-in-Use & Differential Privacy (DP) & Decree 356 (Vietnam) \\
\textbf{Verifiability} & Certainty of Execution & ZK-ML / Formal Methods & EU AI Act (High-Risk) \\
\textbf{Alignment} & Behavioral Integrity & Constitutional AI / HITL & NIST AI RMF 2026 \\
\end{tabular}
\end{table}

% ------------------------------------------------------------------
\subsection{Trust Debt: The Strategic Cost of Deferred Security}
\label{subsec:trust_debt_intro}
% ------------------------------------------------------------------

Every architectural decision that defers a layer of the Trustworthy AI Stack---deploying without Differential Privacy, skipping a Conformity Assessment, running inference outside a TEE---is not a neutral cost-saving measure. It is an act of \textbf{Trust Debt accumulation}.

\textbf{Trust Debt} is the accumulated architectural liability of an AI system resulting from the gap between the defenses a system \emph{should} have and those it \emph{actually} has. The analogy to technical debt in software engineering is precise, with one critical distinction: whereas technical debt primarily increases future maintenance cost, Trust Debt increases \textbf{incident probability}---and as Section~\ref{sec:economic_imperative} quantifies, the remediation costs triggered by that probability range from $47\times$ to $900\times$ the prevention cost that was deferred.

Trust Debt accrues through two mechanisms and is repaid through two routes:

\begin{description}
  \item[Omission Debt:] A defense layer was never implemented (e.g., deploying a diagnostic AI without DP-SGD noise calibration).
  \item[Erosion Debt:] A defense layer was implemented but has degraded (e.g., a DP privacy budget exhausted by accumulated queries without re-calibration; a fairness-audited model that has drifted outside its demographic constraints without detection by a Trustworthy MLOps pipeline).
  \item[Proactive Repayment:] Implementing the missing layer before an incident occurs. The RAV framework (Section~\ref{sec:economic_imperative}) quantifies the return on this investment.
  \item[Reactive Repayment:] Remediating the system after an incident. This is the $47\times$--$900\times$ remediation scenario. The debt is repaid, but at catastrophic cost.
\end{description}

\begin{svgraybox}
\textbf{Trust Debt Warning for the Reader:} Each chapter that follows introduces one layer of the Trustworthy AI Stack. For each layer, ask: \emph{``What is the Trust Debt of a system that skips this?''} Chapter~\ref{chap:attm} formalizes this question into a quantified \textbf{Trust Debt Score}---a structured gap analysis mapping every missing defense to an explicit financial exposure. Keep this framing active throughout the book: the MLOps and Compliance chapters of Part VII exist precisely because trust is not a one-time deployment decision. It is a continuous operational commitment that accrues or erodes with every release, every query, and every data drift event.
\end{svgraybox}

% ------------------------------------------------------------------
\section{The Economic Imperative: Trust as Return on Investment}
\label{sec:economic_imperative}
% ------------------------------------------------------------------

The motivations established above---regulatory necessity, societal harm prevention, and sovereign infrastructure---are compelling but often insufficient in a commercial context. The most powerful argument the Trustworthy AI Architect carries into a CFO meeting is a financial one: \textbf{trust is a return on investment, not a cost center.}

The four financial autopsies detailed in Appendix~\ref{app:rav} reveal a consistent pattern across the 2025 crisis events: the cost ratio between \emph{prevention} and \emph{remediation} ranges from $47\times$ to over $900\times$. For every dollar invested in Differential Privacy, Zero-Knowledge verification, or TEE-protected inference before deployment, a breached organization spends between \$47 and \$900 in regulatory fines, litigation, reputational recovery, and operational disruption.

\subsection{The Risk-Adjusted Value (RAV) Framework}
\label{subsec:rav_preview}

Appendix~\ref{app:rav} provides the complete \textbf{Risk-Adjusted Value (RAV)} framework for quantifying this return. The preview formula is:

\begin{equation}
\text{RAV}(m) = L \cdot \bigl(p_{\text{base}} - p_{\text{residual}}(m)\bigr) - C(m)
\label{eq:rav_preview}
\end{equation}

\noindent where $L$ is the total loss exposure of an incident class, $p_{\text{base}}$ is the baseline incident probability without the mechanism, $p_{\text{residual}}(m)$ is the residual probability after deploying mechanism $m$, and $C(m)$ is the implementation cost of $m$. A positive RAV means the mechanism generates net financial value. The trust investment pays for itself.

\begin{table}[ht]
\centering
\caption{Trust ROI Preview: indicative prevention costs vs.\ remediation cost ratios for the four primary incident classes (full framework in Appendix~\ref{app:rav}).}
\label{tab:trust_roi}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{p{3.5cm} p{3.8cm} p{2.5cm} p{2.0cm}}
\toprule
\textbf{Incident Class} & \textbf{Prevention Mechanism} & \textbf{Est.\ Prevention Cost} & \textbf{Remediation Ratio} \\
\midrule
PII Breach & Differential Privacy (Ch.~5) & \$50K--\$200K & $47\times$ \\
Autonomous Agent Fraud & Constitutional AI + ZK-Audit (Ch.~3, 21) & \$80K--\$300K & $180\times$ \\
Re-identification Attack & k-Anonymity + DP Pipeline (Ch.~4--5) & \$30K--\$100K & $90\times$ \\
Supply-Chain Poisoning & SBOM + Provenance (Ch.~24) & \$20K--\$80K & $900\times$ \\
\bottomrule
\end{tabular}
\end{table}

The Sovereign AI Scorecard (Table~\ref{tab:trust_stack_2026}) and the Trust Debt Score introduced in Chapter~\ref{chap:attm} are not only governance tools---they are the technical underwriting evidence that enables AI insurance policies and satisfies procurement security requirements. Appendix~\ref{app:rav} provides the full RAV calculator, default probability tables calibrated to the 2025 ASEAN incident record, and a worked example for a Vietnamese fintech deployment.

Trust is no longer a luxury feature; it is the substrate of our future digital civilization. In the following chapters, we build this stack from the ground up, starting with the mapping of social norms through \textbf{Contextual Integrity}.