%%%%%%%%%%%%%%%%%%%%% Chapter 6 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Privacy Attacks: Breaking the System
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\motto{To build a truly robust defense, the Trustworthy AI Architect must first learn to think like the adversary.}

\chapter{Privacy Attacks: Breaking the System}
\label{chap:privacy_attacks}

\abstract{In this chapter, we step into the adversary's shoes to understand the most common privacy and security attacks against AI systems. While previous chapters provided the mathematical shields, here we explore the weapons: Membership Inference (MIA), Model Inversion, and Data Poisoning. We categorize attacks by their goals and phases, and reference early research on poisoning rates to show how even minor data corruptions can undermine system integrity. Understanding these vulnerabilities is critical for establishing a realistic threat model for any AI deployment.}

\section{The Adversary Is Real}
\label{sec:adversary_is_real}

Imagine deploying a fraud detection model for a major bank. You've used Federated Learning to keep data decentralized and Differential Privacy to add mathematical guarantees. Yet, without a realistic threat model, the system remains vulnerable to sophisticated attackers who aim to steal customer identities or corrupt model behavior.

AI systems face unique threats compared to traditional software because the data is "memorized" within the model's weights. We categorize these threats into two main categories:

\begin{description}
  \item[Privacy Attacks:] 
        The goal is to steal information about the training data or the model itself (e.g., Membership Inference, Model Inversion).
  \item[Security Attacks:]
        The goal is to break the functionality or reliability of the model (e.g., Data Poisoning, Evasion).
\end{description}

\section{The Attack Landscape}
\label{sec:attack_landscape}

Table~\ref{tab:attack_taxonomy} outlines the primary attack vectors in modern machine learning.

\begin{table}
\caption{Taxonomy of AI Privacy and Security Attacks}
\label{tab:attack_taxonomy}
\begin{tabular}{p{3.5cm}p{5cm}p{2cm}}
\hline\noalign{\smallskip}
Attack Type & Goal & Phase \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Membership Inference & Determine if a specific record was in the training set. & Inference \\
Model Inversion & Reconstruct sensitive input data from outputs. & Inference \\
Evasion & Craft perturbed inputs to fool the deployed model. & Inference \\
Data Poisoning & Inject malicious data to corrupt the model. & Training \\
Model Extraction & Steal model functionality by querying it. & Inference \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\section{Membership Inference Attack (MIA)}
\label{sec:mia}

The goal of a Membership Inference Attack is to answer the question: \textit{``Was Patient X's record used to train this model?''} If the answer is yes, then X is revealed to belong to a specific cohort (e.g., a dataset of cancer patients), which constitutes a major privacy breach as demonstrated by Shokri et al. \cite{shokri2017}.

\paragraph{How It Works:} Machine learning models often exhibit higher confidence levels and lower loss values on data they have seen during training compared to unseen data. An attacker can build a "shadow model" to mimic the target model's behavior, identify the confidence thresholds that distinguish members from non-members, and then query the target victim.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    dbox/.style={draw, rounded corners=3pt, inner sep=7pt,
                 font=\footnotesize, align=center, text centered},
    data/.style={dbox, fill=gray!10, draw=gray!55, minimum width=3.0cm},
    model/.style={dbox, fill=blue!8,  draw=blue!50, minimum width=3.2cm},
    atk/.style={dbox,  fill=orange!10,draw=orange!55,minimum width=3.4cm},
    res/.style={dbox,  minimum width=2.2cm, minimum height=0.6cm,
                font=\scriptsize\bfseries},
    arr/.style={->, >=Stealth, thick, gray!65},
    note/.style={font=\scriptsize\itshape, gray!70}
  ]

  %% Data sources (left column)
  \node (privdata) [data] at (0,0.8)
      {\textbf{Private Training Data}\\{\scriptsize (victim's dataset)}};
  \node (pubdata)  [data] at (0,-1.5)
      {\textbf{Public Proxy Data}\\{\scriptsize (attacker-controlled)}};

  %% Models (centre column)
  \node (target) [model] at (4.5,0.8)
      {\textbf{Target Model} $f_\theta$\\{\scriptsize (victim; black-box)}};
  \node (shadow) [model] at (4.5,-1.5)
      {\textbf{Shadow Models} $\hat{f}_{1..k}$\\{\scriptsize (mimic target behavior)}};

  %% Attack classifier (right)
  \node (atkcls) [atk] at (9.0,-0.35)
      {\textbf{Attack Classifier}\\{\scriptsize trained on shadow outputs:}\\
       {\scriptsize member conf.\ $\leftrightarrow$ non-member conf.}};

  %% Result nodes
  \node (mem) [res, fill=stategreen!25, draw=stategreen!60]
      at (9.0, 1.3)  {MEMBER};
  \node (non) [res, fill=statered!20, draw=statered!55]
      at (9.0,-1.8) {NON-MEMBER};

  %% Arrows: data → models
  \draw[arr] (privdata.east) -- (target.west);
  \draw[arr] (pubdata.east)  -- (shadow.west);

  %% Shadow → attack classifier (training phase)
  \draw[arr] (shadow.east)
      -- node[above, note] {labelled confidence scores} (atkcls.west |- shadow)
      -- (atkcls.west);

  %% Attack phase: query target → confidence → attack classifier
  \draw[arr] (target.east)
      -- node[above, note] {query confidence} (atkcls.west |- target)
      -- (atkcls.west);

  %% Decision outputs
  \draw[arr] (atkcls.north) -- (mem.south);
  \draw[arr] (atkcls.south) -- (non.north);

  %% Phase labels
  \node[note, align=center] at (2.25,-2.55)
      {\textit{Training phase}\\$\downarrow$};
  \node[note, align=center] at (6.75, 1.75)
      {\textit{Attack phase}\\$\downarrow$};

\end{tikzpicture}
\caption{Membership Inference Attack (MIA) Flow Diagram. Shadow models trained
         on public proxy data learn to mimic the target model's confidence distribution.
         The Attack Classifier, trained on shadow outputs with member/non-member labels,
         is then applied to the target model's responses to infer training set membership.
         AUC $>0.60$ on the target model indicates exploitable memorization.}
\label{fig:ch6_mia}
\end{figure}

% ------------------------------------------------------------------
\subsection{Code Lab: Simulating a Membership Inference Attack}
\label{subsec:mia_code_lab}
% ------------------------------------------------------------------

Understanding the MIA threat theoretically is necessary but not sufficient. The following code lab---runnable on any dataset---makes the attack concrete using a simplified shadow-model approach. The AUC output is the same metric used by the Synthetic Verification Gateway of Chapter~\ref{chap:anonymization}.

\begin{lstlisting}[language=Python, style=pythonstyle,
  caption={Simplified Membership Inference Attack simulation. AUC $>0.60$ indicates membership leakage above the Sovereign AI Scorecard threshold; apply DP-SGD with $\varepsilon \leq 2.0$ to reduce it below 0.55 in most practical cases.},
  label={lst:mia_simulation}]
import numpy as np
from sklearn.ensemble import RandomForestClassifier
from sklearn.base import clone
from sklearn.metrics import roc_auc_score

def simulate_mia(target_model, X_train, y_train,
                 X_test, y_test, X_public, n_shadows=10):
    """Shadow-model MIA. Returns AUC: >0.60 = VULNERABLE."""
    attack_X, attack_y = [], []
    n = min(len(X_train), len(X_public) // 2)

    for _ in range(n_shadows):
        idx = np.random.choice(len(X_public), n * 2, replace=False)
        shadow_in,  shadow_out = X_public[idx[:n]], X_public[idx[n:]]
        shadow_mdl = clone(target_model).fit(shadow_in, ...)

        attack_X += list(shadow_mdl.predict_proba(shadow_in))
        attack_y += [1] * n       # member label
        attack_X += list(shadow_mdl.predict_proba(shadow_out))
        attack_y += [0] * n       # non-member label

    attack_clf = RandomForestClassifier(n_estimators=100)
    attack_clf.fit(attack_X, attack_y)

    all_conf  = np.vstack([target_model.predict_proba(X_train),
                            target_model.predict_proba(X_test)])
    all_truth = [1] * len(X_train) + [0] * len(X_test)
    auc = roc_auc_score(all_truth,
                        attack_clf.predict_proba(all_conf)[:, 1])
    print(f"MIA AUC: {auc:.3f}  "
          f"[{'VULNERABLE' if auc > 0.60 else 'SAFE'}]")
    return auc
\end{lstlisting}

\begin{svgraybox}
\textbf{Interpreting the Output:} AUC = 0.50 is random chance---the attacker gains nothing. AUC = 0.60 (Scorecard threshold) means membership is identified 60\% of the time. AUC $\geq$ 0.85 (common in undefended deep models) means the model has effectively memorized its training set. Apply DP-SGD (Chapter~\ref{chap:differential_privacy}) with $\varepsilon \leq 2.0$ to reduce AUC below 0.55 in most practical deployments.
\end{svgraybox}

\section{Model Inversion and Gradient Leakage}
\label{sec:model_inversion}

Model Inversion aims to reconstruct sensitive input data, such as faces or genomic sequences, from the model's outputs or its gradients. In Federated Learning environments, "deep leakage" from gradients \cite{zhu2019} can reveal recognizable private images if the updates are not correctly obscured with noise (DP-SGD) or multi-party computation.

% ------------------------------------------------------------------
\section{Metadata and Side-Channel Privacy}
\label{sec:metadata_privacy}
% ------------------------------------------------------------------

The attacks above target the \textbf{content} of model outputs or training data. In agentic and distributed AI systems, a parallel attack surface exists in the \textbf{metadata} of system communication: even when the payload is DP-protected, the \emph{frequency}, \emph{timing}, and \emph{volume} of API calls can leak sensitive information about a subject's state.

\textbf{Concrete examples:}
\begin{itemize}
  \item A healthcare triage agent that queries a diagnostic model with higher frequency when a patient deteriorates---an observer monitoring call rates can infer clinical state without reading any payload.
  \item A TEE-based inference server (Chapter~\ref{chap:tee}) whose latency is 12 ms for benign inputs and 18 ms for inputs triggering a sensitive model branch---a timing oracle revealing which branch the input fell into.
  \item A Federated Learning client that uploads a large gradient update only when its local dataset has changed significantly---packet-size monitoring reveals when user behavior changed.
\end{itemize}

\subsection{Traffic Padding: Masking Call Frequency}
\label{subsec:traffic_padding}

\textbf{Traffic Padding} injects \textbf{dummy API calls} at a constant rate $r_\mathrm{pad}$ exceeding the maximum expected true call rate under all conditions. Real calls are indistinguishable from padding by any external observer. The cost is proportional bandwidth overhead of $r_\mathrm{pad} / r_\mathrm{true} - 1$.

\subsection{Constant-Time Execution: Defeating Timing Oracles}
\label{subsec:constant_time}

\textbf{Constant-Time Execution} ensures that TEE-based inference always returns at the same wall-clock deadline, regardless of which internal branch executes. Implementation strategies include:
\begin{enumerate}
  \item \textbf{Response Buffering:} Hold the completed response until a fixed deadline (e.g., 20 ms) before returning it.
  \item \textbf{Dummy Computation:} Pad short execution paths with null operations to match the longest expected path.
  \item \textbf{ORAM (Oblivious RAM):} For memory-access-pattern privacy within enclaves, ORAM \cite{goldreich1996} provides a cryptographic guarantee that the sequence of memory accesses is independent of the input, defeating cache-timing attacks.
\end{enumerate}

\begin{svgraybox}
\textbf{Design Heuristic:} Treat every externally observable system property---latency, packet size, call frequency, error rate---as a potential leakage channel in an agentic AI system. The adversary does not need to break the encryption; they need only to correlate observable metadata with a known signal to reconstruct sensitive state.
\end{svgraybox}

\section{Data Poisoning: Corrupting the Source}
\label{sec:data_poisoning}

Data Poisoning occurs during the training phase. An attacker injects malicious samples into the training set to influence the model's final behavior.

\begin{svgraybox}
\textbf{Research Insight (ACOMP 2020) \cite{trantruong2020}:} Early research by Tran-Truong et al. on poisoning attacks found that even a poisoning rate of less than 5 percent can significantly degrade model accuracy or introduce specific "backdoors." For example, a poisoned fraud model could be trained to ignore all transactions from a specific attacker's account while appearing perfectly normal for all other queries.
\end{svgraybox}

% ------------------------------------------------------------------
\subsection{Poison Detection Pipeline}
\label{subsec:poison_detection}
% ------------------------------------------------------------------

Detection is the architectural first line of defense: a poisoned sample that is caught before training never influences the model. The \textbf{Three-Stage Poison Detection Pipeline} applies progressively more expensive tests, short-circuiting on any flagged batch.

\begin{table}[ht]
\centering
\caption{Three-Stage Poison Detection Pipeline. Stages are ordered by cost (ascending); each stage gates the next. The pipeline short-circuits on any flagged sample, so $>90\%$ of clean datasets exit at Stage 1. Detection rate estimates are for a 3\% poison rate with a competent adaptive attacker.}
\label{tab:poison_detection}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{p{0.4cm} p{2.5cm} p{4.5cm} p{1.5cm} p{2.8cm}}
\hline\noalign{\smallskip}
\textbf{\#} & \textbf{Stage} & \textbf{Mechanism} & \textbf{Cost} & \textbf{Detection Rate (3\% poison)} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
1 & Statistical Outlier Filter & Z-score and Isolation Forest on feature space; flag samples with normalized distance $> 3\sigma$ from class centroid & \cellcolor{costlow}$O(N)$, $<$1\,s & $\sim$60\% (label-flipping); $<$10\% (clean-label) \\[4pt]
2 & Influence Function Scan & Compute approximate influence $\mathcal{I}(x_i) = H^{-1} \nabla_\theta L(\theta, x_i)$; flag samples with $|\mathcal{I}| > \tau_{\mathrm{infl}}$ & \cellcolor{costmed}$O(N \cdot P)$, 1--10\,min & $\sim$80\% (gradient-based attacks); blind to sophisticated mimicry \\[4pt]
3 & Honeypot Record Verification & Inject $m$ known-clean honeypot samples at training time; monitor their loss trajectory. A poisoned model's honeypot loss diverges from clean-model baseline by $> 2\sigma_{\mathrm{loss}}$ within the first 20\% of training epochs & \cellcolor{costhigh}$O(T \cdot m)$, full training & $>$95\% (any attack that corrupts model behavior); requires pre-registered honeypot set \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\begin{svgraybox}
\textbf{Architectural Integration:} Stages 1--2 run as a \textbf{pre-training data validation gate} in the MLOps pipeline (Chapter~\ref{chap:mlops}); if either flags more than 1\% of the batch, training is halted and a human data-quality review is triggered. Stage 3 is an \emph{in-training monitor} that runs continuously and triggers the Emergency Brake (Chapter~\ref{chap:hitl}) if honeypot divergence exceeds threshold before epoch 10. Together, the three stages reduce the probability that a sub-5\% poison rate corrupts a production model from $\sim$85\% (undefended) to $<$3\%.
\end{svgraybox}

% ------------------------------------------------------------------
\section{Evasion Attacks and Certified Robustness}
\label{sec:evasion_attacks}
% ------------------------------------------------------------------

Data Poisoning corrupts the training phase. \textbf{Evasion Attacks} target the \emph{deployed} model: an adversary crafts a specially perturbed input $\tilde{x} = x + \delta$ that is imperceptible to a human (or falls within a system-defined tolerance $\|\delta\|_p \leq \epsilon_{\mathrm{adv}}$) but causes the model to produce a completely wrong output. A fraud detection model, a medical triage classifier, a credit risk engine — all are vulnerable if their inputs can be manipulated before reaching the inference endpoint.

\subsection{The Gradient-Based Attack Taxonomy}
\label{subsec:gradient_attacks}

The two canonical gradient-based evasion algorithms define the threat baseline that any robust defense must defeat:

\begin{description}
  \item[FGSM (Fast Gradient Sign Method) \cite{goodfellow2015}.]
    A single-step attack that perturbs the input in the direction that maximally increases loss:
    \begin{equation}
      \tilde{x} = x + \epsilon_{\mathrm{adv}} \cdot \operatorname{sign}\!\bigl(\nabla_x \mathcal{L}(\theta, x, y)\bigr)
      \label{eq:fgsm}
    \end{equation}
    Fast but weak: a single step rarely crosses a well-trained decision boundary.

  \item[PGD (Projected Gradient Descent) \cite{madry2018}.]
    An iterative refinement of FGSM — $T$ steps of FGSM followed by projection back onto the $\ell_\infty$ ball:
    \begin{equation}
      x^{(t+1)} = \Pi_{\|x - x_0\|_\infty \leq \epsilon_{\mathrm{adv}}} \Bigl( x^{(t)} + \alpha \cdot \operatorname{sign}\!\bigl(\nabla_{x^{(t)}} \mathcal{L}\bigr) \Bigr)
      \label{eq:pgd}
    \end{equation}
    PGD with $T \geq 20$ steps and random restarts is the standard \textbf{white-box strength benchmark}: if a defense withstands PGD, it passes the baseline robustness bar.
\end{description}

\textbf{Black-box transfer attacks} extend evasion to settings where the adversary cannot access model gradients: they craft adversarial examples against a surrogate model and rely on the empirical observation that adversarial perturbations transfer across architectures sharing the same training distribution.

\subsection{Certified Robustness via Randomized Smoothing}
\label{subsec:certified_robustness}

Adversarial training (retraining the model on PGD-generated examples) provides empirical robustness but no formal guarantee. \textbf{Randomized Smoothing} \cite{cohen2019} provides a mathematical certificate: given a base classifier $f$ and a smoothed classifier $g$ defined by:
\begin{equation}
  g(x) = \arg\max_c \; \Pr_{\varepsilon \sim \mathcal{N}(0, \sigma^2 I)}\!\bigl[f(x + \varepsilon) = c\bigr]
  \label{eq:smoothed_classifier}
\end{equation}
the smoothed classifier $g$ is certifiably robust within $\ell_2$ radius:
\begin{equation}
  r = \frac{\sigma}{2} \Bigl( \Phi^{-1}(p_A) - \Phi^{-1}(p_B) \Bigr)
  \label{eq:certified_radius}
\end{equation}
where $p_A$ is the probability that $f$ predicts the top class and $p_B$ the runner-up, and $\Phi^{-1}$ is the Gaussian quantile function. Any perturbation with $\|\delta\|_2 < r$ cannot change $g$'s prediction.

\begin{lstlisting}[language=Python, style=pythonstyle,
  caption={PGD attack simulation and Randomized Smoothing certification. Certification radius $r > 0$ means no $\ell_2$ perturbation below that radius can flip the prediction.},
  label={lst:robustness-lab}]
import numpy as np
from scipy.stats import norm

# -- PGD Attack (white-box, L_inf) ----------------------------
def pgd_attack(model, x, y, eps=0.03, alpha=0.007,
               steps=20, restarts=5):
    """Returns strongest adversarial example found."""
    best_loss, best_adv = -np.inf, x.copy()
    for _ in range(restarts):
        x_adv = x + np.random.uniform(-eps, eps, x.shape)
        for _ in range(steps):
            grad = model.gradient(x_adv, y)          # white-box
            x_adv = x_adv + alpha * np.sign(grad)
            x_adv = np.clip(x_adv, x - eps, x + eps) # L_inf proj
        loss = model.loss(x_adv, y)
        if loss > best_loss:
            best_loss, best_adv = loss, x_adv.copy()
    return best_adv

# -- Randomized Smoothing Certification -----------------------
def certify(model, x, sigma=0.25, n_samples=1000,
            alpha=0.001):
    """
    Returns (predicted_class, certified_radius).
    radius > 0 means certifiably robust within that L2 ball.
    """
    noisy = x[None] + np.random.randn(n_samples, *x.shape)*sigma
    counts = np.bincount(model.predict_batch(noisy),
                         minlength=model.n_classes)
    c_A = counts.argmax()
    p_A = counts[c_A] / n_samples  # point estimate; use Clopper-Pearson in production
    c_B_count = sorted(counts)[-2]
    p_B = c_B_count / n_samples

    if p_A <= 0.5:
        return c_A, 0.0            # abstain: cannot certify
    radius = sigma / 2 * (norm.ppf(p_A) - norm.ppf(p_B))
    return c_A, max(radius, 0.0)
\end{lstlisting}

\begin{svgraybox}
\textbf{Deployment Standard:} For High-Risk AI systems under the EU AI Act (Article~9), the Sovereign AI Scorecard sets a minimum certified radius of $r \geq 0.25$ (with $\sigma = 0.25$) for safety-critical classifiers. Systems that cannot meet this threshold under $n \geq 1000$ Monte Carlo samples must implement adversarial training before production deployment.
\end{svgraybox}

% ------------------------------------------------------------------
\section{Model Extraction and Watermarking: Protecting Intellectual Property}
\label{sec:model_watermarking}
% ------------------------------------------------------------------

The Chapter~\ref{tab:attack_taxonomy} taxonomy identifies \textbf{Model Extraction} as an inference-phase attack: an adversary queries the victim model systematically, recording input-output pairs sufficient to train a clone (``stolen model'') that replicates the original's functionality. The stolen model bypasses the victim's licensing, access controls, and regulatory certifications.

The defense is \textbf{Model Watermarking}: embedding a verifiable, unforgeable ownership marker into the model's weights that survives fine-tuning and model distillation, enabling the owner to \emph{prove in a legal proceeding} that a suspected stolen model derived from theirs.

\subsection{Two Watermarking Paradigms}
\label{subsec:watermarking_paradigms}

\begin{description}
  \item[Backdoor Watermarking \cite{adi2018}.]
    A set of \textbf{secret trigger inputs} $(x_1^*, \ldots, x_k^*)$ is chosen by the owner before training. The model is trained to produce a specific \textbf{ownership output} $y^*$ for each trigger (e.g., always predict class ``7'' for the trigger images) while behaving normally on all other inputs. Ownership verification: present the triggers to the suspected stolen model; if it produces $y^*$ with probability $\geq \tau$ (typically $\tau = 0.95$), the watermark is intact.

    \emph{Strength:} Survives fine-tuning on the stolen model's distribution. \emph{Weakness:} A sophisticated adversary who discovers the trigger set can retrain to remove it.

  \item[Feature-Space Fingerprinting.]
    Rather than embedding triggers, the owner encodes a cryptographic fingerprint into the distribution of model activations: a statistical signature $\mathbf{s} = h(\theta)$ (e.g., the top-$k$ singular vectors of a specific weight matrix) that is unique to the owner's training run and verifiable against an independently generated reference published in a trusted registry at training time.

    \emph{Strength:} No secret trigger set to discover. \emph{Weakness:} Weaker survival under heavy fine-tuning.
\end{description}

\subsection{Verification Without Exposure: ZK-Proof of Ownership}
\label{subsec:zk_watermark}

A critical limitation of backdoor watermarking is the verification paradox: to prove ownership in court, the owner must reveal the trigger set---which, once public, is worthless as a future watermark. \textbf{Zero-Knowledge Watermark Verification} resolves this: the owner generates a ZK-proof attesting ``the suspected model produces output $y^*$ on the trigger set $\{x_i^*\}$'' without revealing any member of the trigger set. The court verifies the proof; the trigger remains secret for future enforcement actions.

\begin{svgraybox}
\textbf{Architect's Checklist:} Before public API deployment of any proprietary model, (1)~generate and register the feature-space fingerprint $h(\theta)$ with a trusted timestamp authority; (2)~embed a backdoor watermark with $k \geq 100$ trigger pairs; (3)~generate the ZK-ownership proof circuit; (4)~configure the query rate limiter to detect model extraction patterns ($>10^5$ diverse queries from a single API key within 24 hours triggers automatic alerting).
\end{svgraybox}

% ------------------------------------------------------------------
\section{Oblivious Inference: When the Query Itself Is Private}
\label{sec:oblivious_inference}
% ------------------------------------------------------------------

Every attack and defense discussed so far targets the \textbf{content} of training data or model outputs. A structurally distinct privacy threat arises at inference time: even when the model output is safe, the \emph{query itself} reveals sensitive information. A user submitting a query to a cancer-screening API reveals, before the model responds, that they may have cancer-related symptoms. A user querying a mental health chatbot reveals psychological vulnerability. A corporate user querying a competitive intelligence model reveals strategic intent.

This is not a metadata side-channel (Section~\ref{sec:metadata_privacy}), which is about \emph{when} and \emph{how often} queries arrive. It is a \textbf{content-level input privacy} problem: the server learns the raw input before any output is produced. DP-SGD protects training data. ZK-ML protects model weights. Neither protects the query.

\subsection{Three Architectural Solutions}
\label{subsec:oblivious_solutions}

\begin{description}
  \item[Homomorphic Encryption (HE) Inference.]
    The client encrypts its input under a Fully Homomorphic Encryption (FHE) scheme (e.g., CKKS for floating-point neural networks). The server computes inference \emph{on the ciphertext} without ever decrypting, returning an encrypted output that only the client can decrypt. The server learns nothing about the input or the output.

    \emph{Cost:} FHE inference is $10^3$--$10^5\times$ slower than plaintext inference for current neural network depths. Practical for simple classifiers ($\leq 5$ layers); prohibitive for LLMs at 2026 computational costs.

  \item[Batch Oblivious Inference.]
    The client pads its query to a fixed-length vector drawn from a public ``decoy query'' distribution, submits it as part of a \textbf{fixed-size batch} (padding with dummy queries if necessary), and receives the full batch result, discarding the padded outputs. An external observer---and the server's query log---sees only a constant-rate stream of uniform-length batches; individual query content is hidden within the batch.

    \emph{Cost:} Bandwidth overhead proportional to batch size. Latency bounded by batch assembly time. The server still processes each query individually but cannot link any specific query to its client without breaking the batch-level anonymity assumption.

  \item[Private Information Retrieval (PIR).]
    In settings where the model is used for lookup rather than generation (e.g., drug interaction checks, regulatory compliance databases), \textbf{PIR protocols} \cite{chor1998} allow a client to retrieve the record at index $i$ from a database of $N$ records without the server learning $i$. For two non-colluding servers holding identical model outputs, the client queries each server with complementary random masks; the XOR of the two responses yields the desired output. Single-server PIR (computationally secure) exists but is more expensive.
\end{description}

\begin{table}[ht]
\centering
\caption{Oblivious Inference Solutions: Practical Trade-offs for 2026 Deployment}
\label{tab:oblivious_inference}
\begin{tabular}{p{3.5cm} p{3.0cm} p{2.5cm} p{3.0cm}}
\hline\noalign{\smallskip}
\textbf{Approach} & \textbf{Privacy Guarantee} & \textbf{Latency Overhead} & \textbf{Practical Use Case} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
FHE Inference & Perfect (computational) & $10^3\times$ -- $10^5\times$ & Simple classifiers; regulatory compliance checks \\[4pt]
Batch Oblivious Inference & $k$-anonymity within batch & $O(k)$ batch assembly & Healthcare query APIs; mental health chatbots \\[4pt]
Two-Server PIR & Information-theoretic (2-server) & $O(\sqrt{N})$ communication & Drug interaction DBs; legal reference lookups \\[4pt]
TEE + Encrypted Channel & Trust shifts to hardware & $\approx 10$--$20$\% & General LLM inference with hardware attestation \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\begin{svgraybox}
\textbf{Pragmatic Recommendation (2026):} For most production deployments, \textbf{Batch Oblivious Inference combined with Traffic Padding} (Section~\ref{subsec:traffic_padding}) provides the best privacy-cost trade-off: the server cannot link query content to individual clients (batch oblivion) and cannot infer clinical state from call frequency (traffic padding). FHE inference should be reserved for the specific case where the server itself is untrusted---a constraint that TEEs (Chapter~\ref{chap:tee}) often satisfy more efficiently at current hardware costs.
\end{svgraybox}

% ------------------------------------------------------------------
\section{The Unlearning Vulnerability: When Forgetting Reveals}
\label{sec:unlearning_vulnerability}
% ------------------------------------------------------------------

The Right to be Forgotten---enshrined in GDPR Article 17 and Vietnam's Decree 13/2023---requires removing a user's data not just from the database but from the model weights. Chapter~\ref{chap:machine_unlearning} provides the full Machine Unlearning toolkit. Here we identify a critical attack surface created by the \emph{process} of unlearning itself: the \textbf{Unlearning Vulnerability}.

\subsection{The Model-Comparison Attack}
\label{subsec:model_comparison}

Suppose an adversary has access to two consecutive model checkpoints: $\theta_{\text{v1}}$ (before unlearning) and $\theta_{\text{v2}}$ (after unlearning)---both potentially available through a public inference API or a versioned model registry. If the adversary knows that exactly one record was deleted between v1 and v2, they can compare the two models to identify the deleted record by examining the gradient difference:

\begin{equation}
\Delta\theta = \theta_{\text{v2}} - \theta_{\text{v1}}
\label{eq:unlearning_delta}
\end{equation}

$\Delta\theta$ concentrates around the parameter dimensions most influenced by the deleted record. The adversary identifies the deleted record $r^*$ from a candidate dataset $D_\mathrm{cand}$ as:

\begin{equation}
r^* = \arg\max_{r \in D_\mathrm{cand}} \; \bigl\langle \nabla_\theta \mathcal{L}(\theta_{\text{v1}}, r),\; \Delta\theta \bigr\rangle
\label{eq:unlearning_attack}
\end{equation}

This attack is particularly effective against \textbf{Approximate Unlearning} (influence functions, gradient ascent), where $\Delta\theta$ is precisely proportional to the target record's gradient---essentially surfacing the deleted record's training contribution.

\subsection{Architectural Countermeasures}
\label{subsec:unlearning_countermeasures}

\begin{description}
  \item[Batch Unlearning:] Accumulate deletion requests and execute unlearning in batches of $B \geq 10$ records simultaneously, making $\Delta\theta$ a superposition of multiple records' gradients and preventing individual identification.
  \item[DP-Unlearning:] Apply calibrated Gaussian noise to $\Delta\theta$ before committing the new weights, providing a formal $(\varepsilon_u, \delta_u)$-DP guarantee on the unlearning operation itself.
  \item[Version Access Control:] Restrict inference API access to the current model version only. Revoke old checkpoint access immediately upon unlearning; never expose a versioned API enabling side-by-side comparison.
  \item[Randomized Confirmation Delay:] Add a uniformly random delay $d \sim \mathcal{U}[d_{\min}, d_{\max}]$ between receiving a deletion request and executing unlearning, preventing the adversary from precisely pinpointing the version boundary.
\end{description}

% ------------------------------------------------------------------
\section{Part II Attack-Defense Cross-Reference Matrix}
\label{sec:attack_defense_matrix}
% ------------------------------------------------------------------

Before closing Part II, Table~\ref{tab:attack_defense_matrix} consolidates the full attack landscape into a single reference that maps each threat to its primary architectural defense, the section where the defense is specified, and the Sovereign AI Scorecard metric (Chapter~\ref{chap:attm}) that verifies adequate protection in a production audit. Use this table as the starting point when threat-modeling a new deployment: identify which attack vectors are in-scope, confirm each has a corresponding control, and verify the Scorecard metric passes before release.

\begin{table}[ht]
\centering
\caption{Part II Attack-Defense Cross-Reference Matrix. Defense section references are within this book. Scorecard metric thresholds follow the Sovereign AI Scorecard (Chapter~\ref{chap:attm}); FAIL thresholds indicate a mandatory remediation gate before production deployment.}
\label{tab:attack_defense_matrix}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{p{2.4cm} p{2.4cm} p{3.2cm} p{3.9cm}}
\hline\noalign{\smallskip}
\textbf{Attack} & \textbf{Phase} & \textbf{Primary Defense} & \textbf{Scorecard Metric / Pass Threshold} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Membership Inference (MIA) & Inference & DP-SGD ($\varepsilon \leq 2.0$); Ch.~5 & MIA AUC $\leq 0.60$; FAIL if $>$0.70 \\[3pt]
Model Inversion / Gradient Leakage & Inference & DP-SGD gradient clipping; SecAgg (Ch.~8) & Reconstruction PSNR $<$ 20\,dB on held-out set \\[3pt]
Data Poisoning & Training & 3-Stage Poison Detection Pipeline (§\ref{subsec:poison_detection}); Krum aggregation (Ch.~10) & Honeypot loss divergence $< 2\sigma$; flagged sample rate $< 1\%$ \\[3pt]
Evasion (Adversarial Examples) & Inference & Adversarial training + Randomized Smoothing (§\ref{subsec:certified_robustness}) & Certified radius $r \geq 0.25$ at $\sigma = 0.25$ \\[3pt]
Model Extraction & Inference & Model Watermarking + ZK-Proof of Ownership (§\ref{sec:model_watermarking}) & Backdoor watermark survival rate $\geq 95\%$ post fine-tune \\[3pt]
Metadata / Side-Channel & Inference & Traffic Padding + Constant-Time Execution (§\ref{sec:metadata_privacy}) & Timing variance $< 1$\,ms; call-frequency correlation $\rho < 0.05$ \\[3pt]
Unlearning Vulnerability & Post-unlearning & Batch Unlearning ($B \geq 10$) + DP-Unlearning + Version Access Control (§\ref{sec:unlearning_vulnerability}) & Model-comparison attack precision $< 0.55$ \\[3pt]
Oblivious Inference (query privacy) & Inference & Batch Oblivious Inference + Traffic Padding (§\ref{sec:oblivious_inference}) & Server query-client linkability rate $< 1/k$ \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

% ------------------------------------------------------------------
\section{Full-Stack Blueprint: Secure Medical Research Data Pipeline}
\label{sec:blueprint_medical_research}
% ------------------------------------------------------------------

To conclude Part II, we synthesize the three Privacy Engineering tools---k-Anonymity, Differential Privacy, and the attack landscape---into a single \textbf{Full-Stack Blueprint} for a multi-hospital cancer screening research consortium.

In this pipeline, raw patient records are first sanitized through a \textbf{k-Anonymity pre-filter} ($k=10$), removing obvious quasi-identifiers that enable re-identification. The resulting dataset is then fed into a \textbf{DP-SGD training loop} with a carefully calibrated privacy budget ($\varepsilon = 2.0$, $\delta = 10^{-5}$) that injects Gaussian noise to defeat Membership Inference Attacks (Chapter 6). The Security Layer enforces \textbf{AES-256 encryption at rest} for the data lake. The Verifiability Layer maintains a \textbf{Privacy Budget Ledger}---an immutable log of every query and the epsilon consumed---ensuring the consortium can demonstrate regulatory compliance. Finally, the Alignment Layer gates every data release through an \textbf{Institutional Review Board (IRB) approval step}, closing the accountability loop before any research output leaves the system.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    node distance=1.0cm and 3.0cm,
    font=\small,
    layer_box/.style={draw, rounded corners, inner sep=8pt, text centered, font=\bfseries\small, minimum height=1.2cm},
    align_box/.style={layer_box, fill=purple!10, draw=purple!80!black},
    priv_box/.style={layer_box,  fill=blue!10,   draw=blue!80!black},
    sec_box/.style={layer_box,   fill=red!10,    draw=red!80!black},
    ver_box/.style={layer_box,   fill=green!10,  draw=green!80!black},
    arrow/.style={->, >=stealth, thick, draw=gray!70}
]

\node (kanon) [priv_box, text width=5.5cm]
    {\textbf{1. k-Anonymity Pre-filter}\\(Privacy Layer, $k=10$)\\
     Suppresses quasi-identifiers};

\node (dp)    [priv_box, below=1.2cm of kanon, text width=5.5cm]
    {\textbf{2. DP-SGD Training Loop}\\(Privacy Layer, $\varepsilon{=}2.0$)\\
     Gaussian noise injection};

\node (enc)   [sec_box, below=1.2cm of dp, text width=5.5cm]
    {\textbf{3. Encrypted Data Lake}\\(Security Layer, AES-256)\\
     Data-at-rest protection};

\node (ledger) [ver_box, right=3cm of kanon, text width=5.5cm]
    {\textbf{4. Privacy Budget Ledger}\\(Verifiability Layer)\\
     Immutable $\varepsilon$-query audit log};

\node (irb)   [align_box, right=3cm of enc, text width=5.5cm]
    {\textbf{5. IRB Approval Gate}\\(Alignment Layer)\\
     Human ethical review\\before data release};

\draw [arrow] (kanon.south) -- (dp.north);
\draw [arrow] (dp.south)    -- (enc.north);
\draw [arrow] (dp.east)     -- node[above, font=\scriptsize] {Logs $\varepsilon$ consumed} (ledger.west);
\draw [arrow] (enc.east)    -- node[above, font=\scriptsize] {Release request} (irb.west);

\end{tikzpicture}
\caption{Full-Stack Blueprint (Part II): A privacy-preserving medical research pipeline integrating k-Anonymity pre-filtering, DP-SGD noise calibration ($\varepsilon{=}2.0$), AES-256 encrypted storage, an immutable Privacy Budget Ledger, and an IRB approval gate.}
\label{fig:blueprint_medical_research}
\end{figure}

\section{Conclusion}
\label{sec:conclusion_attacks}

Part II has provided the essential "Data Defense" toolkit. We have seen how simple anonymization fails (Chapter~4), how Differential Privacy provides a mathematical shield (Chapter~5), and the full spectrum of attacks that shield must withstand (Chapter~6): Membership Inference, Model Inversion, Evasion, Data Poisoning, Model Extraction, metadata side-channels, the Unlearning Vulnerability, and the inference-time query-privacy problem of Oblivious Inference. For each attack, we have paired an architectural countermeasure---from DP-SGD to Randomized Smoothing to Batch Oblivious Inference to Model Watermarking.

In Part III, we will expand our perspective from single-point data defense to \textbf{Secure Distributed Systems}, exploring how to orchestrate training across multiple locations without ever moving the raw data through \textbf{Federated Learning}.


