%%%%%%%%%%%%%%%%%%%%% Chapter 4 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Classic Anonymization & The Theory of Quasi-Identifiers
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\motto{``We removed all names and IDs. The data is anonymous now.'' This is the most common, and most dangerous, assumption in data engineering. Trust requires a rigorous mapping of identifiers, not just their removal.}

\chapter{Classic Anonymization \& The Theory of Quasi-Identifiers}
\label{chap:classic_anonymization}

\abstract{In the world of Trustworthy AI Architecture, Chapter 4 is where a project lives or dies. If anonymization is too weak, the system faces catastrophic re-identification; if it is too aggressive, the model's utility vanishes. This chapter provides a rigorous analysis of data transformation versus mathematical privacy guarantees. We define the taxonomy of attributes, analyze the ``K-L-T'' hierarchy of protection, and explore architectural implementation patterns such as Referential Integrity and the Pareto Frontier of utility. We then present formal definitions, algorithms, security proofs, and worked examples for $k$-anonymity and $\ell$-diversity. Finally, we map these techniques to the 2026 standards established by Vietnam's Decree 356.}

\section{The Taxonomy of Attributes}
\label{sec:taxonomy_attributes}

To build a secure anonymization pipeline, the architect must first classify every column in the dataset. We formalize a dataset as a set of tuples $T(ID, QI, SA)$:

\begin{description}
  \item[Identifiers (ID):] Explicit keys such as Full Names or Social Security Numbers. These must be entirely suppressed or replaced with unique tokens.
  \item[Quasi-Identifiers (QI):] Attributes that are not unique on their own but become a ``fingerprint'' when combined. In a 2026 urban environment, a combination of \textit{Birthdate + Gender + ZIP code} can uniquely identify up to 87\% of individuals \cite{sweeney2002}.
  \item[Sensitive Attributes (SA):] The private information we want to protect while still allowing for statistical analysis (e.g., medical diagnosis, salary).
\end{description}

% ------------------------------------------------------------------
\subsection{Quasi-Identifier Discovery: A Systematic Heuristic}
\label{subsec:qi_discovery}
% ------------------------------------------------------------------

Knowing the formal definition of a quasi-identifier is necessary but not sufficient. In practice, every real-world dataset contains dozens of columns, and the architect must determine which subset constitutes the quasi-identifier set $QI$ \emph{before} selecting $k$. Underestimating $QI$ leads to the Linkage Attack failure mode above; overestimating it forces aggressive generalization that destroys utility. The following four-step discovery heuristic is recommended before any anonymization pipeline is initialized.

\begin{enumerate}
  \item \textbf{Column-type classification.} Separate columns into Identifiers (suppress unconditionally: name, SSN, CCCD), Candidate QIs (demographic, geographic, temporal, or behavioral), and Sensitive Attributes (the target of protection). Any column that is (a) potentially available in an external public dataset and (b) not sensitive in its own right is a candidate QI.
  \item \textbf{External linkability audit.} For each candidate QI, query: \emph{Does a publicly available dataset contain this column together with any identifier?} Vietnam-specific sources to check: voter registration data, business registration databases (Cong Thong Tin), social media profile fields. If yes, classify as \textbf{confirmed QI}.
  \item \textbf{Uniqueness test.} For the confirmed QI set, compute the fraction of records that are \emph{unique} on the full cross-product of QI values: $\mathrm{UniquenessRatio}(QI) = |\{t \in T : |EQ(t)| = 1\}| / |T|$. If UniquenessRatio $> 0.01$ (i.e., more than 1\% of records are already unique before any linkage), the QI set is too specific for $k < 20$.
  \item \textbf{Sensitivity expansion check.} Run the Data Food Chain analysis (Section~\ref{sec:data_food_chain}): can any non-sensitive QI column be used to infer the sensitive attribute directly (e.g., ZIP code $\Rightarrow$ income bracket $\Rightarrow$ financial SA)? If the mutual information $I(\text{QI}_i; SA) > 0.1$ nats, elevate $\text{QI}_i$ to a sensitive attribute or apply $t$-closeness.
\end{enumerate}

\begin{svgraybox}
\textbf{Rule of Thumb:} For a Vietnamese healthcare dataset in 2026, a safe baseline $QI$ set is $\{\text{Age-band (5yr)}, \text{Gender}, \text{District}\}$ with $k \geq 15$ (Decree 356 Class B). Adding Occupation or Employer to any published companion dataset immediately upgrades the effective $QI$ set and requires re-evaluation. Run the uniqueness test after every new auxiliary source is identified.
\end{svgraybox}

\section{The K-L-T Hierarchy: Modern Standards}
\label{sec:klt_hierarchy}

To solve the \textbf{Homogeneity Attack} (where all $k$-anonymous subjects share the same diagnosis), we implement $l$-diversity. This requires that each $k$-anonymous group has at least $l$ ``well-represented'' values for the Sensitive Attribute.

% ======================================================================
\subsection{Formal Foundations: $k$-Anonymity and $\ell$-Diversity}
\label{sec:formal_foundations}

The proliferation of large-scale data collection by governments, healthcare systems, and private organisations has created unprecedented opportunities for statistical analysis and research. At the same time it has raised profound concerns about individual privacy. When a data holder wishes to share a microdata table---one row per person---with an external researcher or the public, sensitive attributes such as medical diagnoses, salary, or criminal record may be inferred by an adversary who already possesses partial background knowledge.

Two fundamental formal models address this threat:

\begin{itemize}
  \item \textbf{$k$-Anonymity} (Samarati \& Sweeney, 1998/2002) \cite{sweeney2002}: guarantees that every record is indistinguishable from at least $k-1$ others with respect to quasi-identifier attributes.
  \item \textbf{$\ell$-Diversity} (Machanavajjhala \emph{et al.}, 2006) \cite{machanavajjhala2007}: strengthens $k$-anonymity by requiring diversity within the sensitive attribute of each equivalence class, thereby resisting \emph{homogeneity} and \emph{background-knowledge} attacks.
\end{itemize}

This section provides formal definitions, illustrative examples, algorithmic treatments, security analyses, and a critical comparison of these two models together with their principal extensions.

\subsubsection{Preliminaries and Data Model}
\label{sec:data_model}

\paragraph{Microdata Tables.}

A \emph{microdata table} is a relation $T = (A_1, \ldots, A_n)$ where each tuple $t \in T$ represents a single individual. The attribute set is partitioned into three disjoint groups:
\begin{align*}
  \text{Explicit Identifiers} &: \mathcal{I} \subseteq \{A_1,\ldots,A_n\}\\
  \text{Quasi-Identifiers}    &: QI \subseteq \{A_1,\ldots,A_n\} \setminus \mathcal{I}\\
  \text{Sensitive Attributes} &: SA \subseteq \{A_1,\ldots,A_n\} \setminus (\mathcal{I}\cup QI)
\end{align*}

\begin{itemize}
  \item \textbf{Explicit identifiers} (e.g.\ name, SSN) are suppressed before any release.
  \item \textbf{Quasi-identifiers} (e.g.\ age, gender, ZIP code) can be linked to external databases to re-identify individuals.
  \item \textbf{Sensitive attributes} (e.g.\ disease, salary) represent the information the data holder wishes to protect.
\end{itemize}

\paragraph{Generalisation and Suppression.}

A \emph{generalisation} of a domain $D$ is a surjective mapping $\phi: D \to D'$ such that $|D'| \leq |D|$. Repeated application yields a \emph{generalisation hierarchy} $\mathcal{H}_A$ for attribute $A$. Replacing every value $v$ of $A$ in $T$ by $\phi(v)$ produces the \emph{generalised table} $T^{*}$.

Common generalisation strategies include:
\begin{itemize}
  \item \textit{Value generalisation}: replace a specific value with a more abstract ancestor (e.g.\ \texttt{29} $\to$ \texttt{[25,30)}).
  \item \textit{Suppression}: replace a value with a special symbol $\ast$, representing maximal generalisation.
\end{itemize}

% ----------------------------------------------------------------------
\subsubsection{$k$-Anonymity: Formal Definition and Analysis}
\label{sec:kanon_formal}

\paragraph{Equivalence Classes.}
Given a quasi-identifier $QI$ and a (possibly generalised) table $T^*$, the \emph{equivalence class} of a tuple $t \in T^*$ is
\[
  EQ(t) = \{t' \in T^* \mid t'[QI] = t[QI]\}.
\]
The set $\{EQ(t) \mid t \in T^*\}$ partitions $T^*$ into equivalence classes.

\paragraph{Definition.}
A table $T^*$ satisfies \emph{$k$-anonymity} with respect to quasi-identifier $QI$ if and only if
\[
  \forall\, t \in T^* : |EQ(t)| \geq k.
\]

Intuitively, each individual ``hides'' among at least $k-1$ others who share identical quasi-identifier values. The parameter $k$ is a privacy budget: a larger $k$ offers stronger privacy at the cost of greater information loss.

\paragraph{Illustrative Example ($k=3$).}

Consider the original table:

\begin{center}
\begin{tabular}{p{2cm}cccp{3cm}}
\hline\noalign{\smallskip}
\textbf{Name} & \textbf{Age} & \textbf{Gender} & \textbf{ZIP} & \textbf{Disease}\\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Alice   & 29 & F & 13053 & Flu\\
Bob     & 30 & M & 13068 & Cancer\\
Carol   & 29 & F & 13053 & Heart Disease\\
Dave    & 45 & M & 14853 & Flu\\
Eve     & 47 & M & 14853 & Cancer\\
Frank   & 46 & M & 14850 & Heart Disease\\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{center}

After generalising Age $\to$ [25,35) / [45,50) and ZIP $\to$ 130** / 148**:

\begin{center}
\begin{tabular}{cccp{3cm}}
\hline\noalign{\smallskip}
\textbf{Age} & \textbf{Gender} & \textbf{ZIP} & \textbf{Disease}\\
\noalign{\smallskip}\svhline\noalign{\smallskip}
{[}25,35) & F & 130** & Flu\\
{[}25,35) & M & 130** & Cancer\\
{[}25,35) & F & 130** & Heart Disease\\
{[}45,50) & M & 148** & Flu\\
{[}45,50) & M & 148** & Cancer\\
{[}45,50) & M & 148** & Heart Disease\\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{center}

Each equivalence class has size 3, so the table is 3-anonymous.

\paragraph{Attacks Against $k$-Anonymity.}

Despite its appeal, $k$-anonymity is vulnerable to two well-known attacks when the distribution of sensitive values within an equivalence class is skewed.

\paragraph{Homogeneity Attack.}
If all tuples in an equivalence class share the same sensitive value, an adversary learns the sensitive attribute of every member with certainty---even without any background knowledge. For example, suppose a 3-anonymous equivalence class contains tuples whose \texttt{Disease} field is \{Cancer, Cancer, Cancer\}. An adversary who links any member to this class immediately learns that the individual has cancer.

\paragraph{Background-Knowledge Attack.}
An adversary may exploit external knowledge about the relative likelihood of sensitive values to dramatically narrow the uncertainty even when values are not homogeneous. For example, suppose a class contains \{Flu, Heart Disease, Cancer\} but the adversary knows the individual is a non-smoker. If background statistics show that non-smokers have $<1\%$ chance of lung cancer, the effective uncertainty collapses to two values.

% ----------------------------------------------------------------------
\subsubsection{$\ell$-Diversity: Formal Definition and Analysis}
\label{sec:ldiv_formal}

\paragraph{Motivation.}
$\ell$-Diversity (Machanavajjhala \emph{et al.}, 2006) \cite{machanavajjhala2007} was proposed to address the homogeneity and background-knowledge attacks. The core intuition is that merely ensuring indistinguishability on quasi-identifiers is insufficient; sensitive values within each equivalence class must themselves be sufficiently \emph{diverse}.

\paragraph{Definition.}
Let $q^*$ be an equivalence class and let $S(q^*)$ denote the multiset of sensitive attribute values in $q^*$. The class $q^*$ satisfies \emph{$\ell$-diversity} if it contains at least $\ell$ \emph{well-represented} sensitive values. A table $T^*$ satisfies $\ell$-diversity if every equivalence class satisfies $\ell$-diversity.

The notion of ``well-represented'' gives rise to three principal instantiations:

\paragraph{Distinct $\ell$-Diversity.}
An equivalence class satisfies \emph{distinct $\ell$-diversity} if
\[
  |\{v \mid v \in S(q^*)\}| \geq \ell,
\]
i.e.\ there are at least $\ell$ \emph{distinct} sensitive values.

\paragraph{Entropy $\ell$-Diversity.}
An equivalence class satisfies \emph{entropy $\ell$-diversity} if
\[
  H(S(q^*)) = -\sum_{s \in \mathcal{S}} p(q^*, s)\log p(q^*, s) \;\geq\; \log\ell,
\]
where $p(q^*, s) = |\{t \in q^* \mid t[SA] = s\}| / |q^*|$ and $\mathcal{S}$ is the domain of $SA$.

Entropy $\ell$-diversity is strictly stronger than distinct $\ell$-diversity: it penalises equivalence classes that nominally contain $\ell$ values but concentrate probability mass on a single value.

\paragraph{Recursive $(c,\ell)$-Diversity.}
Let the sensitive values in equivalence class $q^*$ be sorted by frequency: $r_1 \geq r_2 \geq \cdots \geq r_m$. The class satisfies \emph{recursive $(c,\ell)$-diversity} if
\[
  r_1 < c\,(r_\ell + r_{\ell+1} + \cdots + r_m),
\]
for constants $c > 0$ and integer $\ell \geq 2$. This prevents the most common value from dominating even when $\ell$ distinct values are present.

\paragraph{Relationship Between Variants.}

For any equivalence class $q^*$:
\[
  \text{Entropy-}\ell\text{-diversity}
  \;\Rightarrow\;
  \text{Distinct-}\ell\text{-diversity}.
\]
Moreover, recursive $(c, \ell)$-diversity with $c \leq 1/(|\mathcal{S}|-\ell+1)$ implies distinct $\ell$-diversity. The converse implications do not hold in general.

If $H(S(q^*)) \geq \log\ell$ then by the maximum-entropy bound the class must have support on at least $\ell$ distinct values, otherwise the entropy would be at most $\log(\ell - 1) < \log \ell$.

\paragraph{Illustrative Example: $\ell$-Diversity vs.\ $k$-Anonymity.}

Consider two 3-anonymous equivalence classes:

\begin{center}
\begin{tabular}{p{2cm}cc}
\hline\noalign{\smallskip}
 & \textbf{Class A} & \textbf{Class B}\\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Tuple 1 & Cancer & Cancer\\
Tuple 2 & Cancer & Heart Disease\\
Tuple 3 & Cancer & Flu\\
\noalign{\smallskip}\hline\noalign{\smallskip}
Distinct $\ell$ & 1 & 3\\
Entropy $H$     & 0 & $\log 3 \approx 1.585$\\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{center}

Class A satisfies 3-anonymity but \emph{fails} distinct 2-diversity (all values identical). Class B satisfies 3-anonymity \emph{and} distinct 3-diversity (and entropy 3-diversity).

% ----------------------------------------------------------------------
\subsubsection{Algorithms for Achieving $k$-Anonymity and $\ell$-Diversity}
\label{sec:kanon_algorithms}

\paragraph{The Mondrian Algorithm for $k$-Anonymity.}

Mondrian (LeFevre \emph{et al.}, 2006) is a greedy, multidimensional partitioning algorithm for $k$-anonymity. It operates on the full quasi-identifier space without requiring pre-specified hierarchies. The algorithm proceeds as follows:

\begin{enumerate}
  \item If $|T| < 2k$, return $\{T\}$ (cannot split further).
  \item Select the attribute $A^* = \arg\max_{A \in QI} \mathrm{NormalizedRange}(T, A)$.
  \item Compute the median $m$ of $T[A^*]$.
  \item Split: $T_L = \{t \in T \mid t[A^*] \leq m\}$, $T_R = \{t \in T \mid t[A^*] > m\}$.
  \item If $|T_L| \geq k$ and $|T_R| \geq k$, recurse on both partitions.
  \item Otherwise, return $\{T\}$.
\end{enumerate}

\textbf{Complexity.} Each recursive call partitions the data along the median of the widest attribute. With $n$ tuples and $d = |QI|$ dimensions the depth is $O(\log n)$ and the total cost is $O(nd \log n)$.

\paragraph{Extended Mondrian for $\ell$-Diversity.}

To enforce $\ell$-diversity, we augment the split condition: a split is only performed if \emph{both} child partitions satisfy $\ell$-diversity. The diversity check depends on the chosen variant:
\begin{itemize}
  \item \textbf{Distinct}: $|\{t[SA] \mid t \in T\}| \geq \ell$
  \item \textbf{Entropy}: $H(T[SA]) \geq \log \ell$
  \item \textbf{Recursive $(c)$}: recursive $(c, \ell)$-condition on the frequency vector
\end{itemize}

\paragraph{Incognito Algorithm.}

Incognito (LeFevre \emph{et al.}, 2005) performs a bottom-up lattice traversal over all possible generalisation combinations. For each node in the lattice (a specific generalisation of each QI attribute), it checks whether the resulting table satisfies $k$-anonymity. The algorithm exploits the \emph{monotone property}: if a generalisation fails $k$-anonymity, any less-generalised descendant also fails.
\[
  \text{If } T^*(G) \text{ violates } k\text{-anonymity, then } \forall G' \preceq G,\; T^*(G') \text{ also violates } k\text{-anonymity.}
\]
This allows large portions of the lattice to be pruned, reducing the search space from $O(\prod_i |H_{A_i}|)$ to a manageable size in practice.

% ----------------------------------------------------------------------
\subsubsection{Information Loss Metrics}
\label{sec:info_loss_metrics}

Anonymisation necessarily degrades data utility. The following metrics quantify information loss (IL).

\paragraph{Normalised Certainty Penalty (NCP).}
For a numerical attribute $A$ with domain $[A_{\min}, A_{\max}]$ and an equivalence class $q^*$ that generalises $A$ to the interval $[q^*_{\min}, q^*_{\max}]$:
\[
  \mathrm{NCP}_A(q^*) = \frac{q^*_{\max} - q^*_{\min}}{A_{\max} - A_{\min}}.
\]
The overall NCP is averaged over all attributes and all tuples:
\[
  \mathrm{NCP}(T^*) = \frac{1}{n \cdot |QI|} \sum_{t \in T^*} \sum_{A \in QI} \mathrm{NCP}_A(EQ(t)).
\]
$\mathrm{NCP} = 0$ means no loss; $\mathrm{NCP} = 1$ means complete suppression.

\paragraph{Discernibility Metric (DM).}
\[
  \mathrm{DM}(T^*) = \sum_{q^* \in \Pi} |q^*|^2.
\]
Larger equivalence classes incur higher penalty, incentivising algorithms to create balanced partitions.

\paragraph{Average Equivalence Class Size (AECS).}
\[
  \mathrm{AECS}(T^*) = \frac{n}{|\Pi|}.
\]
A low AECS (close to $k$) indicates minimal over-generalisation.

% ----------------------------------------------------------------------
\subsubsection{Security Analysis}
\label{sec:security_analysis}

\paragraph{Privacy Guarantees of $k$-Anonymity.}
Under $k$-anonymity, assuming an adversary with perfect knowledge of the quasi-identifier values of a target individual but no other information, the probability of correctly identifying the target's record is at most $1/k$.

The adversary's target falls into some equivalence class $EQ(t)$ with $|EQ(t)| \geq k$. Each member of the class is indistinguishable on $QI$, so under a uniform prior the probability of selecting the correct record is $1/|EQ(t)| \leq 1/k$.

\paragraph{Limitations of $k$-Anonymity.}
$k$-Anonymity provides \emph{no} guarantee on the posterior probability of a specific sensitive value given the quasi-identifier. In the worst case (all records in a class share the same sensitive value), the posterior equals the prior of 1---i.e.\ the adversary gains full information about the sensitive attribute.

\paragraph{Privacy Guarantees of $\ell$-Diversity.}
Under entropy $\ell$-diversity, for any equivalence class $q^*$ and any sensitive value $s$:
\[
  p(q^*, s) \leq \frac{1}{\ell}.
\]
Consequently, the \emph{positive disclosure probability}---the probability that an adversary correctly infers $s$ as the target's sensitive value---is at most $1/\ell$.

\paragraph{Remaining Vulnerabilities of $\ell$-Diversity.}

\begin{itemize}
  \item \textbf{Skewness attack}: If the overall distribution of a sensitive value is very low (e.g.\ only 1\% of people have HIV), an equivalence class with 1-in-3 HIV records still leaks significantly compared to the population prior.
  \item \textbf{Similarity attack}: If ``diverse'' sensitive values are semantically similar (e.g.\ \{stomach cancer, colon cancer, lung cancer\}), the adversary still learns the individual has cancer.
\end{itemize}

These vulnerabilities motivated $t$-closeness \cite{li2007}, which requires the distribution of sensitive values in each class to be close (in Earth Mover's Distance) to the overall distribution in $T$.

% ----------------------------------------------------------------------
\subsubsection{Summary Comparison of Privacy Models}
\label{sec:comparison_privacy_models}

\begin{table}
\caption{Comparison of privacy models}
\label{tab:compare_privacy}
\centering
\begin{tabular}{p{4cm}cccc}
\hline\noalign{\smallskip}
\textbf{Property} & \textbf{$k$-Anon.} & \textbf{$\ell$-Div.} & \textbf{$t$-Close.} & \textbf{Diff.\ Privacy}\\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Re-identification protection & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
Homogeneity-attack resistance & $\times$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
Skewness-attack resistance    & $\times$ & Partial & $\checkmark$ & $\checkmark$\\
Similarity-attack resistance  & $\times$ & $\times$ & $\checkmark$ & $\checkmark$\\
Composition theorems          & $\times$ & $\times$ & $\times$ & $\checkmark$\\
Auxiliary-info independence   & $\times$ & $\times$ & $\times$ & $\checkmark$\\
Exact data output             & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\times$\\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

% ----------------------------------------------------------------------
\subsubsection{Worked End-to-End Example}
\label{sec:worked_example}

We illustrate the full anonymisation pipeline on a small medical dataset.

\paragraph{Original Data.}

\begin{center}
\begin{tabular}{clccp{3cm}}
\hline\noalign{\smallskip}
\textbf{ID} & \textbf{Age} & \textbf{Gender} & \textbf{ZIP} & \textbf{Disease}\\
\noalign{\smallskip}\svhline\noalign{\smallskip}
1 & 21 & F & 47677 & Ovarian Cancer\\
2 & 22 & F & 47602 & Ovarian Cancer\\
3 & 27 & M & 47678 & Prostate Cancer\\
4 & 24 & M & 47905 & Flu\\
5 & 35 & M & 47909 & Prostate Cancer\\
6 & 36 & M & 47906 & Flu\\
7 & 53 & F & 47605 & Flu\\
8 & 57 & F & 47673 & Flu\\
9 & 56 & M & 47607 & Heart Disease\\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{center}

\paragraph{Step 1: Remove Explicit Identifiers.}
The \textbf{ID} column is suppressed entirely.

\paragraph{Step 2: Apply 3-Anonymity.}
We generalise Age and ZIP to obtain equivalence classes of size $\geq 3$:

\begin{center}
\begin{tabular}{cccp{3cm}}
\hline\noalign{\smallskip}
\textbf{Age} & \textbf{Gender} & \textbf{ZIP} & \textbf{Disease}\\
\noalign{\smallskip}\svhline\noalign{\smallskip}
{[}20,30) & F & 476** & Ovarian Cancer\\
{[}20,30) & F & 476** & Ovarian Cancer\\
{[}20,30) & M & 476** & Prostate Cancer\\
{[}20,30) & M & 479** & Flu\\
{[}30,40) & M & 479** & Prostate Cancer\\
{[}30,40) & M & 479** & Flu\\
{[}50,60) & F & 476** & Flu\\
{[}50,60) & F & 476** & Flu\\
{[}50,60) & M & 476** & Heart Disease\\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{center}

\paragraph{Step 3: Check $\ell$-Diversity.}
Partition into equivalence classes based on identical QI values:

\begin{center}
\begin{tabular}{p{1.5cm}p{2cm}p{7cm}}
\hline\noalign{\smallskip}
\textbf{Class} & \textbf{Tuples} & \textbf{Diseases}\\
\noalign{\smallskip}\svhline\noalign{\smallskip}
$q_1$ & 1,2,3 & \{Ovarian Cancer, Ovarian Cancer, Prostate Cancer\}\\
$q_2$ & 4,5,6 & \{Flu, Prostate Cancer, Flu\}\\
$q_3$ & 7,8,9 & \{Flu, Flu, Heart Disease\}\\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{center}

\textbf{Checking distinct 2-diversity:}
\begin{itemize}
  \item $q_1$: \{Ovarian Cancer, Prostate Cancer\} $\to$ 2 distinct values. $\checkmark$
  \item $q_2$: \{Flu, Prostate Cancer\} $\to$ 2 distinct values. $\checkmark$
  \item $q_3$: \{Flu, Heart Disease\} $\to$ 2 distinct values. $\checkmark$
\end{itemize}

\textbf{Checking entropy 2-diversity} ($H \geq \log 2 = 1$ bit):
\begin{align*}
  H(q_1) &= -\tfrac{2}{3}\log\tfrac{2}{3} - \tfrac{1}{3}\log\tfrac{1}{3} \approx 0.918 < 1.
\end{align*}
$q_1$ \emph{fails} entropy 2-diversity. We must re-partition or apply additional generalisation (e.g.\ merge Gender into $*$ or merge $q_1$ and $q_2$) until all classes pass.

\paragraph{Step 4: Remediation.}
Merge $q_1$ (size 3) with $q_2$ (size 3) into a single class of size 6:
\[
  q_{12}: \{\text{Ovarian Cancer, Ovarian Cancer, Prostate Cancer, Flu, Prostate Cancer, Flu}\}
\]
\[
  H(q_{12}) = -\tfrac{2}{6}\log\tfrac{2}{6} - \tfrac{2}{6}\log\tfrac{2}{6} - \tfrac{2}{6}\log\tfrac{2}{6} = \log 3 \approx 1.585 \geq \log 2. \; \checkmark
\]

The final published table is thus (2, entropy-2)-diverse.

% ----------------------------------------------------------------------
\subsubsection{Practical Considerations}
\label{sec:practical_considerations}

\paragraph{Choosing $k$ and $\ell$.}
The choice of $k$ and $\ell$ depends on:
\begin{enumerate}
  \item \textbf{Sensitivity of data}: Medical records warrant $k \geq 10$ and entropy $\ell \geq 5$; less sensitive data may tolerate $k = 3$.
  \item \textbf{Dataset size}: Small datasets with few records per quasi-identifier profile may require aggressive suppression to achieve large $k$, severely reducing utility.
  \item \textbf{Regulatory requirements}: HIPAA's Safe Harbour rule implicitly targets $k = 20000$ by requiring ZIP codes to be reduced to 3 digits for small populations.
\end{enumerate}

\paragraph{Curse of Dimensionality.}
As $|QI|$ grows, equivalence classes become smaller and harder to fill, forcing larger generalisations or heavier suppression. This is the \emph{curse of dimensionality} in anonymisation. Practical mitigations include:
\begin{itemize}
  \item \textbf{Attribute selection}: Release only a subset of quasi-identifiers.
  \item \textbf{Microaggregation}: Cluster records and replace each cluster with its centroid.
  \item \textbf{Synthetic data}: Generate statistically representative synthetic records via generative models.
\end{itemize}

\paragraph{Sequential Releases and Composition.}
$k$-Anonymity and $\ell$-diversity do not compose well: releasing multiple $k$-anonymous tables derived from the same dataset can enable an adversary to intersect equivalence classes and re-identify individuals. Differential privacy provides tight composition bounds, making it preferable for sequential releases.

\subsection{t-Closeness (Distribution Protection)}
$l$-diversity can still leak information through \textbf{Skewness Attacks} if a group's distribution differs wildly from the global population. $t$-closeness requires that the distribution of a sensitive attribute in any group is close to the global distribution.

\begin{svgraybox}
\textbf{Technical Sidebar: Earth Mover's Distance in t-Closeness}\\
To quantify the ``closeness'' of distributions $P$ (global) and $Q$ (group), we use the \textbf{Earth Mover's Distance (EMD)}. For a numerical sensitive attribute, the distance $D[P, Q]$ is defined as:
\begin{equation}
D[P, Q] = \frac{1}{m-1} \sum_{i=1}^{m-1} \left| \sum_{j=1}^{i} p_j - \sum_{j=1}^{i} q_j \right|
\end{equation}
where $p_j$ and $q_j$ are the probability masses. An architect sets a threshold $t$ (e.g., 0.2) to limit the amount of statistical information a single group can leak.
\end{svgraybox}

\section{Attack Vectors on Classic Models}
\label{sec:attack_vectors_classic}

Even with $k, l,$ and $t$ protection, architects must defend against:
\begin{description}
  \item[The Mosaic Effect:] Using ``secondary releases'' or external datasets (e.g., voter registries) to perform \textbf{Linkage Attacks} that bridge the anonymized data back to real names.
  \item[Background Knowledge Attacks:] Exploiting an adversary's prior knowledge about a specific subject (e.g., ``I know my neighbor does not smoke; therefore, they must be the other record in this group'').
  \item[Structural Attacks (Graph Data):] Anonymizing social links and relationships requires defending against node-level and edge-level re-identification.
\end{description}

% ------------------------------------------------------------------
\subsection{Linkage Attack: Worked Example}
\label{subsec:linkage_worked_example}
% ------------------------------------------------------------------

The Mosaic Effect becomes concrete when traced through a real attack sequence. The following example follows Sweeney's canonical 1997 methodology \cite{sweeney2002} applied to a modern Vietnamese health dataset and a publicly available voter registration database.

\textbf{Setup.} A district hospital publishes a 3-anonymous medical dataset (Age generalized to 5-year bands, Gender retained, ZIP code generalized to district level). A voter registration database is publicly available containing Name, Age (exact), Gender, and full ZIP code.

\begin{table}[ht]
\centering
\caption{Linkage Attack: Three-Step Re-identification Trace. The 3-anonymous medical dataset (left) is joined to the voter registry (right) using the intersection of quasi-identifier values. In Step 3, the equivalence class collapses to a single record.}
\label{tab:linkage_attack}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{p{0.4cm} p{3.2cm} p{4.2cm} p{4.5cm}}
\hline\noalign{\smallskip}
\textbf{Step} & \textbf{Adversary Action} & \textbf{Medical Dataset State} & \textbf{Voter Registry Match} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
1 & Filter by gender & Rows: Gender = F, Age = [25--30), District = Hai Ba Trung & Voter registry: 847 female records in Hai Ba Trung \\[4pt]
2 & Narrow by exact age & Age band [25--30) in registry: voters aged 25, 26, 27, 28, 29 & 847 $\to$ 312 records within age band \\[4pt]
3 & Cross-reference known occupation & Adversary knows target works at ``Vinmec International'' (posted on LinkedIn). Voter registry includes employer field & \cellcolor{costhigh}312 $\to$ \textbf{1 record: Nguyen Thi H.} \\[4pt]
4 & Infer sensitive attribute & Nguyen Thi H. maps to equivalence class $q_1$ & Class $q_1$ contains: \{Diabetes, Diabetes, Hypertension\}. Re-identification probability: $2/3 = 67\%$ for Diabetes \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\textbf{Architectural lesson.} The medical dataset satisfied 3-anonymity against a quasi-identifier set of \{Age-band, Gender, District\}. The attack succeeded because the \emph{effective} quasi-identifier set---expanded by the voter registry---included Employer, reducing the anonymity set to 1. The defense: (1)~enumerate all publicly available datasets that could act as linkage auxiliaries before selecting quasi-identifiers; (2)~apply $t$-closeness or Differential Privacy when the adversary's auxiliary information cannot be bounded; (3)~restrict data recipients via contractual re-identification prohibition (Decree 356 Class B requirement, Table~\ref{tab:decree356_compliance}).

\section{Architectural Implementation}
\label{sec:arch_implementation_anonymization}

\subsection{Static vs. Dynamic Anonymization}
\textbf{SDM (Static Data Masking)} is used for creating one-time research datasets. \textbf{DDM (Dynamic Data Masking)} acts as ``Anonymization-as-a-Service,'' where data is generalized on-the-fly based on the requester's security role and the current $\delta$-presence probability.

\subsection{Referential Integrity and the Pareto Frontier}
In distributed architectures, architects must ensure \textbf{Referential Integrity}: if ``User\_A'' is masked to ``ID\_88'' in the Finance table, they must be masked to ``ID\_88'' in the Healthcare table to allow for relational joins without creating new linkage vulnerabilities.

Ultimately, the architect must navigate the \textbf{Pareto Frontier of Utility}: finding the optimal balance where Information Loss (IL) is minimized while Privacy Risk (PR) is kept below the acceptable threshold for the AI use case.

% ------------------------------------------------------------------
\subsubsection{The Privacy-Utility Trade-off Report Template}
\label{subsubsec:pareto_template}
% ------------------------------------------------------------------

Navigating the Pareto Frontier is not merely a technical optimization---it is a \textbf{communication exercise}. Business stakeholders and regulators require a structured, reproducible format for understanding the privacy-utility trade-off inherent in every parameter selection. The following \textbf{Privacy-Utility Trade-off Report} satisfies both audiences, meeting Decree 356's data-release documentation requirement.

\begin{table}[ht]
\centering
\caption{Privacy-Utility Trade-off Report Template: a standardized format for presenting mechanism parameter selections to business stakeholders and regulators under Decree 356 and the EU AI Act.}
\label{tab:pareto_report}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{p{3.8cm} p{7.5cm}}
\toprule
\textbf{Report Field} & \textbf{Content / Architect's Entry} \\
\midrule
Dataset Description & \textit{[Name; size $N$; sensitivity class A/B/C/D; domain]} \\
Privacy Mechanism & \textit{[k-Anonymity / $\ell$-Diversity / DP-SGD / Synthetic / LDP]} \\
Privacy Parameter(s) & \textit{[e.g., $k=10$; $\varepsilon=1.0$, $\delta=10^{-5}$]} \\
Utility Metric & \textit{[e.g., Model Accuracy, F1-score, Mean Absolute Error]} \\
Baseline Utility (no privacy) & \textit{[e.g., Accuracy = 94.2\%]} \\
Privacy-Adjusted Utility & \textit{[e.g., Accuracy = 89.7\% at $\varepsilon=1.0$]} \\
Utility Loss & \textit{[e.g., $\Delta = -4.5\%$; within $\pm5\%$ acceptable band]} \\
Regulatory Alignment & \textit{[Decree 356 Art.\ X; EU AI Act Art.\ Y; NIST RMF control Z]} \\
MIA-S Audit Score & \textit{[AUC = 0.54; below 0.60 threshold --- PASS]} \\
Approved By & \textit{[Data Protection Officer + date]} \\
\bottomrule
\end{tabular}
\end{table}

Pair this report with a \textbf{Privacy-Utility Curve} so decision-makers can select the operating point that satisfies both the technical requirement and the business constraint.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=10cm, height=5.5cm,
    xlabel={Privacy Budget $\varepsilon$},
    ylabel={Model Accuracy (\%)},
    xmin=0.1, xmax=10,
    ymin=70, ymax=97,
    xtick={0.1,0.5,1,2,4,8,10},
    xticklabels={0.1,0.5,1.0,2.0,4.0,8.0,10},
    grid=major, grid style={dashed, gray!30},
    legend pos=south east, font=\small,
]
\addplot[thick, blue!70!black, mark=*] coordinates {
    (0.1,72)(0.5,80)(1.0,85)(2.0,89)(4.0,92)(8.0,94)(10,95)};
\addlegendentry{DP-SGD Accuracy}
\addplot[thick, red!60!black, dashed] coordinates {(0.1,94.2)(10,94.2)};
\addlegendentry{Baseline (no DP)}
\addplot[only marks, mark=triangle*, mark size=4pt, orange!90!black]
    coordinates {(1.0,85)};
\addlegendentry{Decree 356 Operating Point ($\varepsilon{=}1.0$)}
\addplot[thick, green!55!black, dotted] coordinates {(0.1,89.5)(10,89.5)};
\addlegendentry{Min.\ acceptable utility threshold}
\end{axis}
\end{tikzpicture}
\caption{Privacy-Utility Curve for a diagnostic classifier: Accuracy vs.\ $\varepsilon$. The Decree 356 operating point ($\varepsilon=1.0$) delivers 85\% accuracy, above the minimum threshold. The full Trade-off Report (Table~\ref{tab:pareto_report}) accompanies this visualization for regulatory submission.}
\label{fig:pareto_curve}
\end{figure}

\section{The Vietnam Edge: Decree 356 (2026) Standards}
\label{sec:vietnam_decree_356}

Effective March 2026, \textbf{Decree 356} (replacing Decree 13) defines ``Sensitive Personal Data'' to include transaction history and real-time behavioral tracking. To comply with ``Advanced Safeguards,'' architects must map their $k$-anonymity levels to specific sensitivity tiers before data can be transferred to third parties or across borders.

% ------------------------------------------------------------------
\subsection{Decree 356 Sensitivity-Tier Compliance Table}
\label{subsec:decree356_compliance}
% ------------------------------------------------------------------

Table~\ref{tab:decree356_compliance} operationalizes Decree 356 for anonymization engineers: it maps each sensitivity class to the \emph{minimum} required parameter values, the mandatory additional safeguard, and the relevant article. Every data release from a Vietnamese data controller must be evaluated against this table before a cross-border transfer or third-party data-sharing agreement can be executed.

\begin{table}[ht]
\centering
\caption{Decree 356 Sensitivity-Tier Compliance Requirements for Anonymized Data Release. Sensitivity classes follow the taxonomy of Chapter~\ref{chap:privacy_attacks}. All minimum values are floors, not targets; architects should select the strongest mechanism compatible with required utility.}
\label{tab:decree356_compliance}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{p{1.5cm} p{2.5cm} p{1.4cm} p{1.4cm} p{1.4cm} p{3.8cm}}
\hline\noalign{\smallskip}
\textbf{Class} & \textbf{Examples} & \textbf{Min.\ $k$} & \textbf{Min.\ $\ell$} & \textbf{Max.\ $\varepsilon$} & \textbf{Mandatory Safeguard (Decree 356)} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
\cellcolor{costhigh}A & Biometric, health, political affiliation, EEG & \cellcolor{costhigh}20 & \cellcolor{costhigh}5 & \cellcolor{costhigh}1.0 & Art.~9: Formal Privacy Impact Assessment; DPO sign-off; DP-SGD mandatory for ML \\[4pt]
\cellcolor{costhigh}B & Financial records, precise location history, legal status & \cellcolor{costhigh}15 & \cellcolor{costhigh}4 & \cellcolor{costhigh}2.0 & Art.~9: Data-sharing agreement with contractual re-identification prohibition \\[4pt]
\cellcolor{costmed}C & Behavioral preferences, purchase history, device IDs & \cellcolor{costmed}10 & \cellcolor{costmed}3 & \cellcolor{costmed}4.0 & Art.~6: Purpose limitation documented in processing register \\[4pt]
\cellcolor{costlow}D & Aggregated public statistics, anonymized survey responses & \cellcolor{costlow}5 & \cellcolor{costlow}2 & \cellcolor{costlow}8.0 & Art.~6: Standard data-transfer agreement sufficient \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\begin{svgraybox}
\textbf{Compliance Workflow:} Before any release, (1)~classify the dataset using the sensitivity taxonomy; (2)~verify that all anonymization parameters meet the class minimum; (3)~run the MIA audit (AUC $< 0.60$) and the NNDR check (NNDR $> 0.50$); (4)~attach the Privacy-Utility Trade-off Report (Table~\ref{tab:pareto_report}) and the Decree 356 class determination to the data-sharing agreement. For Class A/B data, the Data Protection Officer must countersign before transfer.
\end{svgraybox}

Practitioners should view $k$-anonymity and $\ell$-diversity not as competing but as complementary layers: choose $k$ to limit re-identification risk, $\ell$ to prevent attribute inference, and consider $t$-closeness or differential privacy when adversaries may have rich auxiliary information or when data will be released repeatedly.

% ==================================================================
\section{Synthetic Data: Promise, Peril, and the
         Privacy-Utility-Fidelity Trilemma}
\label{sec:synthetic-data}
% ==================================================================

\begin{quote}
\textit{``Synthetic data is not anonymized data. It is a statistical
shadow of the original---and shadows can still reveal the shape of
what cast them.''}
\end{quote}

% ------------------------------------------------------------------
\subsection{The Rise of Synthetic Data as a Privacy Primitive}
\label{subsec:synth-rise}
% ------------------------------------------------------------------

The limitations of classical anonymization---the curse of
dimensionality, the Mosaic Effect, and the brittleness of
$k$-anonymity under auxiliary information (Section~\ref{sec:attack_vectors_classic})---have
driven practitioners toward a fundamentally different paradigm:
\textbf{synthetic data generation}.
Rather than transforming or suppressing real records, this approach
uses a generative model to produce an entirely new dataset whose
statistical properties approximate those of the original, without
containing any real individual's record.

The appeal is immediate.
A hospital system can share a synthetic patient dataset with external
researchers without ever exposing a single real patient's history.
A bank can train a fraud-detection model on synthetic transactions
that mirror real fraud patterns without violating data-localization
laws.
As noted in Section~\ref{sec:practical_considerations}, synthetic
data also directly addresses the curse of dimensionality that plagues
classical methods.

The primary generative architectures used in practice are:

\begin{description}
  \item[Generative Adversarial Networks (GANs).]
    A \emph{Generator} network produces synthetic records; a
    \emph{Discriminator} network attempts to distinguish them from real
    ones.
    The adversarial dynamic drives the Generator toward high statistical
    fidelity.
    Variants such as CTGAN and TableGAN are specialized for tabular
    data.

  \item[Variational Autoencoders (VAEs).]
    A probabilistic encoder maps real records into a latent
    distribution; the decoder samples from this distribution to produce
    synthetic records, providing a smoother and more controlled
    generative process.

  \item[Diffusion Models.]
    Increasingly applied to structured data, these models learn to
    denoise records from pure noise back to realistic samples, offering
    strong fidelity for complex distributions.
\end{description}

For structured tabular data---the dominant format in healthcare,
finance, and government---libraries such as \texttt{SDV} (Synthetic
Data Vault) and Gretel.ai provide production-ready pipelines for
fitting and sampling from these generative models.

% ------------------------------------------------------------------
\subsection{The Privacy-Utility-Fidelity Trilemma}
\label{subsec:puf-trilemma}
% ------------------------------------------------------------------

A fundamental architectural tension governs every synthetic data
pipeline.
The \textbf{Privacy-Utility-Fidelity (PUF) Trilemma} states that
these three properties are in perpetual tension: improving any one of
them typically degrades at least one of the others.

\begin{description}
  \item[Fidelity] measures how closely the synthetic data's statistical
    distribution matches the original.
    High fidelity means the generative model has learned the real
    data's correlations, marginals, and joint distributions accurately.

  \item[Utility] measures how well models trained on synthetic data
    perform when deployed on real data.
    A high-fidelity dataset is often---but not always---a high-utility
    dataset, as downstream model performance depends on which specific
    features and correlations were preserved.

  \item[Privacy] measures the degree to which synthetic records resist
    re-identification and attribute-inference attacks against the
    original training population.
\end{description}

The trilemma manifests as follows.
A generative model trained to maximize fidelity will tend to memorize
rare or outlier records, since these are hardest to model
statistically and the model is penalized most for missing them.
This memorization is the primary source of privacy failure.
Conversely, aggressively regularizing the generative model to prevent
memorization (e.g., through Differential Privacy, discussed in
Chapter~\ref{chap:differential_privacy}) introduces noise that degrades both fidelity
and utility.
The architect must navigate this frontier deliberately, rather than
assuming that generating synthetic data is inherently private.

% --- TikZ diagram -----------------------------------------------
\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    vertex/.style={circle, draw, minimum size=2.2cm,
                   font=\small\bfseries, align=center},
    edge/.style={thick, <->, >=Stealth}
  ]
  % Three vertices of the trilemma triangle
  \node[vertex] (P) at (90:3.2cm)  {Privacy};
  \node[vertex] (F) at (210:3.2cm) {Fidelity};
  \node[vertex] (U) at (330:3.2cm) {Utility};

  % Edges with tension labels
  \draw[edge] (P) -- (F)
    node[midway, left,  font=\footnotesize\itshape, xshift=-4pt]
         {DP noise reduces\ fidelity};
  \draw[edge] (F) -- (U)
    node[midway, below, font=\footnotesize\itshape, yshift=-6pt]
         {high fidelity\ aids utility};
  \draw[edge] (U) -- (P)
    node[midway, right, font=\footnotesize\itshape, xshift=4pt]
         {memorization\ leaks privacy};

  % Optimal design point in the centre
  \node[draw, dashed, rounded corners, font=\footnotesize,
        fill=white, align=center] at (0,0)
       {Optimal\\Design\\Point};
\end{tikzpicture}
\caption{The Privacy-Utility-Fidelity (PUF) Trilemma.
  No synthetic data system can simultaneously maximize all three
  properties.
  The architect's task is to identify the optimal operating point for
  the specific threat model and downstream use case.}
\label{fig:puf-trilemma}
\end{figure}
% ----------------------------------------------------------------

% ------------------------------------------------------------------
\subsection{The Core Threat: Membership Inference on Synthetic
            Datasets}
\label{subsec:mia-synthetic}
% ------------------------------------------------------------------

The most dangerous misconception in synthetic data deployment is the
belief that, because no real record appears verbatim in the synthetic
dataset, the privacy of the original training population is
guaranteed.
This is false.

A \textbf{Membership Inference Attack against synthetic data}
(MIA-S) asks a different question from the classical MIA described in
Chapter~\ref{chap:classic_anonymization}.
Rather than querying the generative model directly, the adversary is
given access only to the published synthetic dataset~$\tilde{T}$ and
attempts to determine whether a specific real record~$r$ was part of
the original training data~$T$ used to fit the generative model.

The attack proceeds in three steps:

\begin{enumerate}
  \item \textbf{Shadow Generative Models.}
    The adversary trains multiple ``shadow'' generative models on
    datasets that either include or exclude a known target record~$r$.

  \item \textbf{Statistical Fingerprinting.}
    The adversary computes statistical proximity metrics
    (e.g., nearest-neighbour distances, density estimates) between~$r$
    and the published synthetic dataset~$\tilde{T}$.

  \item \textbf{Threshold Classification.}
    Records from the original training set tend to have synthetic
    ``neighbours'' that are statistically closer than records that
    were never in~$T$.
    An attack classifier learns this threshold.
\end{enumerate}

Formally, let
\begin{equation}
  d\!\left(r,\tilde{T}\right)
    = \min_{\tilde{r}\,\in\,\tilde{T}} \|r - \tilde{r}\|
\end{equation}
be the distance from real record~$r$ to its nearest synthetic
neighbour.
Empirical research demonstrates that this distance is systematically
smaller for records $r \in T$ than for $r \notin T$, revealing
membership with attack AUCs exceeding $0.80$ on medical datasets
trained without privacy constraints---far above the random baseline
of $0.50$.

\paragraph{The Mechanism: Why Generative Models Leak.}
Generative models leak membership information for the same reason all
machine-learning models do: they \emph{overfit}.
A GAN that achieves near-perfect fidelity has, in effect, memorized
the statistical structure of every record in the training set,
including rare individuals who occupy sparse regions of the feature
space.
When the synthetic data is released, these memorized patterns remain
embedded in the distribution of~$\tilde{T}$.

Outlier records---patients with rare disease combinations, customers
with unusual transaction patterns---are the most vulnerable.
The generative model disproportionately ``remembers'' them, because
they represent the hardest part of the distribution to model
correctly, and the synthetic dataset consequently produces near-copies
of these individuals.

% ------------------------------------------------------------------
\subsection{Evaluation Metrics: Measuring the Trilemma}
\label{subsec:synth-metrics}
% ------------------------------------------------------------------

The architect requires quantitative tools to navigate the trilemma.
The following metrics form a standard evaluation suite.

\paragraph{Fidelity Metrics.}
\begin{itemize}
  \item \textbf{Marginal Distribution Distance.}
    Column-wise comparison of real vs.\ synthetic distributions using
    Jensen--Shannon Divergence or Wasserstein distance.
  \item \textbf{Correlation Matrix Difference.}
    The Frobenius norm of the difference between real and synthetic
    pairwise correlation matrices, capturing whether inter-feature
    relationships are preserved.
\end{itemize}

\paragraph{Utility Metrics.}
\begin{itemize}
  \item \textbf{Train on Synthetic, Test on Real (TSTR).}
    Train a downstream classifier on~$\tilde{T}$ and evaluate its
    accuracy on a held-out real test set.
    The gap between TSTR accuracy and Train on Real, Test on Real
    (TRTR) accuracy quantifies the utility loss.
\end{itemize}

\paragraph{Privacy Metrics.}
\begin{itemize}
  \item \textbf{Nearest-Neighbour Distance Ratio (NNDR).}
    For each real record $r \in T$, compute
    \begin{equation}
      \mathrm{NNDR}(r)
        = \frac{d(r,\,\tilde{T})}{d(r,\;T \setminus \{r\})}.
    \end{equation}
    A low NNDR indicates the synthetic data contains near-copies of
    real records---a privacy failure.

  \item \textbf{Membership Inference Attack AUC (MIA-S AUC).}
    Empirically evaluate a shadow-model MIA against the published
    synthetic dataset.
    An AUC significantly above $0.5$ indicates measurable privacy
    leakage.
\end{itemize}

Table~\ref{tab:synth-thresholds} summarizes recommended thresholds
for production deployment.

\begin{table}[ht]
\centering
\caption{Recommended evaluation thresholds for synthetic data
         release decisions.}
\label{tab:synth-thresholds}
\begin{tabular}{p{4cm}p{3.5cm}p{4.5cm}}
\hline\noalign{\smallskip}
\textbf{Metric} &
\textbf{Acceptable Threshold} &
\textbf{Action if Violated} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Marginal JS Divergence
  & $<\!0.15$ per column
  & Increase model capacity or training epochs \\[4pt]
TSTR\,/\,TRTR Accuracy Ratio
  & $>\!0.90$
  & Retune generative model \\[4pt]
NNDR (5th percentile)
  & $>\!0.50$
  & Apply DP-SGD to generative model training \\[4pt]
MIA-S AUC
  & $<\!0.60$
  & Apply DP or increase regularization \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

% ------------------------------------------------------------------
\subsection{Architectural Defences: Bridging to Differential
            Privacy}
\label{subsec:synth-dp-bridge}
% ------------------------------------------------------------------

The principled solution to membership inference leakage in synthetic
data is to apply \textbf{Differentially Private training to the
generative model itself}.
Rather than post-hoc filtering of the synthetic output, DP is applied
during the training of the GAN or VAE using the DP-SGD algorithm
(formalized in Chapter~\ref{chap:differential_privacy}).

A DP-trained generative model provides a formal guarantee: the
distribution of the synthetic dataset~$\tilde{T}$ is
$(\varepsilon, \delta)$-indistinguishable regardless of whether any
single record~$r$ was present in~$T$ or not.
This is precisely the Differential Privacy guarantee of
Section~\ref{sec:dp_formal_definition}, now applied to the generative
process rather than to a query function.

The practical workflow for a privacy-preserving synthetic data
pipeline is shown in Listing~\ref{lst:dp-synthetic}.

\begin{lstlisting}[language=Python, style=pythonstyle,
  caption={Conceptual DP-Synthetic Data Pipeline using CTGAN and
           Opacus-style DP-SGD.},
  label={lst:dp-synthetic}
]
# Step 1: Load and preprocess real sensitive data
real_data = load_sensitive_dataset()

# Step 2: Initialise generative model with DP-SGD optimizer.
# The epsilon budget must be pre-agreed with the
# Data Protection Officer.
generator = CTGAN(epochs=300)
privacy_engine = DPSGDEngine(
    epsilon=1.0,
    delta=1e-5,
    max_grad_norm=1.0
)
privacy_engine.attach(generator.discriminator_optimizer)

# Step 3: Train under DP constraints.
# DP-SGD clips and noises gradients internally.
generator.fit(real_data)

# Step 4: Evaluate the PUF trilemma before release.
synthetic_data = generator.sample(n=len(real_data))
evaluate_fidelity(real_data, synthetic_data)
evaluate_utility(real_data, synthetic_data)
run_mia_audit(real_data, synthetic_data)  # Must pass before publication
\end{lstlisting}

The critical caveat is that DP training of generative models incurs a
significant fidelity cost, particularly for high-dimensional tabular
data with many sensitive columns.
The architect must treat the privacy budget~$\varepsilon$ as a
first-class design parameter, negotiating its value against the
downstream utility requirements \emph{before} training begins.

% ------------------------------------------------------------------
\subsection{Synthetic Data and Vietnam's Decree 356 (2026)}
\label{subsec:synth-decree356}
% ------------------------------------------------------------------

Under Decree 356 (Section~\ref{sec:vietnam_decree_356}), synthetic
data occupies an ambiguous legal position.
The Decree does not explicitly exempt synthetic datasets from its
data-transfer restrictions, and the Vietnamese Data Protection
Authority has taken the position that synthetic data \emph{derived
from Sensitive Personal Data} retains its sensitive classification
unless the data controller can demonstrate---through a formal Privacy
Impact Assessment---that the re-identification risk falls below the
prescribed threshold.

In practice, a DP-trained generative model with a formally audited
$\varepsilon \leq 1.0$ and a documented NNDR audit report provides
the strongest basis for demonstrating compliance when sharing
synthetic health or financial data across organizational boundaries.
Data controllers should maintain the generative model's training
provenance and the full evaluation
suite results as part of their compliance documentation.

% ------------------------------------------------------------------
% ------------------------------------------------------------------
\subsection{The Synthetic Verification Gateway}
\label{subsec:synth-verification-gateway}
% ------------------------------------------------------------------

Even a DP-trained generative model should not be released without a final automated audit. The \textbf{Synthetic Verification Gateway} is a middleware architectural pattern that intercepts every synthetic dataset release request, executes an automated MIA audit, and either issues an \textbf{Audit Certificate} (if the MIA-S AUC is below threshold) or \textbf{blocks the release} with a remediation report.

The threshold is $\text{AUC}_{\text{MIA}} < 0.60$ (Sovereign AI Scorecard, Chapter~\ref{chap:attm}): a score at or above this level indicates membership leakage distinguishable from chance at a practically significant level.

\begin{lstlisting}[language=Python, style=pythonstyle,
  caption={Synthetic Verification Gateway: automated MIA-S audit middleware. The certificate hash provides an immutable, verifiable link between the published dataset and its privacy audit---satisfying Decree 356 documentation requirements.},
  label={lst:synth-gateway}]
# Synthetic Verification Gateway
# Run before any synthetic dataset release.
from privacymeter import MembershipInferenceAudit
import hashlib, json

AUDIT_THRESHOLD = 0.60  # Sovereign AI Scorecard threshold

def synthetic_verification_gateway(real_train, real_holdout, synthetic):
    auditor = MembershipInferenceAudit()
    auc = auditor.run(real_train=real_train,
                      real_holdout=real_holdout,
                      synthetic=synthetic)
    if auc >= AUDIT_THRESHOLD:
        raise ReleaseBlockedError(
            f"MIA-S AUC={auc:.3f} exceeds threshold {AUDIT_THRESHOLD}. "
            "Apply DP-SGD to the generative model and re-generate."
        )
    cert = {"status": "APPROVED", "mia_auc": round(auc, 4),
            "threshold": AUDIT_THRESHOLD}
    cert["certificate_id"] = hashlib.sha256(
        json.dumps(cert).encode()).hexdigest()
    return cert
\end{lstlisting}

\subsection{Conclusion: Synthetic Data as a Complement, Not a
            Substitute}
\label{subsec:synth-conclusion}
% ------------------------------------------------------------------

Synthetic data is a powerful addition to the privacy-preserving
architect's toolkit, but it is not a replacement for the mathematical
guarantees of Differential Privacy.
It is best understood as a \textbf{complementary layer}: one that
trades the structural rigidity of $k$-anonymity for statistical
realism, at the cost of introducing new leakage vectors that require
active mitigation.

The key architectural principles are:

\begin{itemize}
  \item Always evaluate synthetic datasets against a formal MIA audit
    before publication---never assume that generation implies privacy.
  \item Apply DP-SGD to the generative model when the underlying data
    is sensitive, accepting the fidelity cost as the price of a formal
    guarantee.
  \item Report the full Privacy-Utility-Fidelity trilemma metrics
    alongside the synthetic dataset as part of the data release
    documentation.
\end{itemize}

With these principles in place, synthetic data can responsibly serve
as the training substrate for models that must never touch real PII.
In the next chapter, we formalize the mathematical framework that
underpins both DP-trained generative models and the broader class of
privacy-preserving mechanisms: \textbf{Differential Privacy}, the
gold standard that transforms these empirical defences into provable
guarantees.

% ------------------------------------------------------------------
\section{Privacy Mechanism Selection: A Diagnostic Decision Tree}
\label{sec:privacy_mechanism_selection}
% ------------------------------------------------------------------

Having established the technical characteristics of k-Anonymity, $\ell$-Diversity, $t$-Closeness, Differential Privacy, and Synthetic Data, the practitioner faces the most common real-world challenge: \emph{which primitive do I use for this specific deployment?} The following \textbf{Privacy Mechanism Selection Decision Tree} provides a diagnostic flowchart based on six discriminating questions.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    node distance=1.1cm and 2.6cm,
    font=\small,
    decision/.style={draw, diamond, aspect=3.2, inner sep=3pt, text centered,
                     font=\bfseries\scriptsize, fill=yellow!8, draw=orange!70!black},
    mech/.style={draw, rounded corners, inner sep=6pt, text centered,
                 font=\bfseries\scriptsize, minimum width=3.6cm},
    kanon/.style={mech, fill=blue!10,   draw=blue!70!black},
    dpbox/.style={mech, fill=green!10,  draw=green!60!black},
    ldpbox/.style={mech, fill=orange!10, draw=orange!70!black},
    synth/.style={mech, fill=purple!10, draw=purple!70!black},
    tclose/.style={mech, fill=red!10,   draw=red!60!black},
    arrow/.style={->, >=stealth, thick, draw=gray!70}
]

\node (d1) [decision, text width=4.5cm]
    {Raw data accessible\\to the consumer?};
\node (d2) [decision, below=1.4cm of d1, text width=4.5cm]
    {$N < 5{,}000$ or tabular\\with few attributes?};
\node (d3) [decision, below=1.4cm of d2, text width=4.5cm]
    {Mobile-first / server\\cannot be trusted?};
\node (d4) [decision, below=1.4cm of d3, text width=4.5cm]
    {Genomic / high-dim with\\rich auxiliary knowledge?};
\node (d5) [decision, below=1.4cm of d4, text width=4.5cm]
    {Repeated or sequential\\query releases?};

\node (synth_n) [synth,  right=2.8cm of d1, text width=3.8cm]
    {\textbf{Synthetic Data}\\DP-trained GAN/VAE\\(Sec.~\ref{sec:synthetic-data})};
\node (kanon_n) [kanon,  right=2.8cm of d2, text width=3.8cm]
    {\textbf{$k$-Anonymity}\\+ $\ell$-Diversity\\(Sec.~\ref{sec:k_l_t_hierarchy})};
\node (ldp_n)   [ldpbox, right=2.8cm of d3, text width=3.8cm]
    {\textbf{Local DP (LDP)}\\Randomized Response\\on-device noise (Ch.~5)};
\node (tclose_n)[tclose, right=2.8cm of d4, text width=3.8cm]
    {\textbf{$t$-Closeness + DP}\\Distribution protection\\+ formal guarantee};
\node (dp_n)    [dpbox,  right=2.8cm of d5, text width=3.8cm]
    {\textbf{DP-SGD (Central)}\\R\'{e}nyi accounting\\Opacus / TF Privacy (Ch.~5)};
\node (default_n)[dpbox, below=1.4cm of d5, text width=3.8cm]
    {\textbf{DP-SGD (Default)}\\$\varepsilon \leq 1.0$; Gaussian mech.\\Conservative baseline};

\draw [arrow] (d1.east)  -- node[above, font=\scriptsize] {No}  (synth_n.west);
\draw [arrow] (d1.south) -- node[right, font=\scriptsize] {Yes} (d2.north);
\draw [arrow] (d2.east)  -- node[above, font=\scriptsize] {Yes} (kanon_n.west);
\draw [arrow] (d2.south) -- node[right, font=\scriptsize] {No}  (d3.north);
\draw [arrow] (d3.east)  -- node[above, font=\scriptsize] {Yes} (ldp_n.west);
\draw [arrow] (d3.south) -- node[right, font=\scriptsize] {No}  (d4.north);
\draw [arrow] (d4.east)  -- node[above, font=\scriptsize] {Yes} (tclose_n.west);
\draw [arrow] (d4.south) -- node[right, font=\scriptsize] {No}  (d5.north);
\draw [arrow] (d5.east)  -- node[above, font=\scriptsize] {Yes} (dp_n.west);
\draw [arrow] (d5.south) -- node[right, font=\scriptsize] {No}  (default_n.north);

\end{tikzpicture}
\caption{Privacy Mechanism Selection Decision Tree: six diagnostic questions guide the architect to the appropriate privacy primitive for their specific deployment scenario. Each terminal references the implementing section or chapter.}
\label{fig:privacy_selection_tree}
\end{figure}

\begin{svgraybox}
\textbf{Vietnam Decree 356 Guidance (2026):} For ``Sensitive Personal Data'' (health, biometrics, financial behavior), Decree 356 recommends $\varepsilon \leq 1.0$ with Gaussian mechanism or a DP-trained generative model. $k$-Anonymity with $k \geq 10$ is accepted for ``General Personal Data.'' Local DP is explicitly endorsed for mobile telemetry collection from consumer devices under Article 18.
\end{svgraybox}

% ------------------------------------------------------------------
\section{From the Trenches: Practitioner's Challenge}
\label{sec:ch4_challenge}
% ------------------------------------------------------------------

\begin{tcolorbox}[challengebox, title={Scenario: The Cross-Border Health Audit}]
\textbf{The Context:} 
A European research hospital wants to collaborate with your ASEAN-based health-tech startup. They require access to a 50,000-record dataset of patient triage outcomes to train their LLM diagnostic assistant. 

\textbf{The Architect's Dilemma:} 
The legal team demands that the data \"cannot be linked to any individual.\" Your lead data scientist suggests achieving $k=10$ anonymity and suppressing Zip codes. Your compliance officer suggests using an off-the-shelf LLM to generate a synthetic sibling dataset. Which mechanism do you deploy, and how do you justify it to the C-suite and the EU regulators?
\end{tcolorbox}

\begin{tcolorbox}[scorecardbox, title={Sovereign AI Scorecard Evaluation}]
\textbf{Option A: $k$-Anonymity ($k=10$)}
\begin{itemize}
    \item \textit{Cost:} Low (O(nd log n) complexity).
    \item \textit{Utility:} Moderate (coarsens data features).
    \item \textit{Regulatory Compliance:} \textbf{Fail.} High-dimensional medical datasets routinely suffer from the Curse of Dimensionality. European regulators consider $k$-anonymity insufficient against modern background-knowledge and Linkage attacks (as seen in the Netflix Prize breach).
\end{itemize}

\textbf{Option B: DP-Synthetic Data ($\varepsilon = 1.0$)}
\begin{itemize}
    \item \textit{Cost:} High (requires DP-SGD training on generative models).
    \item \textit{Utility:} Good (retains statistical properties without 1-to-1 matching).
    \item \textit{Regulatory Compliance:} \textbf{Pass.} A DP-trained generative model provides formal mathematical deniability against Membership Inference Attacks, satisfying GDPR's strict anonymisation exemption while cleanly passing through ASEAN data localization checks.
\end{itemize}

\textbf{The Verdict: Option B.} 
At 50,000 records, the dataset is large enough for DP-SGD to converge with reasonable fidelity. The computational overhead of DP-Synthetic Data is justified by the outright elimination of regulatory cross-border transfer risk.
\end{tcolorbox}
