%%%%%%%%%%%%%%%%%%%%% Chapter 3 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Socio-Technical Trust and Human Agency
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\motto{Trust is not a mathematical proof; it is a shared expectation between human and machine. In the agentic era, accountability requires a clear hand on the steering wheel.}

\chapter{Socio-Technical Trust and Human Agency}
\label{chap:hitl}

\abstract{True trustworthiness exists at the intersection of technical reliability and human agency. By 2026, the industry has realized that a system providing perfectly private answers remains a failure if it lacks "Social Legibility" and a clear hierarchy of control. This chapter explores the evolution of Socio-Technical Trust, specifically the regulatory and architectural shift toward Human-on-the-Loop (HOTL) and Human-in-Command (HiC) models. We analyze the technical implementation of "The Emergency Brake," explore the move toward rigorous Conformal Prediction for uncertainty quantification, and provide a 2026 case study on autonomous healthcare triage.}

\section{The Interdependency of Trust}
\label{sec:interdependency_trust}

In 2026, the ``Trust Gap'' has become the primary bottleneck for AI adoption. This gap is the distance between what an autonomous agent \textit{can} do and what a human operator \textit{understands} it is doing. We have moved beyond the ``Black Box'' era into one where \textbf{Social Legibility} is a technical requirement.

Trust is highest when two layers are harmonized:
\begin{description}
  \item[Technical Trust:] The model's inherent stability, resilience, and mathematical privacy guarantees (e.g., $\epsilon$-DP, robust recovery).
  \item[Social Trust:] The presence of ``Institutional Backstops''---regulations, audit trails, and the assurance that a human or organization remains responsible for the agent's actions.
\end{description}

\section{HITL, HOTL, and Human-in-Command (HiC)}
\label{sec:agency_spectrum}

The 2026 regulatory environment, specifically Vietnam's AI Law and the updated NIST framework, defines a spectrum of human agency that architects must implement:

\begin{description}
  \item[Human-in-the-Loop (HITL):] Active human participation in every decision (e.g., a doctor approving every individual prescription from an AI).
  \item[Human-on-the-Loop (HOTL):] The agent acts autonomously, but a human monitors the stream and can intervene or override in real-time.
  \item[Human-in-Command (HiC):] A high-level governing requirement where humans set the boundaries, constitutional constraints, and ``Emergency Brakes'' within which the agent operates.
\end{description}

Architecturally, the \textbf{Emergency Brake Pattern} is a hardcoded circuit-breaker that immediately suspends agent execution if a core safety or privacy boundary (defined in the Norm Matrix from Chapter 2) is approached.

\subsection{The Post-Brake Recovery Protocol}
\label{subsec:post-brake-recovery}

When the Emergency Brake trips, the agent enters a suspended ``Grey'' state within the Confidence Spectrum. A critical failure in many early autonomous architectures was neglecting the restart mechanism: simply un-pausing the agent often resulted in an immediate re-triggering of the brake, trapping the system in an infinite failure loop.

To resolve this, the Architect must implement a rigorous \textbf{Human-in-Command (HiC) Transition Protocol} consisting of three phases:

\begin{enumerate}
    \item \textbf{Context Freezing and State Dump:} The precise memory state, active tool-call queue, and the specific sequence of LLM generation steps that triggered the brake are cryptographically signed and frozen in the ZK-Audit Log. The agent is entirely sequestered from new inputs.
    \item \textbf{Human Deliberation Interface:} The Trust Dashboard filters the ``AI Noise'' and presents the frozen state to a human supervisor. It explicitly highlights the specific Datalog predicate or constitutional boundary that the agent attempted to violate, providing deterministic context for the suspension.
    \item \textbf{State Mutation and Safe Restart:} Crucially, the human supervisor must not merely click ``Approve'' or ``Deny.'' The recovery protocol requires the operator to mutate the agent's state before re-engaging the run loop. This may involve injecting a corrective system prompt into the context window, forcibly clearing the tool queue, or dynamically revoking access to the offending API. This state mutation ensures the agent resumes with altered parameters, effectively breaking the causal loop that caused the initial suspension.
\end{enumerate}

\section{Calibrated Confidence \& Conformal Prediction}
\label{sec:uncertainty_handover}

A trustworthy agent must not only be accurate but also ``self-aware.'' We have moved past simple softmax probabilities, which were prone to ``Confident Hallucinations,'' toward \textbf{Conformal Prediction} \cite{ouyang2022}. By integrating human feedback directly into the model's behavioral policy through techniques like \textbf{Reinforcement Learning from Human Feedback (RLHF)} \cite{ouyang2022}, we create a system that is fundamentally aligned with human agency.

Conformal Prediction provides a mathematical framework for generating prediction sets that are guaranteed to contain the true value with a pre-specified probability (e.g., $95\%$). When the prediction set grows too large (indicating high ambiguity), the agent initiates an \textbf{Uncertainty Handover}, passing the decision to a human expert.

\section{Socio-Technical Oversight Architecture}
\label{sec:oversight_arch}

Designing for oversight requires addressing the ``Multi-Stakes Pattern,'' where the interests of the Subject, the Architect, and the Auditor must be balanced. Key components include:
\begin{itemize}
    \item \textbf{Verifiable Audit Trails:} Using immutable ledgers or Zero-Knowledge logs to record agent actions without exposing raw sensitive training data.
    \item \textbf{Cognitive Ergonomics:} Reducing ``AI Noise'' by filtering agent logs into a human-readable \textbf{Trust Dashboard} that highlights only the most critical decision-branches for review.
\end{itemize}

% ------------------------------------------------------------------
%  Colour palette for UI state indicators
% ------------------------------------------------------------------
\definecolor{stategreen}{RGB}{  56, 168,  80}
\definecolor{stateamber}{RGB}{ 255, 180,   0}
\definecolor{statered}{RGB}{   220,  53,  53}
\definecolor{stategray}{RGB}{  130, 130, 130}
\definecolor{uibg}{RGB}{       245, 247, 250}
\definecolor{patternbg}{RGB}{  235, 243, 255}

% ------------------------------------------------------------------
%  tcolorbox style for design-pattern callouts
% ------------------------------------------------------------------
\tcbset{
  patternbox/.style={
    enhanced,
    colback=patternbg,
    colframe=blue!55!black,
    fonttitle=\bfseries\small,
    title={Design Pattern},
    left=6pt, right=6pt, top=4pt, bottom=4pt,
    breakable
  }
}

% ==================================================================
\subsection{Trust Dashboards: Design Patterns for Uncertainty
            Handover Interfaces}
\label{subsec:trust-dashboards}
% ==================================================================

\begin{quote}
\textit{``A model that knows it is uncertain but cannot communicate
that uncertainty to a human operator has not solved the trust
problem---it has hidden it behind a screen.''}
\end{quote}

The calibrated confidence framework of
Section~\ref{sec:uncertainty_handover} established the
mathematical trigger for an Uncertainty Handover: when the
Conformal Prediction set $\mathcal{C}(x)$ grows beyond a
pre-specified cardinality threshold $\kappa$, the agent must
escalate to a human expert.
What the mathematical framework does not specify is the equally
critical question of \emph{how} that escalation is communicated.

This is not a peripheral UX concern.
The empirical literature on clinical decision support consistently
demonstrates two failure modes that destroy the value of any
technically correct handover signal:

\begin{description}
  \item[Automation Bias.]
    Human operators who interact with AI systems over extended
    periods systematically over-trust the system's outputs,
    accepting recommendations without critical evaluation even
    when the system explicitly flags uncertainty.
    The Trust Dashboard must be designed to \emph{interrupt}
    this bias, not accommodate it.

  \item[Alarm Fatigue.]
    When uncertainty alerts are frequent, visually undifferentiated,
    or cognitively demanding to interpret, operators habituate to
    them and begin ignoring them.
    In the healthcare context, alarm fatigue is directly linked
    to adverse patient outcomes.
    The Trust Dashboard must be designed to be \emph{rare,
    salient, and immediately actionable}.
\end{description}

This subsection provides five evidence-based design patterns for
Uncertainty Handover interfaces, grounded in the Human Factors
and Human-Computer Interaction literature and operationalised for
the specific constraints of the Vietnamese rural clinician
population introduced in
Section~\ref{sec:healthcare_triage}.
Each pattern is accompanied by a formal specification, a
rationale, and an implementation sketch.

% ------------------------------------------------------------------
\subsubsection{The Confidence Spectrum: A Unified Visual Language}
\label{subsubsec:confidence-spectrum}
% ------------------------------------------------------------------

Before individual design patterns can be applied, the Trust
Dashboard must establish a \textbf{Confidence Spectrum}---a
unified visual and semantic language that maps the model's
internal Conformal Prediction set size to a discrete set of
operator-facing \textbf{Operational States}.

\paragraph{Formal State Definition.}
Let $|\mathcal{C}(x)|$ denote the cardinality of the Conformal
Prediction set for query $x$, and let $n_{\mathrm{class}}$ be
the total number of output classes.
We define four Operational States:

\begin{equation}
  \label{eq:operational-states}
  \mathrm{State}(x) =
  \begin{cases}
    \textsc{Autonomous}  & \text{if } |\mathcal{C}(x)| = 1 \\
    \textsc{Advisory}    & \text{if } 2 \leq |\mathcal{C}(x)|
                              \leq \kappa_{\mathrm{lo}} \\
    \textsc{Handover}    & \text{if } \kappa_{\mathrm{lo}} <
                              |\mathcal{C}(x)| \leq \kappa_{\mathrm{hi}} \\
    \textsc{Suspended}   & \text{if } |\mathcal{C}(x)| >
                              \kappa_{\mathrm{hi}} \text{ or }
                              |\mathcal{C}(x)| = n_{\mathrm{class}}
  \end{cases}
\end{equation}

where $\kappa_{\mathrm{lo}}$ and $\kappa_{\mathrm{hi}}$ are
deployment-specific thresholds set by the architect in
consultation with clinical domain experts.
For the Vietnamese rural triage system, we recommend
$\kappa_{\mathrm{lo}} = 2$ and $\kappa_{\mathrm{hi}} = 4$
for a 7-class triage severity taxonomy.

Table~\ref{tab:operational-states} maps each state to its
visual encoding, operator expectation, and required UI response.

\begin{table}[ht]
\centering
\caption{Confidence Spectrum: Operational States, their visual
  encodings, and required Trust Dashboard responses.
  Colour coding mirrors internationally standardised traffic-light
  semantics to minimise cross-cultural training burden.}
\label{tab:operational-states}
\small
\begin{tabular}{p{2.1cm}p{1.2cm}p{1.6cm}p{2.4cm}p{4.0cm}}
\hline\noalign{\smallskip}
\textbf{State}
  & \textbf{Colour}
  & \textbf{$|\mathcal{C}(x)|$}
  & \textbf{Operator Role}
  & \textbf{Required UI Response} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}

\textsc{Autonomous}
  & \cellcolor{stategreen}\textcolor{white}{Green}
  & $= 1$
  & Monitor
  & Display single recommendation with confidence score.
    No action required. Audit log entry created silently. \\[4pt]

\textsc{Advisory}
  & \cellcolor{stateamber}{Amber}
  & $2$--$\kappa_{\mathrm{lo}}$
  & Review
  & Display ranked candidate set with SHAP attribution.
    Soft prompt: ``Please confirm before proceeding.''
    One-click confirm or override. \\[4pt]

\textsc{Handover}
  & \cellcolor{statered}\textcolor{white}{Red}
  & $\kappa_{\mathrm{lo}}$--$\kappa_{\mathrm{hi}}$
  & Decide
  & Full Uncertainty Handover Panel (Pattern~3).
    Hard interrupt: model output suppressed until operator
    acts. Timer visible. ZK-audit log entry mandatory. \\[4pt]

\textsc{Suspended}
  & \cellcolor{stategray}\textcolor{white}{Grey}
  & $> \kappa_{\mathrm{hi}}$
  & Escalate
  & Model output fully suppressed. Emergency escalation
    protocol activated. Senior clinician paged.
    Case flagged for retrospective review. \\

\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

Figure~\ref{fig:confidence-spectrum-bar} visualises the
Confidence Spectrum as it would appear in a clinical Trust
Dashboard header.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    bar/.style={minimum height=0.65cm, minimum width=3.0cm,
                draw=white, line width=1.5pt},
    label/.style={font=\footnotesize\bfseries, align=center},
    sublabel/.style={font=\scriptsize\itshape, align=center}
  ]
  %% Four state bars
  \node[bar, fill=stategreen] (auto) at (0,0)     {};
  \node[bar, fill=stateamber] (adv)  at (3.1,0)   {};
  \node[bar, fill=statered]   (hand) at (6.2,0)   {};
  \node[bar, fill=stategray]  (susp) at (9.3,0)   {};

  %% State labels inside bars
  \node[label, text=white]  at (auto.center)
      {\textsc{Autonomous}};
  \node[label]              at (adv.center)
      {\textsc{Advisory}};
  \node[label, text=white]  at (hand.center)
      {\textsc{Handover}};
  \node[label, text=white]  at (susp.center)
      {\textsc{Suspended}};

  %% Cardinality labels below
  \node[sublabel] at (auto.south) [yshift=-8pt]
      {$|\mathcal{C}|=1$};
  \node[sublabel] at (adv.south)  [yshift=-8pt]
      {$|\mathcal{C}|=2$--$\kappa_{\mathrm{lo}}$};
  \node[sublabel] at (hand.south) [yshift=-8pt]
      {$|\mathcal{C}|=\kappa_{\mathrm{lo}}$--$\kappa_{\mathrm{hi}}$};
  \node[sublabel] at (susp.south) [yshift=-8pt]
      {$|\mathcal{C}|>\kappa_{\mathrm{hi}}$};

  %% Arrow indicating increasing uncertainty
  \draw[->, >=Stealth, thick, gray!70]
      ([xshift=-10pt]auto.west) -- ([xshift=10pt]susp.east)
      node[right, font=\scriptsize\itshape, gray]
          {increasing uncertainty $\longrightarrow$};

\end{tikzpicture}
\caption{The Confidence Spectrum as displayed in the Trust
  Dashboard header.
  The currently active state is highlighted; the others are shown
  at reduced opacity.
  Colour coding follows traffic-light semantics to leverage
  pre-existing operator intuition across cultural contexts.}
\label{fig:confidence-spectrum-bar}
\end{figure}

% ------------------------------------------------------------------
\subsubsection{Design Pattern 1: Progressive Disclosure}
\label{subsubsec:pattern-progressive-disclosure}
% ------------------------------------------------------------------

\begin{tcolorbox}[patternbox,
  title={Design Pattern 1 --- Progressive Disclosure}]
\textbf{Problem:} Displaying full model uncertainty at all times
overwhelms operators in \textsc{Autonomous} state and masks
critical signals when the state escalates to
\textsc{Handover}.

\textbf{Solution:} Reveal information in layers keyed to the
Operational State.
In \textsc{Autonomous} state, show only the top recommendation
and a single confidence percentage.
In \textsc{Advisory} state, expand to show the full candidate
set ranked by Conformal set probability.
In \textsc{Handover} state, unlock the full Uncertainty Handover
Panel (Pattern~3), SHAP attributions (Pattern~4), and the
escalation control (Pattern~5).
In \textsc{Suspended} state, suppress all model output and show
only the escalation pathway.

\textbf{Rationale:} Progressive disclosure is the primary
Human Factors countermeasure against automation bias.
By making the operator's cognitive load proportional to the
model's uncertainty, the interface trains operators to associate
increased information density with increased urgency, counteracting
the habituation that produces alarm fatigue.

\textbf{Anti-pattern:} A constant full-detail display that shows
SHAP values and prediction sets for every query, regardless of
confidence.
This is the most common implementation error in clinical AI
dashboards and is the primary driver of automation bias in
deployed systems.
\end{tcolorbox}

% ------------------------------------------------------------------
\subsubsection{Design Pattern 2: The Uncertainty Thermometer}
\label{subsubsec:pattern-thermometer}
% ------------------------------------------------------------------

\begin{tcolorbox}[patternbox,
  title={Design Pattern 2 --- The Uncertainty Thermometer}]
\textbf{Problem:} Raw Conformal Prediction set cardinality
($|\mathcal{C}(x)| = 3$) is not intuitively interpretable by a
rural clinician with limited AI literacy.
Presenting probability scores directly (e.g., ``68\% confident'')
has been shown to be misinterpreted under time pressure, with
operators systematically treating any probability above 50\% as
certain.

\textbf{Solution:} Replace numerical probability displays with a
\textbf{Uncertainty Thermometer}: a vertical bar whose fill level
maps the normalised prediction set size to a visceral visual
metaphor.
The thermometer fills from green (empty, $|\mathcal{C}|=1$) to
red (full, $|\mathcal{C}|=n_{\mathrm{class}}$).
A threshold marker at $\kappa_{\mathrm{lo}}$ provides a
pre-attentive visual cue for the \textsc{Handover} boundary.

\textbf{Formal mapping:}
\begin{equation}
  \label{eq:thermometer-fill}
  \mathrm{Fill}(x)
    = \frac{|\mathcal{C}(x)| - 1}{n_{\mathrm{class}} - 1}
    \in [0, 1]
\end{equation}

\textbf{Rationale:} Thermometer metaphors leverage pre-existing
cognitive schemas for ``more is worse'' that are culturally
robust across the Vietnamese rural population.
Unlike percentage displays, they do not require numeracy to
interpret correctly under time pressure.
The threshold marker exploits pre-attentive processing: the
operator perceives a boundary crossing without conscious
deliberation.

\textbf{Placement:} The Uncertainty Thermometer should occupy a
fixed, peripheral position in the dashboard---visible in the
operator's parafoveal field at all times---rather than appearing
only when uncertainty is high.
Constant visibility trains the operator's peripheral attention
system to register state changes automatically.
\end{tcolorbox}

% ------------------------------------------------------------------
\subsubsection{Design Pattern 3: The Uncertainty Handover Panel}
\label{subsubsec:pattern-handover-panel}
% ------------------------------------------------------------------

\begin{tcolorbox}[patternbox,
  title={Design Pattern 3 --- The Uncertainty Handover Panel}]
\textbf{Problem:} When \textsc{Handover} state is triggered, the
operator must make a consequential decision quickly and correctly.
The interface must simultaneously communicate: what the model
considered, why it is uncertain, what the stakes are, and what
the operator's options are---without inducing cognitive overload.

\textbf{Solution:} The \textbf{Uncertainty Handover Panel} is a
modal overlay that activates exclusively on transition to
\textsc{Handover} state and blocks all other interface interaction
until the operator acts.
It contains exactly four zones, each with a strict information
budget:

\begin{enumerate}
  \item \textbf{Alert Zone} (top, full width, red background):
    A single sentence in plain Vietnamese stating the
    uncertainty trigger.
    \emph{No numerical probabilities in this zone.}

  \item \textbf{Candidate Zone} (left 60\%):
    The full Conformal Prediction set $\mathcal{C}(x)$,
    displayed as a ranked list of at most five items.
    Each candidate shows its plain-language label and a
    visual confidence bar (not a number).
    The top candidate is pre-selected but not pre-confirmed.

  \item \textbf{Evidence Zone} (right 40\%):
    Three to five key clinical features that most influenced
    the model's uncertainty, rendered as a horizontal bar
    chart using SHAP values.
    Labels must use clinical terminology consistent with
    the operator's training, not feature engineering names.

  \item \textbf{Action Zone} (bottom, full width):
    Three and only three buttons:
    \textbf{[Confirm Top Recommendation]},
    \textbf{[Select Alternative]}, and
    \textbf{[Escalate to Senior Clinician]}.
    A visible countdown timer ($T_{\mathrm{respond}} = 120$\,s
    for triage contexts) applies light time pressure without
    inducing panic.
    After timeout, the system automatically activates
    \textsc{Suspended} state and pages the senior clinician.
\end{enumerate}

\textbf{Rationale:} The modal overlay enforces the
Human-in-the-Loop (HITL) requirement of
Section~\ref{sec:agency_spectrum} by making it architecturally
impossible for the operator to ignore the handover.
The information budget (one sentence, five candidates, five
features, three buttons) is derived from Miller's Law---the
cognitive limit of $7 \pm 2$ items in working
memory---adapted conservatively for
high-stress clinical environments where working memory capacity
is reduced.
\end{tcolorbox}

Figure~\ref{fig:handover-panel-layout} illustrates the spatial
layout of the Uncertainty Handover Panel.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    zone/.style={draw, rounded corners=3pt, minimum width=11.2cm,
                 font=\small, align=left, inner sep=8pt},
    every node/.append style={anchor=north west}
  ]

  %% Panel background
  \fill[uibg, rounded corners=6pt]
      (-0.2, 0.4) rectangle (11.4, -11.6);
  \draw[statered, line width=2pt, rounded corners=6pt]
      (-0.2, 0.4) rectangle (11.4, -11.6);

  %% Zone 1: Alert
  \fill[statered, rounded corners=3pt]
      (0, 0) rectangle (11.2, -1.4);
  \node[text=white, font=\small\bfseries, align=center,
        anchor=center] at (5.6, -0.7)
      {UNCERTAINTY ALERT --- PLEASE REVIEW AND DECIDE};

  %% Zone 2: Candidate (left 60%)
  \fill[white, rounded corners=3pt]
      (0, -1.6) rectangle (6.6, -8.8);
  \node[font=\footnotesize\bfseries, xshift=6pt] at (0,-1.6)
      [yshift=-12pt]{CANDIDATE DIAGNOSES};
  \foreach \y/\label/\w in {
      -2.4/{1. Acute Appendicitis}/5.2,
      -3.4/{2. Mesenteric Adenitis}/3.8,
      -4.4/{3. Ovarian Cyst (if applicable)}/3.1,
      -5.4/{4. Gastroenteritis}/2.4,
      -6.4/{5. Irritable Bowel Syndrome}/1.6
  }{
    \node[font=\footnotesize, xshift=8pt] at (0,\y) {\label};
    \fill[stateamber, rounded corners=2pt, opacity=0.7]
        (3.6, \y+0.3) rectangle (3.6+\w*0.18, \y-0.05);
  }
  \draw[stategreen, dashed, line width=1pt]
      (0.1,-2.15) rectangle (6.5,-2.85);
  \node[font=\scriptsize\itshape, stategreen, xshift=8pt]
      at (0,-2.9) {Pre-selected (not confirmed)};

  %% Zone 3: Evidence (right 40%)
  \fill[white, rounded corners=3pt]
      (6.8, -1.6) rectangle (11.2, -8.8);
  \node[font=\footnotesize\bfseries, xshift=6pt] at (6.8,-1.6)
      [yshift=-12pt]{KEY EVIDENCE (SHAP)};
  \foreach \y/\feat/\val in {
      -2.5/{Rebound tenderness}/0.82,
      -3.4/{WBC count $>$12k}/0.71,
      -4.3/{Pain migration}/0.58,
      -5.2/{Fever $>$38.5C}/0.44,
      -6.1/{Nausea duration}/0.29
  }{
    \node[font=\scriptsize, xshift=6pt] at (6.8,\y) {\feat};
    \fill[blue!55, rounded corners=1pt]
        (6.9, \y-0.28) rectangle (6.9+\val*3.8, \y-0.58);
    \node[font=\scriptsize, xshift=3pt]
        at (6.9+\val*3.8, \y-0.43) {\val};
  }

  %% Uncertainty Thermometer (far right)
  \fill[gray!15, rounded corners=2pt]
      (10.7, -1.6) rectangle (11.15, -8.8);
  \fill[stateamber, rounded corners=2pt]
      (10.7, -5.6) rectangle (11.15, -8.8);
  \draw[statered, dashed, line width=0.8pt]
      (10.65,-5.6) -- (11.2,-5.6)
      node[right, font=\tiny, statered] {$\kappa$};
  \node[font=\tiny, rotate=90, anchor=center]
      at (10.925,-5.2) {UNCERTAINTY};

  %% Zone 4: Action
  \fill[white, rounded corners=3pt]
      (0, -9.0) rectangle (11.2, -11.4);
  \fill[stategreen, rounded corners=3pt]
      (0.2, -9.2) rectangle (3.5, -10.4);
  \node[text=white, font=\footnotesize\bfseries, align=center]
      at (1.85,-9.8) {CONFIRM TOP\\RECOMMENDATION};
  \fill[stateamber, rounded corners=3pt]
      (3.7, -9.2) rectangle (7.5, -10.4);
  \node[font=\footnotesize\bfseries, align=center]
      at (5.6,-9.8) {SELECT\\ALTERNATIVE};
  \fill[statered, rounded corners=3pt]
      (7.7, -9.2) rectangle (11.0, -10.4);
  \node[text=white, font=\footnotesize\bfseries, align=center]
      at (9.35,-9.8) {ESCALATE TO\\SENIOR CLINICIAN};

  %% Countdown timer
  \node[font=\footnotesize, align=center, statered]
      at (5.6,-10.9)
      {Response required within: \textbf{1:47} \quad
       \textit{(auto-escalates at 0:00)}};

\end{tikzpicture}
\caption{Spatial layout of the Uncertainty Handover Panel
  (Design Pattern~3).
  The panel activates as a modal overlay on transition to
  \textsc{Handover} state.
  Ranked candidate set, SHAP evidence bars,
  Uncertainty Thermometer, and three-action control
  are the only elements present.
  All other dashboard content is blurred behind the overlay.}
\label{fig:handover-panel-layout}
\end{figure}

% ------------------------------------------------------------------
\subsubsection{Design Pattern 4: Contextual SHAP Narration}
\label{subsubsec:pattern-shap-narration}
% ------------------------------------------------------------------

\begin{tcolorbox}[patternbox,
  title={Design Pattern 4 --- Contextual SHAP Narration}]
\textbf{Problem:} Raw SHAP value bar charts, while
mathematically correct, are not
interpretable by clinicians without data-science training.
Presenting them without translation into clinical language
creates a display that satisfies the explainability requirement
technically while failing it practically.

\textbf{Solution:} Wrap the SHAP computation in a
\textbf{Narration Layer} that translates feature importances into
one or two sentences of plain clinical language, generated by a
locally-deployed Small Language Model (SLM) that has been fine-tuned
on the institution's clinical terminology.
The narration is displayed above the SHAP bar chart as the
primary explanation, with the chart as a secondary verification
resource for trained users.

\textbf{Example narration} (generated from SHAP values in
Figure~\ref{fig:handover-panel-layout}):
\begin{quote}
\small\itshape
``The system is most uncertain because the patient's rebound
tenderness and elevated white cell count both suggest acute
appendicitis, but the absence of fever migration makes
alternative diagnoses plausible.
Please assess the right iliac fossa manually.''
\end{quote}

\textbf{Formal narration constraint:}
The narration must satisfy three constraints verified
programmatically before display:
(1)~maximum 40 words;
(2)~Flesch-Kincaid reading grade level $\leq 8$
(appropriate for a secondary-educated Vietnamese reader);
(3)~no numerical probabilities or SHAP scores in the text.

\textbf{Anti-pattern:} Narrations that begin with
``The model predicts\ldots'' or ``With X\% confidence\ldots''
These phrasings have been shown to significantly increase
automation bias by priming the operator to defer to the
model rather than exercise independent judgement.
\end{tcolorbox}

% ------------------------------------------------------------------
\subsubsection{Design Pattern 5: The Graceful Degradation
               Ladder}
\label{subsubsec:pattern-degradation-ladder}
% ------------------------------------------------------------------

\begin{tcolorbox}[patternbox,
  title={Design Pattern 5 --- The Graceful Degradation Ladder}]
\textbf{Problem:} The Uncertainty Handover is not a binary
event---it is a continuous process that may require multiple
escalation steps before a final decision is reached.
Without a defined escalation protocol, operators in rural
settings with limited specialist access may be left holding an
unresolved case with no clear next action.

\textbf{Solution:} Implement a \textbf{Graceful Degradation
Ladder}: a pre-defined, role-aware escalation sequence that the
Trust Dashboard follows automatically when operator response
does not resolve the uncertainty within defined time limits.

\begin{equation}
  \label{eq:degradation-ladder}
  \mathrm{Ladder} =
  \bigl[
    \underbrace{\text{Model}}_{\text{step 0}}
    \to
    \underbrace{\text{Nurse}}_{\text{step 1}}
    \to
    \underbrace{\text{GP}}_{\text{step 2}}
    \to
    \underbrace{\text{Specialist (remote)}}_{\text{step 3}}
    \to
    \underbrace{\text{Transfer protocol}}_{\text{step 4}}
  \bigr]
\end{equation}

At each step, the Trust Dashboard:
(1)~displays the current escalation step prominently;
(2)~surfaces the contact mechanism for the next-step
responder (pager ID, Zalo contact, telemedicine link);
(3)~maintains a live timer showing elapsed time since
initial handover;
(4)~writes a timestamped, ZK-verified audit log entry.

\textbf{Architectural requirement:} The Degradation Ladder must
be configured per deployment by clinical governance staff,
not by the AI architect.
The architect's role is to implement the ladder mechanism and
the audit infrastructure; the clinical team defines the roles,
time limits, and escalation contacts.
This separation of concerns is the technical expression of
the Human-in-Command (HiC) principle of
Section~\ref{sec:agency_spectrum}.

\textbf{Suspension guarantee:} If the ladder reaches step 4
without resolution, the AI system enters
\textsc{Suspended} state (Equation~\eqref{eq:operational-states})
and cannot be reactivated for this case without a senior
clinician's authenticated override.
\end{tcolorbox}

% ------------------------------------------------------------------
\subsubsection{Implementation: The Trust Dashboard Controller}
\label{subsubsec:dashboard-implementation}
% ------------------------------------------------------------------

Listing~\ref{lst:dashboard-controller} provides a minimal
implementation of the Trust Dashboard state machine, connecting
the Conformal Prediction output of
Section~\ref{sec:uncertainty_handover} to the five design
patterns above.

\begin{lstlisting}[
  style=pythonstyle,
  caption={Trust Dashboard Controller: maps Conformal Prediction
           set cardinality to Operational States and triggers the
           appropriate design pattern.},
  label={lst:dashboard-controller}
]
from enum import Enum
from dataclasses import dataclass, field
from typing import Optional
import time

class OperationalState(Enum):
    AUTONOMOUS = "autonomous"   # |C(x)| = 1
    ADVISORY   = "advisory"     # 2 <= |C(x)| <= kappa_lo
    HANDOVER   = "handover"     # kappa_lo < |C(x)| <= kappa_hi
    SUSPENDED  = "suspended"    # |C(x)| > kappa_hi

@dataclass
class DashboardConfig:
    n_classes:      int   = 7     # triage severity classes
    kappa_lo:       int   = 2     # advisory->handover boundary
    kappa_hi:       int   = 4     # handover->suspended boundary
    respond_secs:   int   = 120   # handover response time limit
    max_narration_words: int = 40 # Pattern 4 constraint
    max_fk_grade:   float = 8.0   # Pattern 4 reading level

@dataclass
class HandoverPacket:
    """Data required to render the Uncertainty Handover Panel."""
    state:          OperationalState
    candidates:     list       # (label, score) tuples
    top_features:   list       # (feature, shap_val) tuples
    narration:      str
    thermometer_fill: float    # Eq. (3.2)
    timestamp:      float = field(default_factory=time.time)
    audit_token:    Optional[str] = None

class TrustDashboardController:
    def __init__(self, config, audit_logger,
                 slm_narrator, escalation_ladder):
        self.cfg      = config
        self.auditor  = audit_logger
        self.narrator = slm_narrator
        self.ladder   = escalation_ladder

    def classify_state(self, prediction_set):
        """Map |C(x)| to Operational State (Eq. 3.1)."""
        size = len(prediction_set)
        if   size == 1:                  return OperationalState.AUTONOMOUS
        elif size <= self.cfg.kappa_lo:  return OperationalState.ADVISORY
        elif size <= self.cfg.kappa_hi:  return OperationalState.HANDOVER
        else:                            return OperationalState.SUSPENDED

    def thermometer_fill(self, prediction_set):
        """Normalised fill level (Eq. 3.2)."""
        return (len(prediction_set) - 1) / (self.cfg.n_classes - 1)

    def build_handover_packet(self, prediction_set, shap_values):
        """Construct the full HandoverPacket for UI rendering."""
        state = self.classify_state(prediction_set)

        # Generate SLM narration (Pattern 4)
        narration = self.narrator.generate(
            candidates   = prediction_set,
            top_features = shap_values[:5],
            max_words    = self.cfg.max_narration_words,
            max_fk_grade = self.cfg.max_fk_grade,
            forbidden_patterns = [r'\d+%', r'probability'],
        )

        # Write ZK-verified audit log entry
        audit_token = self.auditor.log(
            state      = state.value,
            candidates = prediction_set,
            features   = shap_values[:5],
            narration  = narration,
        )

        return HandoverPacket(
            state            = state,
            candidates       = prediction_set[:5],
            top_features     = shap_values[:5],
            narration        = narration,
            thermometer_fill = self.thermometer_fill(prediction_set),
            audit_token      = audit_token,
        )

    def handle(self, prediction_set, shap_values):
        """Main entry point: build packet, activate ladder if needed."""
        packet = self.build_handover_packet(prediction_set, shap_values)

        if packet.state == OperationalState.SUSPENDED:
            self.ladder.escalate(
                step    = 1,
                packet  = packet,
                timeout = self.cfg.respond_secs,
            )
        return packet
\end{lstlisting}

% ------------------------------------------------------------------
\subsubsection{Evaluation: The Trust Dashboard Acceptance Protocol}
\label{subsubsec:dashboard-evaluation}
% ------------------------------------------------------------------

Before production deployment, a Trust Dashboard implementation
must pass a structured \textbf{Acceptance Protocol} conducted
with representative end-users---in the rural clinic context, a
stratified sample of nurses and general practitioners from the
target communes.

The protocol evaluates five outcomes, each with a defined
minimum threshold:

\begin{table}[ht]
\centering
\caption{Trust Dashboard Acceptance Protocol: evaluation outcomes,
  measurement instruments, and minimum acceptable thresholds.
  All thresholds are conservative estimates appropriate for a
  patient-safety-critical clinical AI deployment.}
\label{tab:dashboard-acceptance}
\begin{tabular}{p{3.4cm}p{3.8cm}p{2.0cm}p{2.4cm}}
\hline\noalign{\smallskip}
\textbf{Outcome}
  & \textbf{Measurement Instrument}
  & \textbf{Sample Size}
  & \textbf{Min.\ Threshold} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
State Recognition Accuracy
  & 10-scenario quiz: operator identifies
    correct Operational State from screenshot
  & $n \geq 20$
  & $\geq 90\%$ \\[4pt]
Handover Response Time
  & Median time from panel activation to
    operator action, measured in simulation
  & $n \geq 20$
  & $\leq 45$\,s \\[4pt]
Automation Bias Rate
  & \% of trials where operator confirms
    a deliberately incorrect top candidate
    in \textsc{Handover} state
  & $n \geq 20$
  & $\leq 10\%$ \\[4pt]
Alarm Fatigue Index
  & Self-reported desensitisation score
    (Likert 1--5) after 2-hour simulation
  & $n \geq 20$
  & $\leq 2.0$ \\[4pt]
Narration Comprehension
  & Cloze test: operator identifies correct
    clinical action from narration alone
  & $n \geq 20$
  & $\geq 85\%$ \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

The Automation Bias Rate threshold of $\leq 10\%$ deserves
particular emphasis.
A system where one in ten handovers results in uncritical
acceptance of an uncertain AI recommendation is not a
trustworthy system---it is a system that has successfully
\emph{disguised} its uncertainty from its operators.
This threshold must be re-evaluated quarterly in production
through randomised ``trust probes'': deliberate injection of
synthetic high-uncertainty cases into the live query stream to
verify that operators continue to engage critically with
\textsc{Handover} panel activations.

% ------------------------------------------------------------------
\subsubsection{Conclusion: The Interface Is the Architecture}
\label{subsubsec:dashboard-conclusion}
% ------------------------------------------------------------------

The five design patterns of this section---Progressive Disclosure,
the Uncertainty Thermometer, the Handover Panel, Contextual SHAP
Narration, and the Graceful Degradation Ladder---are not cosmetic
additions to the Trustworthy AI Stack.
They are the human-facing implementation of the socio-technical
contract established in Section~\ref{sec:agency_spectrum}.

A system that produces mathematically correct Conformal Prediction
sets, writes cryptographically verified audit logs, and satisfies
$(\varepsilon, \delta)$-DP guarantees, but presents its uncertainty
to a rural clinician through a text field labelled
``Confidence: 0.43''---has failed the most important
architectural test of all: the test of human comprehension under
pressure.

The Trust Dashboard is not where the architecture ends.
It is where the architecture proves whether it was worth building.

\section{Case Study: Agentic Healthcare Triage}
\label{sec:healthcare_triage}

Consider a 2026 autonomous triage system deployed in a Vietnamese rural clinic. The system handles initial patient intake:
\begin{itemize}
    \item \textbf{Autonomy:} It processes vitals and medical history using a localized SLM.
    \item \textbf{The Handover:} Using conformal prediction, if a patient's symptoms are outside its high-confidence training distribution, it flags the file for an ``Urgent Human Review.''
    \item \textbf{Verification:} Every triage decision is recorded with a ZK-proof that verifies the model was executed correctly against the clinic's authorized safety weights from the Ministry of Health.
\end{itemize}

\section{Conclusion of Part I}
\label{sec:conclusion_part1}

The Strategic Vision \& Foundations of Trustworthy AI are now complete. We have the \textbf{2026 Vision} (Chapter 1), the \textbf{Sociotechnical Norms} (Chapter 2), and the \textbf{Human Agency Framework} (Chapter 3). 

We have established that trust is built on four action-oriented pillars: Security, Privacy, Verifiability, and Alignment. In Part II, we descend from these high-level philosophies into the rigorous technical landscape of \textbf{Privacy Engineering \& Data Defense}, beginning with the classical methods that formed the bedrock of early privacy technology.