%%%%%%%%%%%%%%%%%%%%% Chapter 8 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Secure Aggregation: Cryptography & MPC
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\motto{Secure Aggregation allows a server to compute the average of client updates without ever seeing any single update in the clear.}

\chapter{Secure Aggregation: Cryptography \& MPC}
\label{chap:secure_aggregation}

\abstract{Building on the decentralized architecture of Federated Learning, this chapter introduces the cryptographic protocols necessary to protect model updates from a curious or compromised central server. We explore the principles of Multi-Party Computation (MPC) and Secure Aggregation, using the analogy of "sealed envelopes" to explain pairwise masking. We analyze popular schemes like Bonawitz SecAgg \cite{bonawitz2017} and Homomorphic Encryption (CKKS), while addressing the critical trade-offs between cryptographic strength and system performance.}

\section{The Curious Server Problem}
\label{sec:curious_server}

In standard Federated Learning, clients send their model updates (gradients) to a central aggregator. While the raw data stays on the device, the gradients themselves can leak sensitive information. A "curious" server---one that follows the protocol but attempts to learn as much as possible---can use gradient inversion to reconstruct input data. Secure Aggregation is the cryptographic guardrail that prevents this by ensuring the server only ever sees the \textit{aggregate} sum, never the individual contributions.

\begin{svgraybox}
\textbf{The Analogy: Sealed Envelopes.} Imagine 100 hospitals want to sum their patient counts without revealing individual counts to a coordinator. Each hospital adds a unique random mathematical "mask" to their count before placing it in a sealed envelope. When the coordinator sums all the envelopes, the masks cancel each other out perfectly, revealing only the total sum.
\end{svgraybox}

\section{The Math of Pairwise Masking}
\label{sec:masking_math}

The goal of Secure Aggregation is to compute $\sum w_i$ without revealing any $w_i$. This is achieved through pairwise masking. 

Each client $i$ generates a random seed $s_{i,j}$ shared with client $j$. The client then uploads a masked update $u_i = w_i + r_i$, where $r_i$ is the sum of shared secrets:
\begin{equation}
r_i = \sum_{j \in N_i} (s_{i,j} - s_{j,i}) \pmod P
\end{equation}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    client/.style={draw, rounded corners=4pt, fill=green!8, draw=green!55,
                   inner sep=8pt, font=\footnotesize\bfseries, align=center,
                   minimum width=3.2cm, minimum height=1.1cm},
    srv/.style={draw, rounded corners=5pt, fill=blue!10, draw=blue!60,
                inner sep=10pt, font=\small\bfseries, align=center,
                minimum width=3.6cm, minimum height=1.6cm},
    mask/.style={draw, rounded corners=2pt, fill=orange!8, draw=orange!50,
                 inner sep=5pt, font=\scriptsize, align=center, minimum width=3.2cm},
    arr/.style={->, >=Stealth, thick, gray!65},
    note/.style={font=\scriptsize\itshape}
  ]

  %% Three clients on left
  \node (c1) [client] at (0, 2.2)  {\textbf{Client 1}\\$w_1$ (true update)};
  \node (c2) [client] at (0, 0)    {\textbf{Client 2}\\$w_2$ (true update)};
  \node (c3) [client] at (0,-2.2)  {\textbf{Client 3}\\$w_3$ (true update)};

  %% Masked values computed by each client
  \node (m1) [mask, right=1.4cm of c1]
      {$u_1 = w_1 + r_1$\\{\scriptsize $r_1 = s_{12} - s_{21} + s_{13} - s_{31}$}};
  \node (m2) [mask, right=1.4cm of c2]
      {$u_2 = w_2 + r_2$\\{\scriptsize masks sum to zero across all clients}};
  \node (m3) [mask, right=1.4cm of c3]
      {$u_3 = w_3 + r_3$\\{\scriptsize $r_3 = s_{31} - s_{13} + s_{32} - s_{23}$}};

  %% Server on right
  \node (srv) [srv] at (9.5,0)
      {\textbf{Aggregation Server}\\[4pt]
       $\sum_i u_i = \sum_i w_i$\\[2pt]
       {\scriptsize masks cancel: $\sum_i r_i = 0$}\\[2pt]
       {\scriptsize server sees \emph{only} the sum}};

  %% Arrows client → mask
  \draw[arr] (c1.east) -- (m1.west);
  \draw[arr] (c2.east) -- (m2.west);
  \draw[arr] (c3.east) -- (m3.west);

  %% Masked updates → server
  \draw[arr] (m1.east) -- node[note, above, sloped] {blinded transit} (srv.north west);
  \draw[arr] (m2.east) -- node[note, above] {blinded transit} (srv.west);
  \draw[arr] (m3.east) -- node[note, below, sloped] {blinded transit} (srv.south west);

  %% Shared-secret annotations between clients
  \draw[<->, gray!40, dashed, thin] (c1.south) -- node[note, left] {$s_{12}$} (c2.north);
  \draw[<->, gray!40, dashed, thin] (c2.south) -- node[note, left] {$s_{23}$} (c3.north);
  \draw[<->, gray!40, dashed, thin, bend left=25]
      (c1.south west) to node[note, left] {$s_{13}$} (c3.north west);

  \node[note, gray!60, align=center] at (0,-3.3)
      {Pairwise seeds $s_{ij}$ exchanged\\via Diffie-Hellman; never sent to server};

\end{tikzpicture}
\caption{Secure Aggregation Logic: Pairwise masking. Each client adds a cancelling
         random mask $r_i$ derived from shared seeds $s_{ij}$ before uploading. Because
         $\sum_i r_i = 0$ across all clients, the server recovers $\sum_i w_i$ exactly
         without ever observing any individual update $w_i$ in the clear.}
\label{fig:ch8_secagg}
\end{figure}

When the server aggregates all $u_i$, every secret $s_{i,j}$ is added exactly once and subtracted exactly once across the entire network, causing the masks to effectively vanish: $\sum u_i = \sum w_i$.

\section{Popular Schemes and Protocols}
\label{sec:secagg_protocols}

Choosing the right protocol depends on the computational constraints of the clients and the required robustness to network dropout.

\begin{table}
\caption{Comparison of Secure Computation Schemes}
\label{tab:secagg_schemes}
\begin{tabular}{p{3cm}p{3cm}p{4.5cm}}
\hline\noalign{\smallskip}
Scheme & Category & Best For \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Bonawitz SecAgg & MPC & Standard FL with high client dropout. \\
CKKS & Homomorphic Enc. & Deep Learning (Approximate arithmetic). \\
Paillier & Partial HE & Simple summation tasks. \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\section{Trade-offs: Privacy vs. Performance}
\label{sec:secagg_tradeoffs}

Cryptographic defenses are not "free"; they introduce significant overhead that must be balanced by the Trustworthy AI Architect.

\begin{description}
  \item[Communication Overhead:] 
        Protocols like SecAgg require extra rounds for key exchange and secret sharing, which can increase latency in low-bandwidth environments.
  \item[Computational Cost:]
        Homomorphic Encryption (HE) can be 100x to 1000x slower than plaintext computation.
  \item[Robustness to Dropout:]
        If a client drops out after sending their masked update but before the unmasking phase, the server cannot recover the aggregate without specific "secret sharing recovery" mechanisms.
\end{description}

% ------------------------------------------------------------------
\section{Implementation Sketch: Secure FL with Flower and SecAgg}
\label{sec:secagg_implementation}
% ------------------------------------------------------------------

The theory of pairwise masking becomes architectural reality with \textbf{Flower (\texttt{flwr})} \cite{flower}, an open-source FL framework with a built-in SecAgg+ protocol. The following sketch wires a SecAgg-protected aggregation server with DP-noised clients in under 30 lines.

\begin{lstlisting}[language=Python, style=pythonstyle,
  caption={Secure FL with Flower \texttt{flwr} and \texttt{SecAggPlusWorkflow}. The server attaches pairwise masking via the workflow; each client applies DP-SGD noise \emph{before} the mask is applied, giving double-layer protection: even if the SecAgg mask were broken, the DP noise prevents gradient inversion.},
  label={lst:flower_secagg}]
import flwr as fl
from flwr.server.workflow import SecAggPlusWorkflow
from opacus import PrivacyEngine
import torch

# --- Server: attach SecAgg+ workflow to standard FedAvg ---------
strategy = fl.server.strategy.FedAvg(
    fraction_fit=0.8, min_fit_clients=5, min_available_clients=5,
)
workflow = SecAggPlusWorkflow(
    num_shares=5,               # secret-sharing degree
    reconstruction_threshold=3, # minimum shares to reconstruct
)
fl.server.start_server(
    server_address="0.0.0.0:8080",
    config=fl.server.ServerConfig(num_rounds=10),
    strategy=strategy, workflow=workflow,
)

# --- Client: DP-SGD noise applied before SecAgg mask ------------
class SecureFlClient(fl.client.NumPyClient):
    def __init__(self, model, train_loader):
        self.engine = PrivacyEngine()
        self.model, self.opt, self.loader = \
            self.engine.make_private_with_epsilon(
                module=model,
                optimizer=torch.optim.SGD(model.parameters(), lr=0.01),
                data_loader=train_loader,
                epochs=1, target_epsilon=3.0,
                target_delta=1e-5, max_grad_norm=1.0)

    def fit(self, parameters, config):
        # DP-SGD local training; Flower runtime applies SecAgg mask after
        for X, y in self.loader:
            self.opt.zero_grad()
            torch.nn.CrossEntropyLoss()(self.model(X), y).backward()
            self.opt.step()
        eps = self.engine.get_epsilon(delta=1e-5)
        return self.model.parameters(), len(self.loader), {"eps": eps}
\end{lstlisting}

\begin{svgraybox}
\textbf{Architect's Note --- Layer Ordering:} DP noise must be injected \emph{before} the SecAgg mask, not after. The correct stack is: \textbf{[per-sample gradient clip]} $\to$ \textbf{[Gaussian noise injection]} $\to$ \textbf{[pairwise SecAgg masking]} $\to$ \textbf{[server aggregation]}. Adding noise after the mask would mean the DP accounting applies to the \emph{aggregate}, not to each individual update — violating the per-client privacy guarantee.
\end{svgraybox}

\section{Conclusion}
\label{sec:conclusion_secagg}

Secure Aggregation ensures that the central server remains a neutral aggregator rather than a potential spy. By combining the decentralization of FL with the cryptographic protection of MPC, we build the first layer of trust in distributed systems. However, cryptography only protects data \textit{in transit}. To protect the computation \textit{in use}, we must turn to the hardware layer, which we will explore in the next chapter on Trusted Execution Environments.


