%%%%%%%%%%%%%%%%%%%%% Chapter 24 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AI Supply Chain Security: SBOM & Provenance
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\motto{Trust is not just about the final model weights; it is about the integrity of every link in the AI supply chain.}

\chapter{AI Supply Chain Security: SBOM \& Provenance}
\label{chap:supply_chain}

\abstract{A modern AI model is the product of an incredibly complex supply chain involving raw datasets, open-source base models, third-party libraries, and highly specialized hardware. In this chapter, we address the security of the entire machine learning pipeline. We define the concept of a "Software Bill of Materials" (SBOM) for AI, explore methods for tracking Model Provenance and lineage, and analyze frameworks like SLSA (Supply-chain Levels for Software Artifacts) for ensuring the integrity of the transformation from raw data to a deployed model. Securing the supply chain ensures that the trust we build into the model cannot be undermined by a compromised dependency or a hidden vulnerability.}

\section{The ML Supply Chain: A New Surface for Attack}
\label{sec:ml_supply_chain}

The Trustworthy AI Architect must look beyond the weights of the final model and toward the process that created them. The ML supply chain includes three primary vectors:

\begin{description}
  \item[Data Integrity:] Ensuring that the training data has not been poisoned (Chapter 6) or tampered with before reaching the training pipeline.
  \item[Model Ancestry:] Tracking the lineage of base models (e.g., Llama-3, PhoGPT) to ensure they have not been compromised with backdoors.
  \item[Code Provenance:] Verifying the integrity of the libraries (Opacus, Flower, PyTorch) and the CI/CD pipelines used to train and deploy the model.
\end{description}

\section{SBOM for AI: Documenting the Ingredients}
\label{sec:sbom_ai}

Just as food products list their ingredients, AI models must provide a \textbf{Software Bill of Materials (SBOM)}. An AI SBOM is a formal, machine-readable record of all the components that went into the model's creation. This includes the versions of the training scripts, the checksums of the datasets, and the specific hyperparameter configurations.

By maintaining an SBOM, an organization can quickly identify if it is vulnerable to a newly discovered security flaw in one of its dependencies or if its training data has been subject to a known poisoning campaign.

\section{Model Provenance: Tracking Lineage}
\label{sec:provenance_lineage}

\textbf{Model Provenance} is the ability to reconstruct the "autobiography" of an AI model. The architect should maintain a signed lineage graph that tracks:
\begin{itemize}
    \item \textbf{Input Artifacts:} Digital signatures for datasets and base models.
    \item \textbf{Transformation Logs:} Records of the training environments and hardware root-of-trust (Chapter 9).
    \item \textbf{Certification Proofs:} Links to formal verification proofs (Chapter 23) and ZK-verification keys (Chapter 22).
\end{itemize}

\section{Securing the Pipeline with SLSA}
\label{sec:slsa}

To achieve high levels of supply chain security, architects should adopt the \textbf{SLSA} framework. SLSA provides a set of incrementally more rigorous requirements for building software, ensuring that artifacts are built in isolated, ephemeral environments and that every build step is recorded in a non-falsifiable audit trail. For AI, this means that the ``weights'' are not just a file, but a verifiably correct output of a secure build process.

% ==================================================================
\section{Detecting Synthetic Contamination in the AI Supply Chain}
\label{sec:synthetic-contamination}
% ==================================================================

\begin{quote}
\textit{``The most insidious supply chain attack of 2026 is not
a poisoned gradient or a backdoored checkpoint.
It is a training dataset that looks perfectly clean---because
it was generated by a model that was itself trained on
generated data, recursively, until the boundary between human
knowledge and hallucinated knowledge has dissolved.''}
\end{quote}

Sections~\ref{sec:ml_supply_chain}--\ref{sec:slsa} addressed
the integrity of the AI supply chain from the perspective of
malicious tampering: a poisoned dataset, a backdoored base model,
a compromised CI/CD pipeline.
This section addresses a fundamentally different threat: the
\textbf{accidental contamination} of training data with
AI-generated content that degrades model quality not through
adversarial intent, but through recursive self-reference.

This is the problem of \textbf{Model Collapse}---a 2025--2026
phenomenon in which large language models trained on data
generated by earlier generations of language models exhibit
systematic degradation in output diversity, factual accuracy,
and distributional coverage.
The collapse is not a single-event failure; it is a gradual,
self-reinforcing process that may take months or years to
become observable in downstream metrics, by which time the
contaminated data has propagated through the global data
supply chain to affect every model trained on web-crawled
corpora.

For the Trustworthy AI Architect, model collapse is a
\emph{supply chain integrity} problem that demands the same
architectural discipline as data poisoning---but requires
entirely different detection mechanisms.

% ------------------------------------------------------------------
\subsection{The Model Collapse Mechanism: Recursive Poisoning
            of the Knowledge Pool}
\label{subsec:model-collapse-mechanism}
% ------------------------------------------------------------------

The mechanism of model collapse can be formalised as a
recursive degradation of the training distribution.

\paragraph{Generation 0: Human Data.}
Let $\mathcal{D}_0$ be the original training distribution,
generated entirely by human authors.
A model $\mathcal{M}_0$ trained on $\mathcal{D}_0$ learns to
approximate the true distribution $p_{\mathrm{human}}(x)$.

\paragraph{Generation 1: First Synthetic Layer.}
$\mathcal{M}_0$ generates synthetic text $\hat{\mathcal{D}}_1$
that is published on the web.
A new training crawl ingests a mixture:
\begin{equation}
  \label{eq:contaminated-distribution}
  \mathcal{D}_1
    = (1 - \gamma_1) \cdot \mathcal{D}_0
    + \gamma_1 \cdot \hat{\mathcal{D}}_1
\end{equation}
where $\gamma_1 \in [0, 1]$ is the synthetic contamination
ratio.
Model $\mathcal{M}_1$ trained on $\mathcal{D}_1$ now
learns a distribution that over-represents the modes
favoured by $\mathcal{M}_0$ and under-represents the
distributional tails of $p_{\mathrm{human}}$.

\paragraph{Generation $k$: Recursive Amplification.}
After $k$ generations:
\begin{equation}
  \label{eq:recursive-collapse}
  \mathcal{D}_k
    = (1 - \gamma_k) \cdot \mathcal{D}_{k-1}
    + \gamma_k \cdot \hat{\mathcal{D}}_k
  \qquad
  \text{with } \hat{\mathcal{D}}_k \sim \mathcal{M}_{k-1}
\end{equation}

Each generation amplifies the mode collapse of the previous
generation.
The distributional tails---minority languages, rare medical
terminology, non-standard dialects, edge-case legal
reasoning---are systematically eroded.
By generation $k = 3$--$5$, empirical studies report a
20--40\% reduction in lexical diversity and a measurable
increase in factual hallucination rate.

\paragraph{The Trust Implication.}
For a Trustworthy AI system, synthetic contamination is not
merely a quality problem.
It is a \textbf{fairness failure}: if the training data
under-represents minority distributions because synthetic
data has overwritten them, the model's predictions for those
populations will degrade---exactly the failure mode that the
Conformal Prediction and Formal Verification mechanisms of
earlier chapters were designed to detect.
A model trained on recursively contaminated data may pass
accuracy benchmarks on majority populations while silently
failing on the distributional tails where trust matters most.

% ------------------------------------------------------------------
\subsection{A Taxonomy of Synthetic Contamination Sources}
\label{subsec:contamination-taxonomy}
% ------------------------------------------------------------------

Not all synthetic data is a contamination risk.
The architect must distinguish between three categories:

\begin{description}
  \item[Type~A: Controlled Synthetic Data.]
    Data generated intentionally by the organisation for a
    specific purpose (data augmentation, privacy-preserving
    synthetic datasets as described in Chapter~4).
    Controlled synthetic data is \emph{labelled} as synthetic
    in the SBOM and calibrated against a known base
    distribution.
    This is not a contamination risk; it is a design choice.

  \item[Type~B: Disclosed Third-Party Synthetic Data.]
    Data from external providers that is labelled as
    AI-generated (e.g., synthetic benchmarks, LLM-generated
    translations).
    The label allows the architect to control the
    contamination ratio $\gamma$ and apply appropriate
    weighting or filtering.

  \item[Type~C: Undisclosed Web-Crawled Synthetic Data.]
    This is the critical threat vector.
    Web-crawled corpora (Common Crawl, LAION, FineWeb)
    increasingly contain AI-generated text that is not labelled
    as such.
    A 2025 analysis estimated that 10--15\% of new English web
    content is AI-generated; for Vietnamese, the ratio may be
    lower but is growing rapidly as LLM adoption accelerates
    in Southeast Asia.
    This data enters the training pipeline silently, with no
    SBOM entry, no provenance record, and no contamination
    ratio estimate.
\end{description}

Table~\ref{tab:contamination-sources} summarises the three
contamination types with their detection difficulty and
management strategies.

\begin{table}[ht]
\centering
\caption{Synthetic contamination source taxonomy.
  Detection difficulty reflects the cost and reliability of
  identifying the contamination type in a raw training corpus.}
\label{tab:contamination-sources}
\small
\begin{tabular}{p{2.0cm}p{2.2cm}p{2.0cm}p{5.0cm}}
\hline\noalign{\smallskip}
\textbf{Type}
  & \textbf{Provenance}
  & \textbf{Detection Difficulty}
  & \textbf{Management Strategy} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}

Type~A (Controlled)
  & Internal, labelled
  & N/A
  & Record in SBOM with generation method,
    base model, and calibration report. \\[4pt]

Type~B (Disclosed)
  & Third-party, labelled
  & Low
  & Verify label integrity; apply contamination
    ratio cap $\gamma_{\max}$. \\[4pt]

Type~C (Undisclosed)
  & Unknown, unlabelled
  & High
  & Apply statistical detection (Section~\ref{subsec:detection-framework});
    quarantine flagged documents. \\

\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

% ------------------------------------------------------------------
\subsection{The Synthetic Contamination Detection Framework}
\label{subsec:detection-framework}
% ------------------------------------------------------------------

Detecting Type~C contamination---undisclosed AI-generated text
in a web-crawled corpus---requires a multi-signal approach.
No single detector is reliable in isolation; the architect
must fuse multiple weak signals into a composite
\textbf{Provenance Score} that estimates the probability that
a document is AI-generated.

\paragraph{Signal 1: Perplexity Profiling.}
AI-generated text tends to have lower perplexity under
reference language models than human-written text, because
generative models produce text that is, by construction,
high-probability under their own distribution.
The architect computes the perplexity of each document under
a reference model $\mathcal{M}_{\mathrm{ref}}$ that was
\emph{not} used to generate the training data:
\begin{equation}
  \label{eq:perplexity-score}
  s_{\mathrm{ppl}}(d)
    = 1 - \frac{
        \mathrm{PPL}_{\mathcal{M}_{\mathrm{ref}}}(d)
      }{
        \mathrm{PPL}_{\mathrm{median}}(\mathcal{D}_{\mathrm{human}})
      }
\end{equation}
A score $s_{\mathrm{ppl}}$ close to $1$ indicates
suspiciously low perplexity; a score near $0$ indicates
perplexity consistent with human text.

\paragraph{Signal 2: N-gram Repetition Analysis.}
AI-generated text exhibits characteristic n-gram repetition
patterns that differ from human writing: higher frequency of
``safe'' transitional phrases, lower lexical diversity in
long documents, and systematic over-representation of common
collocations.
The architect computes a \textbf{Repetition Ratio}:
\begin{equation}
  \label{eq:repetition-score}
  s_{\mathrm{rep}}(d)
    = 1 - \frac{|\text{unique n-grams in } d|}{|\text{total n-grams in } d|}
\end{equation}
for $n = 4$ (4-grams), normalised against the median
repetition ratio of a verified human reference corpus.

\paragraph{Signal 3: Watermark Detection.}
An increasing number of LLM providers embed statistical
watermarks in their outputs (e.g., soft watermarking via
biased token sampling).
The architect runs a battery of known watermark detectors
against each document.
A positive watermark detection is a strong (but not
conclusive) signal of synthetic origin.

\paragraph{Signal 4: Distributional Divergence.}
For each document, the architect computes the KL divergence
between the document's token distribution and the expected
distribution of the document's language and domain:
\begin{equation}
  \label{eq:divergence-score}
  s_{\mathrm{div}}(d)
    = D_{\mathrm{KL}}\!\left(
        P_d \,\|\, P_{\mathrm{ref}}^{\mathrm{domain}(d)}
      \right)
\end{equation}
Anomalously low divergence (the document is ``too typical'')
is a contamination signal.

\paragraph{Composite Provenance Score.}
The four signals are fused into a composite
\textbf{Provenance Score} $\mathcal{P}(d) \in [0, 1]$:
\begin{equation}
  \label{eq:provenance-score}
  \mathcal{P}(d)
    = v_1 \cdot s_{\mathrm{ppl}}(d)
    + v_2 \cdot s_{\mathrm{rep}}(d)
    + v_3 \cdot s_{\mathrm{wm}}(d)
    + v_4 \cdot s_{\mathrm{div}}(d)
\end{equation}
where $\sum_i v_i = 1$ and $s_{\mathrm{wm}} \in \{0, 1\}$
is the binary watermark detection result.

A document with $\mathcal{P}(d) \geq \theta_{\mathrm{quarantine}}$
is flagged for quarantine and manual review before inclusion
in the training pipeline.
The recommended default threshold is
$\theta_{\mathrm{quarantine}} = 0.65$, calibrated to achieve
$\leq 5\%$ false positive rate on verified human corpora.

% ------------------------------------------------------------------
\subsection{SBOM Integration: Recording Synthetic Provenance}
\label{subsec:sbom-synthetic-provenance}
% ------------------------------------------------------------------

The Provenance Score must be recorded in the AI SBOM
(Section~\ref{sec:sbom_ai}) as a first-class metadata field
for every dataset artifact.
We extend the SBOM schema with three new fields:

\begin{description}
  \item[\texttt{synthetic\_contamination\_ratio}:]
    The estimated fraction $\hat{\gamma}$ of synthetic content
    in the dataset, computed as the fraction of documents with
    $\mathcal{P}(d) \geq \theta_{\mathrm{quarantine}}$.

  \item[\texttt{provenance\_scan\_date}:]
    The date on which the contamination scan was last executed.
    Scans must be repeated whenever the dataset is updated or
    augmented.

  \item[\texttt{provenance\_scan\_method}:]
    The version of the detection framework used, including the
    reference model $\mathcal{M}_{\mathrm{ref}}$, the watermark
    detectors deployed, and the threshold
    $\theta_{\mathrm{quarantine}}$.
\end{description}

Listing~\ref{lst:contamination-scanner} provides a reference
implementation of the Synthetic Contamination Scanner that
computes the Provenance Score for each document in a training
corpus and outputs the SBOM-compatible contamination report.

\begin{lstlisting}[language=Python, style=pythonstyle,
  caption={Synthetic Contamination Scanner: computes per-document
           Provenance Scores and generates SBOM-compatible
           contamination reports for AI training pipelines.},
  label={lst:contamination-scanner}
]
from dataclasses import dataclass, field
from typing import List, Tuple
import hashlib, json, math

@dataclass
class ProvenanceResult:
    doc_id:         str
    perplexity_sig: float   # s_ppl in [0, 1]
    repetition_sig: float   # s_rep in [0, 1]
    watermark_sig:  float   # s_wm in {0, 1}
    divergence_sig: float   # s_div in [0, 1]
    provenance_score: float # P(d) in [0, 1]
    flagged:        bool

# -- Default signal weights (Equation 24.6) -----------------------
DEFAULT_WEIGHTS = {
    "perplexity":  0.30,
    "repetition":  0.25,
    "watermark":   0.25,
    "divergence":  0.20,
}

class SyntheticContaminationScanner:
    def __init__(self, ref_model, watermark_detectors,
                 human_ref_corpus,
                 weights=None,
                 quarantine_threshold=0.65):
        self.ref_model   = ref_model
        self.wm_detects  = watermark_detectors
        self.human_ref   = human_ref_corpus
        self.weights     = weights or DEFAULT_WEIGHTS
        self.threshold   = quarantine_threshold
        # Pre-compute reference statistics
        self._ref_ppl_median = self._median_perplexity(
            human_ref_corpus)
        self._ref_rep_median = self._median_repetition(
            human_ref_corpus)

    def scan_document(self, doc_id, text):
        """Compute Provenance Score for a single document."""
        s_ppl = self._perplexity_signal(text)
        s_rep = self._repetition_signal(text)
        s_wm  = self._watermark_signal(text)
        s_div = self._divergence_signal(text)

        score = (
            self.weights["perplexity"]  * s_ppl
          + self.weights["repetition"]  * s_rep
          + self.weights["watermark"]   * s_wm
          + self.weights["divergence"]  * s_div
        )
        return ProvenanceResult(
            doc_id=doc_id,
            perplexity_sig=round(s_ppl, 4),
            repetition_sig=round(s_rep, 4),
            watermark_sig=round(s_wm, 4),
            divergence_sig=round(s_div, 4),
            provenance_score=round(score, 4),
            flagged=(score >= self.threshold),
        )

    def scan_corpus(self, documents):
        """Scan all documents; return results + SBOM report."""
        results = [self.scan_document(d_id, text)
                   for d_id, text in documents]
        flagged = [r for r in results if r.flagged]

        sbom_report = {
            "synthetic_contamination_ratio":
                round(len(flagged) / max(len(results), 1), 4),
            "total_documents": len(results),
            "flagged_documents": len(flagged),
            "quarantine_threshold": self.threshold,
            "signal_weights": self.weights,
            "provenance_scan_method": {
                "ref_model": str(self.ref_model),
                "watermark_detectors":
                    [str(d) for d in self.wm_detects],
            },
        }
        return results, sbom_report

    # -- Signal implementations ----------------------------
    def _perplexity_signal(self, text):
        ppl = self.ref_model.perplexity(text)
        return max(0.0, min(1.0,
            1.0 - ppl / self._ref_ppl_median))

    def _repetition_signal(self, text, n=4):
        tokens  = text.split()
        ngrams  = [tuple(tokens[i:i+n])
                   for i in range(len(tokens) - n + 1)]
        if not ngrams:
            return 0.0
        raw = 1.0 - len(set(ngrams)) / len(ngrams)
        return max(0.0, min(1.0,
            raw / max(self._ref_rep_median, 1e-6)))

    def _watermark_signal(self, text):
        for detector in self.wm_detects:
            if detector.detect(text):
                return 1.0
        return 0.0

    def _divergence_signal(self, text):
        # Placeholder: KL divergence vs domain reference
        return 0.0  # Requires domain-specific implementation

    @staticmethod
    def _median_perplexity(corpus):
        # Returns pre-computed median PPL of human ref corpus
        return getattr(corpus, "median_ppl", 45.0)

    @staticmethod
    def _median_repetition(corpus):
        return getattr(corpus, "median_rep", 0.12)
\end{lstlisting}

% ------------------------------------------------------------------
\subsection{Operational Recommendations}
\label{subsec:contamination-recommendations}
% ------------------------------------------------------------------

Based on the detection framework and the model collapse
mechanism, we provide four operational recommendations for
architects building training pipelines in 2026:

\begin{enumerate}
  \item \textbf{Scan before ingestion, always.}
    Every dataset artifact entering the training pipeline
    must pass through the Synthetic Contamination Scanner
    before inclusion.
    The SBOM entry must record the contamination ratio
    $\hat{\gamma}$ and the scan date.
    Datasets without a provenance scan record are treated
    as Type~C (undisclosed) by default.

  \item \textbf{Enforce a contamination ratio cap.}
    Set a maximum allowable synthetic contamination ratio
    $\gamma_{\max}$ for the training corpus.
    The recommended default for High-Risk Vietnamese
    deployments is $\gamma_{\max} = 0.15$ (15\% synthetic
    content by document count).
    Exceeding this threshold triggers a mandatory review
    by the Data Protection Officer.

  \item \textbf{Maintain a verified human reference corpus.}
    The detection framework requires a
    \emph{verified human reference corpus} against which
    perplexity and repetition baselines are calibrated.
    This corpus must be curated from sources with provable
    human authorship (pre-2020 publications, manually
    verified Wikipedia revisions, government gazettes)
    and must be updated annually to reflect evolving
    linguistic patterns.

  \item \textbf{Cross-reference with the Trust Debt Registry.}
    A synthetic contamination ratio that exceeds the
    Conformity Assessment threshold should be recorded as a
    Type~2 (Staleness) debt entry in the Trust Debt
    Registry (Section~\ref{sec:trust-tech-debt}), because
    the model's training data no longer reflects the
    distribution that was certified.
    This creates an automatic link between supply chain
    integrity and the operational debt management framework
    of Chapter~25.
\end{enumerate}

% ------------------------------------------------------------------
\section{Hardware Integrity: From Silicon to System}
\label{sec:hardware_integrity}
% ------------------------------------------------------------------

The supply chain attacks covered so far---dataset poisoning, synthetic contamination, adversarial model weights---all target the \emph{software} layers of the AI stack. Below the software, however, sits a hardware substrate whose integrity is equally critical and far harder to audit. A model trained with perfect DP-SGD, verified with ZK-ML, and deployed in a TEE can still be compromised if the \textbf{silicon} executing those computations is itself malicious.

\subsection{Hardware Trojans: Silent Saboteurs in the Chip}
\label{subsec:hardware_trojans}

A \textbf{Hardware Trojan} is a malicious modification to an integrated circuit's design or manufacturing mask, introduced during fabrication at a foundry or during the chip design phase. Unlike software vulnerabilities, Hardware Trojans are:
\begin{itemize}
  \item \textbf{Persistent:} They survive software updates, OS reinstallation, and enclave re-provisioning.
  \item \textbf{Targeted:} A Trojan can be designed to activate only when a specific model weight pattern or instruction sequence is executed, making it invisible under standard functional testing.
  \item \textbf{Physically undetectable} without destructive analysis (die delayering, scanning electron microscopy) or side-channel measurement equipment.
\end{itemize}

For AI systems deployed in critical infrastructure (power grids, financial settlement, military logistics), the threat model must include the possibility that the GPU or NPU executing inference has been compromised at the silicon level.

\begin{table}[ht]
\centering
\caption{Hardware Attack Vectors Against AI Systems and Architectural Countermeasures}
\label{tab:hardware_attacks}
\begin{tabular}{p{3.0cm} p{4.0cm} p{5.0cm}}
\hline\noalign{\smallskip}
\textbf{Attack Vector} & \textbf{AI System Impact} & \textbf{Countermeasure} \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Hardware Trojan (chip design) & Selective output manipulation; exfiltration of model weights via covert channel & Trusted foundry programs (DARPA TRUST); formal verification of RTL netlists; side-channel behavioral fingerprinting \\[4pt]
Supply chain substitution (counterfeit chips) & Incorrect computation; reduced TEE entropy quality & Cryptographic chip authentication (PUF-based attestation); mandatory procurement from NIST-listed vendors \\[4pt]
Cache timing side-channel (Spectre/Meltdown class) & Model weight extraction; training data inference from memory access patterns & ORAM for TEE memory access (Chapter~\ref{chap:privacy_attacks}); constant-time inference paths \\[4pt]
Power side-channel (SCA) & Reconstruction of secret model parameters from power trace correlation & Differential power analysis countermeasures in enclave firmware; power noise injection \\[4pt]
Rowhammer / DRAM fault injection & Bit-flip attacks corrupting model weights or ZK-proof verification tables & ECC memory mandated for all AI inference servers; fault-tolerant weight encoding \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\end{table}

\subsection{Practical Hardware Assurance for AI Architects}
\label{subsec:hardware_assurance}

The Trustworthy AI Architect cannot inspect chip silicon directly. Instead, the architect's toolkit comprises three practical assurance strategies:

\begin{description}
  \item[1. Behavioral Fingerprinting.] Establish a \textbf{golden reference profile} of the hardware platform under a known-safe workload (inference on a published benchmark model). Periodically replay this workload and compare latency histograms, power consumption curves, and output bit patterns against the reference. Statistically significant deviations trigger a Type-2 (Staleness) Trust Debt entry and a hardware replacement request.
  \item[2. PUF-Based Attestation.] \textbf{Physical Unclonable Functions (PUFs)} exploit manufacturing variations to generate unique, unforgeable hardware fingerprints. For critical deployments, require that every compute node presents a PUF-signed attestation report as part of its TEE attestation chain. This adds a hardware-rooted identity layer that cannot be spoofed by software.
  \item[3. Diversity and Redundancy.] For the highest-criticality workloads, execute inference across chips from \textbf{two independent semiconductor supply chains} (e.g., AMD + Intel, or domestic + allied-nation foundries) and flag any output divergence exceeding a hardware-calibrated tolerance. This is the silicon-level analogue of Byzantine-fault-tolerant aggregation (Chapter~\ref{chap:fl_security}).
\end{description}

\subsection{Conflict-Free Sustainability in AI Hardware}
\label{subsec:conflict_free}

The physical hardware on which Trustworthy AI runs requires rare-earth elements and conflict minerals whose extraction chains carry significant social and environmental risks. Cobalt (used in lithium-ion batteries for edge inference devices), tantalum (capacitors in high-density compute boards), and tungsten (chip interconnects) are among the minerals listed under the OECD Due Diligence Guidance for Responsible Supply Chains of Minerals from Conflict-Affected and High-Risk Areas.

\begin{svgraybox}
\textbf{The Social Trust Layer of Hardware:} A model that protects the privacy of its users while running on hardware whose supply chain funds armed conflict in Eastern Congo has not achieved genuine Social Trust---it has merely shifted the harm downstream, outside the visibility of its architects. The \textbf{RBA (Responsible Business Alliance) Code of Conduct} and the \textbf{Conflict Minerals Reporting Template (CMRT)} provide standardized audit frameworks that Trustworthy AI teams should require from their hardware procurement partners as part of the supply chain SBOM (Section~\ref{sec:supply_chain_sbom}).
\end{svgraybox}

The architect's practical action is to extend the Software Bill of Materials (SBOM) to a \textbf{Hardware Bill of Materials (HBOM)}, documenting:
\begin{enumerate}
  \item Chip manufacturer, foundry country, and process node for every compute component.
  \item Conflict mineral sourcing declarations (CMRT form) from the chip manufacturer.
  \item PUF attestation certificate and the trusted authority that issued it.
  \item Date of last behavioral fingerprint comparison against the golden reference.
\end{enumerate}

The HBOM becomes part of the system's verifiable trust evidence package alongside the software SBOM, model card, and Privacy Budget Ledger.

\section{Conclusion}
\label{sec:conclusion_supply_chain}

Securing the AI supply chain ensures that our trust is anchored in the entire lifecycle, not just the final output. By extending supply chain security to include the detection and management of synthetic contamination---the silent, recursive poisoning of the global knowledge pool---the architect builds a training pipeline that is resilient not only to adversarial tampering, but to the emergent systemic risks of a world in which AI increasingly trains on its own outputs. With these security measures in place, we can confidently move toward the operational challenges of maintaining this trust in production. In the next chapter, we explore \textbf{Trustworthy MLOps} and continuous monitoring.


